Listening to the room: disrupting activity of dorsolateral prefrontal cortex impairs learning of room acoustics in human listeners

Version 1

Abstract

Abstract

Navigating complex sensory environments is critical to survival, and brain mechanisms have evolved to cope with the wide range of surroundings we encounter. To determine how listeners learn the statistical properties of acoustic spaces, we assessed their ability to perceive speech in a range of noisy and reverberant rooms. Listeners were also exposed to repetitive transcranial stimulation (rTMS) to disrupt the dorsolateral prefrontal cortex (dlPFC) activity, a region believed to play a role in statistical learning. Our data suggest listeners rapidly adapt to statistical characteristics of an environment to improve speech understanding. This ability is impaired when rTMS is applied bilaterally to the dlPFC. The data demonstrate that speech understanding in noise is best when exposed to a room with reverberant characteristics common to human-built environments, with performance declining for higher and lower reverberation times, including fully anechoic (non-reverberant) environments. Our findings provide evidence for a reverberation “sweet spot” and the presence of brain mechanisms that might have evolved to cope with the acoustic characteristics of listening environments encountered every day.

Introduction

Learning occurs over multiple time scales—evolutionary, developmental and moment-to-moment—to support a diverse range of abilities: navigating complex sensory environments (Bregman, 1994; Lewicki et al., 2014; Smith & Lewicki, 2006), expressing intricate behaviours in social settings (Gariépy et al., 2014; van den Bos et al., 2013), or acquiring communication skills such as bird song (Brainard & Doupe, 2002; Lauay et al., 2004; Marler, 1970) or spoken language (Aslin et al., 1998; Saffran et al., 1996; Saffran, 2003). Although some learning requires active or explicit involvement in a task (Huyck & Wright, 2011; Mathews et al., 1989; Rebuschat, 2015), perhaps with a system of rewards and punishments to make the learning ‘stick’ (Barberis, 2013; Schultz, 2002; Wächter et al., 2009), other forms of learning seem automatic or implicit (Reber, 1967), acquired with individuals seemingly unaware it is taking place. This type of learning, referred to as ‘statistical learning’ (Ambrus et al., 2020; Saffran et al., 1996), ‘sequence learning’ (Nissen & Bullemer, 1987; Vékony et al., 2022) or ‘sequential learning’ (Conway & Christiansen, 2001; Vékony et al., 2022), is thought to entail automatic and incidental extraction of regularities or patterns within external stimuli or in the environment (Conway, 2020; Takács et al., 2021). Evident across sensory modalities to support automatic learning of tonal (Saffran et al., 1999) or linguistic (Saffran et al., 1996) sequences, strings of letters (Reber, 1967), visual scenes and shapes (Fiser & Aslin, 2001), visual-motor patterns (Nissen & Bullemer, 1987), and even tactile input (Conway & Christiansen, 2005) statistical learning appears to be a unitary, domain-general phenomenon, potentially governed by a single mechanism or neurocognitive principle (Conway, 2020; Kirkham et al., 2002).

Increasing evidence suggests that statistical learning is a critical part of how listeners deal with complex and cluttered acoustic scenes. Potential background sounds such as rain or insects—referred to as sound textures (Hicks & McDermott, 2024; McDermott et al., 2013; McWalter & McDermott, 2018) as well as changes in the regularity of sound patterns (Barascud et al., 2016; Bianco et al., 2020) are processed—seemingly unconsciously—in terms of their summary statistics. This form of statistical learning likely contributes to our ability to follow conversations in background noise (‘cocktail party listening’; Cherry, 1953) and to deal with reverberant spaces where multiple, delayed copies of the same sound reach a listener in the form of reflections from walls and other acoustically opaque surfaces (Blesser & Salter, 2009; Sabine, 1953; Schroeder, 1962). Though listeners rely on early-arriving sound energy to determine source location—suppressing potentially conflicting localization cues in later-arriving, often more intense, sound energy (Blesser & Salter, 2009; Bradley et al., 1999; Culling et al., 2003; Houtgast & Steeneken, 1985; Nielsen & Dau, 2010)—the perception of reflected sound energy is informative of the listening environment more broadly (Bronkhorst & Houtgast, 1999; Shinn-Cunningham, 2000). This includes whether environments are indoors or outdoors, their identity and dimensions (Brumm & Naguib, 2009; Cabrera et al., 2005; Kolarik et al., 2021; Zahorik & Wightman, 2001), as well as the number of occupants or potential interfering sources (Bradley et al., 1999; Culling et al., 2003; Hawley et al., 2004; Houtgast & Steeneken, 1985; Nielsen & Dau, 2010; Peissig & Kollmeier, 1997).

Nevertheless, despite its potential utility for understanding background features of a sound environment, the accumulation over time of late-arriving, reverberant sound energy is thought to generate an additional burden on listening performance beyond that from sound energy direct from interfering sources (Houtgast & Steeneken, 1973; Knudsen, 1929; Lochner & Burger, 1961; Santon, 1976; Shinn-Cunningham & Kawakyu, 2003), smearing the acoustic waveform and occluding temporal gaps that might otherwise be helpful for ‘glimpsing’ speech in background noise (Cooke, 2006). If listeners are able to utilize late-arriving, reverberant energy of known acoustic environments to supress disruptive spatial cues and enhance speech understanding (Brandewie & Zahorik, 2010, 2013; Vlahou et al., 2019; Watkins, 2005a, 2005b), this suggests that the acoustic characteristics of an environment can be learned and stored for later use in complex listening tasks.

Here, using an ecologically relevant listening task—understanding speech in background noise—we assessed the ability of human listeners to learn the statistical structure of different sound environments defined by their RT60, the time it takes for reverberant energy to decay by 60 decibels (dB), and found that speech understanding in background noise improved over time. Specifically, when asked to report words from unfamiliar and semantically uninformative spoken sentences of varying duration in noise convolved with the reverberant qualities of different acoustic environments, listeners’ performance improved with increasing sentence duration, and with repeated exposure to each environment (Brandewie & Zahorik, 2010, 2013). This capacity for learning the acoustic environment to aid speech understanding was diminished when repetitive transcranial magnetic stimulation (rTMS) was applied bilaterally to impair the function of dorsolateral prefrontal cortex (dlPFC), a cortical locus hypothesised to contribute to statistical learning (Ambrus et al., 2020; Vékony et al., 2022). Counter to expectations that reverberant energy can only harm speech understanding, listeners’ abilities to leverage the knowledge of a sound environment (talker identity, speech corpus, noise characteristics, spatial configuration) to improve speech understanding was best when a ‘typical’ room with reverberant characteristics close to the average of a wide range of common (built) environments was included as one of the three environments presented in a single experimental run. Performance declined systematically when this room was switched for one with more, or less, reverberation, including a fully anechoic (i.e., non-reverberant or dry) environment in which listeners initially trained on the recall task. Importantly in the context of other potentially learnable acoustic features, talker identity (three male and three female) represented a random variable in our experimental design. Evidence for a reverberation ‘sweet spot’ suggests the existence of brain mechanisms adapted to the longer-term structure of common listening environments (Traer & McDermott, 2016).

Our data suggest listeners rapidly ‘tune in’ or adapt to the background (the statistical structure of the sound environment, including its reverberation profile) and use this knowledge to improve their performance in a listening task. This benefit of learned listening appears most evident when encountering levels of reverberation commonly experienced in everyday settings. The data also support the view that listeners retain information about the acoustic background long enough for it to influence listening performance at some later time, consistent with in vivo experimental evidence of increased adaptive capacity for neural learning of sound environments upon repeated exposure to those environments (Dean et al., 2005, 2008; Ivanov et al., 2022) and that cortical circuits modulate this capacity (Robinson et al., 2016). If long-term learning of reverberant environments relies on, or even establishes, a preferred range of RT60s for speech understanding, it suggests modifications might be required to listening assessments and technologies—routinely developed under anechoic conditions—to generate optimal performance outcomes.

Materials and methods

Participants

All 62 participants (53 females, ages between 19-26 years old (mean ± SD = 22 ± 2 years old)) included in the study were Australian native-English speakers, had normal pure tone thresholds (< 20 dB HL tested at octave intervals between 0.5-8 kHz (Hughson & Westlake, 1944); Interacoustics Hearing Aid Fitting Analyzer Affinity 2.0 Audiometry) and normal middle ear function (assessed using standard 226 Hz tympanometry; Titan, Interacoustics). To ensure normal outer hair cell function, all participants were screened for Distortion Product Otoacoustic Emissions (DPOAEs) between 0.5-10 kHz (stimulus parameters: f1/f2 =1.2, f1 =65 dB SPL; f2 =55 dB SPL; responses parameters: SNR > 6 dB, signal level > −10 dB SPL, and reliability > 98% (DPOAE440; Titan, Interacoustics). All participants had normal steady-state ipsilateral broadband noise middle ear muscle reflexes (MEMR) (Titan, Interacoustics).

Acoustic stimuli

To assess listeners’ abilities to understand speech in background noise across different sound environments, they were asked to verbally report keywords spoken by a talker virtually located in front of the participant that were masked by white noise arriving from a virtual source 90° to the left (Figure 1A) (Brandewie & Zahorik, 2010, 2013). A binaural configuration was necessary as speech intelligibility improvements following exposure to room acoustics is significantly reduced under monaural listening conditions (Brandewie & Zahorik, 2010). The speech stimuli used in this study were from the Coordinate Response Measure (CRM) corpus (Bolia et al., 2000), and all combinations of, |Callsigns| (‘Baron’, ‘Eagle’, ‘Charlie’, ‘Tiger’, ‘Arrow’, ‘Ringo’, ‘Laker’, ‘Hoper’), |Colors| (four monosyllabic choices, ‘red’, ‘white’, ‘blue’, or ‘green’), and |Numbers| (the English digits between ‘one’ and ‘eight’) were used in this experiment (Figure 1B) (Brandewie & Zahorik, 2010, 2013). The masking white noise was randomly generated on a PC using MATLAB: RRID:SCR_001622 (The MathWorks, Inc. of Natick, MA: https://au.mathworks.com/) and preceded the speech stimuli by 150 ms, during which its amplitude linearly increased from zero to full scale. The masker was present throughout the speech and ended (without ramping) with the speech.

Experimental environment.

A. Speech-in-noise immersed in real reverberant environments reproduced in an anechoic chamber over a 41-loudspeaker array. All speech material was rendered as if it was originating from directly in front of a listener and was masked by spatially separated (90° to the left) white noise at −15 dB signal-to-noise ratio (SNR). The noise masker level was fixed at 70 dB SPL. The speech material was convolved with the measured impulse response of a variety of different real rooms, from anechoic to highly reverberant real rooms such as an underground car park (see Table 1 methods for acoustic details). B. Structure of the speech material. Sentences from the Coordinated Response Measures (CRM) corpus were modified to vary their duration depending on how many words preceded the target phrase i.e., ‘carrier phrase’ (CP) length. Participants were asked to recognize and repeat target words: |Color| from a list of four and |Number| from a 1-8 list. C. Example of the experimental paradigm. Trials consisted of ‘carrier phrases’ spoken in a specific room i.e., ‘go to blue 1 now’ in the Lecture Room. Participants were never exposed to the same room consecutively and the task lasted no longer than 45 minutes. D. Sensitivity to |Color| and |Number| combined in the Lecture Room, Open-Plan Office and Underground Car Park. Mean d’ (measure of accuracy calculated as: Z (correct responses) – Z (false alarm)] denoted as circles in the boxplot (n=22 for all rooms). The horizontal line denotes the median. Upper and lower limits of the boxplot represent 1st (q1) and 3rd (q3) quartiles respectively, while whiskers denote the interquartile range (IQR =q3-q1). As previously reported (Brandewie & Zahorik, 2010, 2013), we observed an increase in performance for longer ‘carrier phrases’ in all environments. https://doi.org/10.25949/24295342.v1. All images in this figure were generated using artificial intelligence, with the exception of the Anechoic Room photograph, which was taken by the authors

Virtual Environments

Reverberant environments were reproduced by convolving the generated acoustic signals with Room Impulse Responses (RIRs) obtained using 62-channel microphone array recordings of real rooms that were subsequently decoded into 41 higher-order Ambisonic (HOA) channels (Badajoz-Davila et al., 2020; Weisser et al., 2019). These final decoded channels corresponded to the spherical array of 41 Tannoy V8 concentric loudspeakers (Tannoy) installed in the anechoic chamber at the Australian Hearing Hub where testing took place. To optimize the “directionality” of the acoustic stimuli, the speech and masking noise signals were separately convolved with RIRs that had enhanced direct sound components in loudspeakers associated with their respective virtual source locations (i.e., 0° azimuth loudspeaker for the target talker and 90° azimuth loudspeaker for the masking noise) (Badajoz-Davila et al., 2020; Weisser et al., 2019). In the anechoic condition, RIRs only incorporating these enhanced direct sound components were convolved with the speech and masking noise signals. The SNR for speech-in-noise during data collection was −15 dB to avoid ceiling performance and was manipulated after convolution with the RIR by adjusting the gain of the speech target relative to a fixed masker level of 70 dB SPL. All signal processing was performed on a PC using MATLAB (Mathworks) and the final spatialized signal were presented via the RME MADI sound card (RME Audio) to two RME 32-channel digital-to-analog converters (M-32, RME Audio). These, in turn, fed 11 Yamaha XM4180 power amplifiers (Yamaha) that drove the loudspeaker array.

To compare our recordings to RIR data recorded and analyzed by Traer and McDermott (2016), we also calculated RT60s for each room by integrating across the 41 HOA channels of each RIR and then calculating the median RT60 value in 31 frequency sub-bands with centre frequencies ranging from 80 Hz to 10 kHz (Traer & McDermott, 2016). We refer to these measures of reverberation as RT60traer. The range of sub-bands and their frequency selectivity was chosen to match that of the human ear (Glasberg & Moore, 1990; McDermott & Simoncelli, 2011). See SI Materials and Methods, (Traer & McDermott, 2016), for further information concerning how individual RT60traers were extracted from each sub-band.

Identity of the selected sound environments

Three rooms were selected, with similar Reverberation Times (RT60s) to those employed by Brandewie and Zahorik (2013). As Bajadoz-Davila et al. (2020) and Brandewie and Zahorik (2013) calculated these RT60s according to ISO-3382 guidelines (ISO, 2009), they are referred to in the text as RT60iso: Lecture room (RT60iso = 0.46 s), Open-Plan Office (RT60iso = 0.96 s) and Underground Car Park (RT60iso = 2.42 s) (Table 1). Additionally, we assessed performance in different combinations of rooms to determine whether a specific combination was necessary to improve speech performance as ‘carrier phrase length’ was increased. For each combination, Lecture Room was swapped with one of three virtual environments: Anechoic (RT60iso = n/a), Living Room (RT60iso = 0.33 s) and Highly Reflective Room (RT60iso of 1.55 s) (Table 1). Naïve participants were recruited for each room combination.

Continuous theta-burst stimulation

Applying repetitive Transcranial Magnetic Stimulation (rTMS)—specifically continuous theta burst stimulation (cTBS)—to the right and left dlPFC, elicits a period of depressed cortical excitability in the targeted area that lasts about an hour post-stimulation (Gamboa et al., 2010; Hoogendam et al., 2010; Huang et al., 2005), the time during which speech recall in reverberant rooms was assessed. Participants were screened for contra-indications to rTMS (see Supplemental Information. Transcranial Magnetic Stimulation screening form). Two cTBS conditions (‘real’ and ‘sham’ TMS) were counterbalanced across normal-hearing participants, and participants were blinded to the type of manipulation they might receive.

cTBS intensity was individually measured for every participant (i.e., both ‘real’ and ‘sham’ TMS conditions) as a function of their corticospinal excitability assessed with single-pulse TMS motor evoked potentials (MEPs). First, single-pulse TMS (Magstim Rapid2 system, Magstim) was delivered over the left and right primary motor cortex to determine the optimal site for MEP elicitation and to determine resting and active motor thresholds for each participant and each side of the brain. A 70 mm figure-of-eight coil was orientated at 45° to the scalp with current flowing posterior-anterior across the primary motor cortex. Coil position and angle was adjusted until the optimal site for consistent elicitation of MEPs was identified. Once the optimal site for stimulation was located, this was marked on the scalp. The resting motor threshold was then determined by the experimenter visually identifying the minimal single-pulse TMS intensity necessary to elicit a MEP from the right and left first dorsal interosseous muscle, while the hand was at rest in 5 out of 10 consecutive stimulations (Rothwell et al., 1999; Sandrini et al., 2011).

Participants’ resting left and right motor thresholds ranged from 46-57% of the maximum stimulator output (mean = 51 ± 5). Participants’ individual motor threshold for left and right hemisphere were recorded to set the intensity for bilateral cTBS stimulation which was administered by placing the same figure-of-eight coil over the right and left dlFPC located over electrode positions F3 and F4 (10-20 system EEG) (Herwig et al., 2003; Jurcak et al., 2005). Bilateral TMS was performed serially i.e., first the right or the left dlFPC (order was alternated among subjects), was chosen to limit possible compensation effects of the non-stimulated hemisphere (Ambrus et al., 2020). Each cTBS burst consisted of three pulses at 50 Hz, with bursts repeated at a frequency of 5 Hz, applied continuously for 40s, and delivered at an intensity of 80% of the right or left resting motor threshold (Huang et al., 2005).

Procedure

All participants completed first a familiarization task that consisted of 10 trials (i.e., different ‘carrier phrase lengths’) presented in anechoic conditions at 0 dB SNR where listeners were asked to report |Color| and |Number|. Participants exposed to ‘real’ or ‘sham’ TMS completed the familiarization task after cTBS procedures. All participants performed the familiarization phase with 80-100% accuracy, confirming that all participants including exposed to cTBS understood the task and procedural learning was achieved and not affected by either ‘sham’ or ‘real’ TMS exposure. The CRM sentences were spoken by six talkers (3 males and 3 females) and were of varying duration where all, some, or none of the preceding sentence (the ‘carrier phrase’ (CP, Figure 1B)) before ‘|Color| |Number|’ was included (Brandewie & Zahorik, 2010, 2013). Thus, for CP0, listeners heard ‘|Color| |Number| now’, CP1—‘go to |Color| |Number| now’, CP2—‘|Callsign| go to |Color| |Number| now’, CP3—‘Ready |Callsign| go to |Color| |Number| now’ (Figure 1B and C). After each phrase was presented, participants reported the |Color| and |Number| they heard to the experimenter and performance was assessed based on keywords correctly identified. When collecting performance data in reverberant environments at −15 dB SNR (SNR that allows a level of performance similar to (Brandewie & Zahorik, 2010, 2011, 2013; Zahorik, 2009)), the room environment was selected randomly for each trial except that the same room could not appear in two consecutive trials. The target |Color|, |Number| and talker were selected randomly for each trial. For each experimental session, three different listening environments were assessed, and listeners were presented 360 trials (30 repeats of each ‘room’ x ‘carrier phrase length’ condition) with a total test time of 45 min per session. Each listener participated in one session only. Throughout the session, listeners were required to verbally repeat the appropriate |Color| and |Number| combination they heard from corpus lists located to the sides of the 0° azimuth loudspeaker. The CRM is a close-set corpus, therefore participants were provided with |Color| and |Number| choices to select from (Brandewie & Zahorik, 2010, 2011, 2013; Pavel Zahorik, 2009). The experimenter continuously monitored participants’ responses while scoring performance, but no feedback was provided.

Speech Performance Analysis and Time course-fittings of mean cumulative hit rates

d’ was first calculated individually for both |Color| and |Number| in different room/carrier combinations using Equation 1, where z(H) and z(F) were the z transforms of the hit rate and false alarm, respectively. Hit rates consisted of the correct |Color| or |Number| being selected, whereas false alarms were considered when the same |Color| or |Number| was reported as an incorrect response to the presentation of any other |Color||Number| combination. To prevent infinite values of d’, hit rates/false alarms of 1 were set to 0.99, and corresponding false alarms/hit rates of 0 were set to 0.01 leading to maximal/minimal d’ values of ± 4.65. The total d’ for a carrier phrase length in a particular room was calculated by averaging |Color| and |Number| d’ values for that combination of room and ‘carrier phrase length’ across individuals.

Given how d’ was calculated for |Color| and |Number|, it was not considered a useful metric to describe performance across time in different environments: the paucity of data for each |Color| and |Number| affecting the temporal resolution of any generated curve. Therefore, to understand how the average performance developed in each listening environment as a function of time, we averaged cumulative hit rate across all subjects, after a 5-point moving average was implemented for each individual trace (equivalent to ∼7 seconds) and plotted this as a function of the mean exposure time accumulated in each environment in question. Time courses of the developing performance were quantified by fitting a double exponential function to the curves, whose model is described as follows:

Where f(t) is the fit as a function of time; magfast/magslow are the magnitudes and τfast/τslow are the time constants for the two exponential fits. C is a constant term that describes their offset on the y-axis.

Constrained optimization of the fits was achieved using the fmincon function in Matlab to find local minima in mean-squared errors. Fits were reinitialized using a combination of 50 values linearly spaced between 50 s and 500 s for τslow and between 0 s and 100 s for τfast; upper bounds of 2000 s and 200 s were set for τslow and τfast respectively. The optimal fit for a listening environment was selected from the resulting 2450 fits by identifying the fit with the smallest objective function value, fval. During optimization, fits were weighted by their variance; therefore, greater importance was attributed to performance at later exposure times. Adjusted R2 values are quoted for each fit in Supplemental Figure 1, 4 and 5. For analysis of ‘global learning’, we calculated the absolute percentage deviation of cumulative hit rate curves relative to the final time point in each curve i.e., their Final Hit Rate (FHR), we then calculated back in time to determine where fitted curves to the data first deviated from FHR by a threshold of 10%.

Statistical analysis

Repeated measures ANOVAs (rANOVA) were performed to assess if speech understanding (d’) was affected by the factors: rooms (levels: Lecture Room, Open-Plan Office and Carpark) and carrier phrase length (levels: CP0, CP1, CP2 and CP3), in a within-subjects factor analysis. Univariate ANOVA was employed to assess whether speech understanding was affected by the factors: TMS exposure (levels Sham and real TMS), rooms (levels: Lecture Room, Open-Plan Office and Carpark) and carrier phrase length (levels: CP0, CP1, CP2 and CP3) in a between-subjects factor analysis.

Univariate ANOVA was also used to assess whether speech understanding was affected by room-context (i.e., the 3rd room environment in which Open-Plan Office and Underground Car Park are learnt). Only one factor was assessed in this between-subjects room context (levels: Anechoic room, Living room, Lecture room and Highly Reflectant room). Effect sizes were calculated for all statistical analysis (Partial Eta-squared (ŋp2) or Cohen’s d when appropriate). Further two-tailed t-tests (alpha=0.05, with Bonferroni corrections for multiple comparisons) were also performed. The statistical analysis was performed in SPSS: RRID:SCR_002865 (IBM Corp. Released 2023. IBM SPSS Statistics for Windows, Version 29.0.2.0 Armonk, NY).

Power Analysis

Initial sample size estimation (≥ 18) was computed using G*Power: RRID:SCR_013726 (Faul et al., 2007) (Effect size f = 0.25; α err prob = 0.05; Power (1-β err prob) = 0.95). However, given the large effect size observed in previous studies (Brandewie & Zahorik, 2010, 2013; Srinivasan & Zahorik, 2012; Pavel Zahorik & Brandewie, 2016) using sample sizes = 9-16 listeners, we expected relatively large effect sizes with a sample ≥ 10. All variables variance were tested for normal distribution (Shapiro-Wilk test, (Shapiro & Wilk, 1965).

Data availability statement

All data and related metadata were deposited in Figshare, an appropriate public repository (https://doi.org/10.25949/24295342.v1)

Code Information

All custom code used for data collection and analysis will be available upon request.

Results

Human listeners can incorporate knowledge of a listening environment, specifically its reverberant characteristics, to improve speech understanding in noise (Brandewie & Zahorik, 2010, 2013; Srinivasan & Zahorik, 2012; Zahorik & Brandewie, 2016). To understand how this learning of sound environments is achieved, we recreated acoustic characteristics of real rooms using an array of loudspeakers located in an anechoic chamber (Figure 1A) and assessed listeners’ performance in a speech-in-noise task using sentences from the Coordinate Response Measure (CRM) corpus— “Ready |Callsign| go to |Color| |Number| now” (Figure 1B). The CRM corpus is commonly used to quantify listening performance in noisy or cluttered environments (Bolia et al., 2000). Listeners reported the |Color| (one of four monosyllabic choices, ‘red’, ‘white’, ‘blue’, or ‘green’) and the |Number| (the English digits between ‘one’ and ‘eight’) they heard for sentences of varying duration where all, some, or none of the preceding sentence (the ‘carrier phrase length’) before ‘|Color| |Number|’ was included (Figure 1B). Importantly, words in the CRM phrase preceding ‘|Color| |Number|’ are uninformative as to the color and number (Brandewie & Zahorik, 2010, 2013). CRM sentences—complete or partial (see Methods and Figure 1B)—were presented as if originating from in front of a participant, whilst a randomly generated white noise was presented from a source 90° to their left. Listeners’ abilities to recognize |Color| and |Number| from whole or partial CRM sentences were assessed. Although sentences varied in duration from trial to trial, they always contained the element ‘|Color| |Number| now’ embedded in noise were convolved with the impulse response of the real rooms (Figure 1C). Sentence length varied by adjusting the duration of the preceding ‘carrier phrase’ (CP; Figure 1B-C), which was always constructed from the same CRM sentence embedded in noise and convolved with the same impulse response as the remainder of the phrase. Thus, for CP0, listeners heard only ‘|Color| |Number| now’, CP1—‘go to |Color| |Number| now’, CP2—‘|Callsign| go to |Color| |Number| now’, CP3—‘Ready |Callsign| go to |Color| |Number| now’. After each phrase was presented, participants verbally reported the |Color| and |Number| they heard to the experimenter and performance was assessed based on keywords correctly identified (Figure 1D).

We first confirmed that understanding speech in background noise depends on the reverberant characteristics of the listening spaces from which impulse responses were obtained. Overall, RT60isos [RT60s according to ISO-3382 guidelines (ISO, 2009)] of the (six) environments varied from fully anechoic to a highly reverberant underground car park with RT60iso = 2.42 s. Our initial assessment (Figure 1D) examined listening performance in three environments: a Lecture Room (RT60iso = 0.45 s), an Open-Plan Office (RT60iso = 0.96 s), and an Underground Car Park (RT60iso = 2.42 s). Listeners reported |Color| and |Number| from CRM phrases of varying length, spoken by one of 6 talkers (3 female, 3 male) in one of the three environments. ‘Carrier phrase length’ and talker were interleaved in a pseudorandom order to avoid the same environment being presented in consecutive trials (e.g., Figure 1C). Overall, listeners performed better (quantified in terms of d’ values for reporting the correct |Color| and |Number|) in the Lecture Room—the least reverberant of the 3 environments (Figure 2A)—with a significant main effect of room [rANOVA: F (2,42) = 76.75, p<0.001, ŋp2 = 0.79]. Bonferroni-corrected post-hoc pairwise comparisons demonstrated that performance in the Lecture Room was significantly better than in the Open-Plan Office [mean difference = 0.41, p < 0.001] and Underground Car Park [mean difference = 1.05, p < 0.001]. Performance was also significantly better in the Open-Plan Office when compared to the Underground Car Park [mean difference = 0.65, p < 0.001].

Learning effects in three real rooms.

A. Overall Performance (d’) (including correct and incorrect responses to all |Colors| and |Numbers|) in the Lecture Room (LR), Open-Plan Office (OPO) and Car Park (CP). Mean d’ denoted as circles in the boxplot (n=22 for all rooms). The horizontal line denotes the median. Upper and lower limits of the boxplot represent 1st (q1) and 3rd (q3) quartiles respectively, while whiskers denote the interquartile range (IQR =q3-q1). B. Time course of performance in each room for 22 listeners. Correct responses (Hit Rate) across time spent in each room are shown, solid curves in color correspond to mean data after 5-time point moving averages, equivalent to ∼7s, were applied). Optimal time course-fittings (using a two-phase association model) are plotted as color markers in each room. Associated shaded areas representing standard errors of the mean. Dashed coloured lines represent the time point at which participants reached a stable performance in each environment [i.e., ± 10% of the Final Hit Rate (FHR)], here ‘global learning’. Notice that no statistical differences were found for ‘global learning’ among the different rooms. C. Shows a correlation analysis between FHR and the time where 10% of the FHR is achieved in each environment. Significant negative correlations (Pearson) were observed for Lecture Room, and Open Plan Office, suggesting the earlier participants reach a stable performance, the higher their FHR is. D. Performance for each carrier phrase length, d’ was significantly better as the ‘carrier phrases length’ increased except for CP2 vs. CP3 (n.s), where a roll over effect was observed. E. Performance to |Color| and |Number| for CP0 i.e., our proxy for short-term adaptation the carrier phrase where exposure to an environment remained minimal. Here all the environments have been collapsed i.e., rANOVA, main effect of “target word”. F-G. Hit Rate for Initial (1-2) and Steady (9-10) trials for keword |Color| in the CP0 condition, (F) in Reverberant conditions, indicating a significant improvement in performance for Steady trials, likely due to the accumulation of acoustic/environmental knowledge i.e., meta-adaptation (G) in anechoic (no echoes) showing a lack of improvement in performance i.e., lack of meta-adaptation in the absence of reverberation. https://doi.org/10.25949/24295342.v1

Statistical learning of reverberant environments occurs over long and short-time courses

Statistical learning likely occurs over different time courses subject to a range of possible brain mechanisms controlling or modulating the different cadences over which learning emerges (Robinson et al., 2016; Simpson et al., 2014). We sought to distinguish short-term from longer-term learning by assessing benefit to word recall of prior exposure to the listening environment over multiple time courses. The design of our paradigm—with talker, length of carrier phrase, target words |Color| & |Number| and reverberation time (listening environment) constituting random variables—means that the initial phase of learning may also be highly variable; listeners, idiosyncratically, likely experience very different parameters over the first few trials, making it difficult to assess the rate at which learning accumulates over these early epochs, and we indeed found this to be the case (see Supplemental Figure 1). Given this, we assessed the rate at which listeners accumulate knowledge to improve listening performance across the task by assessing ‘global likelihood learning’—defined here as the time at which participants achieved stable performance in each environment and quantified as the time point (backwards in time) at which performance reached ± 10% of the Final Hit Rate (FHR)]. For the 22 listeners, this measure of global learning (Figure 2B) was similar across all rooms (a total of 180 trials per room and average trial length of 1.5s): Lecture Room: [42.95 ± 38.35 s]; Open-Plan Office: [54.50 ± 29.50 s] and Car Park: [56.95 ± 33.15 s], with a one-way ANOVA revealing no statistical differences in global learning for the three reverberant environments, suggesting listeners learned these environments at the same rate.

We next wondered whether the FHR achieved by a participant within an environment was related to the time point at which global learning had manifest. To this end, a Pearson’s correlation analysis revealed FHR and the timepoint at which stable performance was achieved were negatively correlated for the Lecture Room: r(22) = −0.63, p = 0.002, 95% CI [−0.84, −0.32] and Open Plan Office r(22) =-0.49, p = 0.02, 95% CI [−0.74, −0.11], but not for the more-highly reverberant Car Park r(22) = 0.02, p = 0.94, 95% CI [−0.29, 0.39], Figure 2C. This analysis suggests that the earlier in time a listener achieves asymptotic performance (i.e., the time point defined as global learning)—presumably by accumulating knowledge about these environments over time—the higher their overall performance (defined by FHR) in the task, at least in the two less-reverberant environments assessed here.

We next analysed whether speech understanding—quantified in terms of d’— improves as the length of the ‘carrier phrase’ was increased (Figure 2D). Consistent with Brandewie and Zahorik (2010; 2013), we found a significant main effect of length of ‘carrier phrase’: [rANOVA: F (3,63) =26.59, p<0.001, ŋp2 =0.56]. Bonferroni-corrected post-hoc pairwise comparisons revealed a significant effect for most carrier phrase lengths (Supplemental Table 1) other than between the two longest, CP2 vs. CP3. This plateau in performance between CP2 and CP3 suggests an upper limit in the ability to exploit/accumulate information in reverberant listening environments with increasing sound duration, consistent with Brandewie and Zahorik’s (Brandewie & Zahorik, 2013) observation of a plateau (and a decline in some reverberant environments) in listening performance with increasing length of carrier phrase, and the existence of an upper limit of listeners to exploit prior exposure to sound environments to benefit listening (Hicks & McDermott, 2024; McDermott et al., 2013; McWalter & McDermott, 2018).

We hypothesized that if prior exposure to the statistics of reverberant rooms arises from an increase in the length of the carrier phrase, then performance to |Number| will always be better than to |Color| for the shortest carrier phrase (CP0: ‘|Color| |Number| now’; Figure 2E) as |Number| is subject to a longer preceding phrase than |Color|. Assessed in terms of reporting |Color| and |Number| alone, listeners performed significantly better (i.e., showed greater sensitivity) for |Number| compared to |Color|: main effect of target word: [rANOVA F (1,21) =48.79, p<0.001, ŋp2 =0.70], [mean difference = 0.38, p < 0.001] despite there being twice as many (eight compared to four) possibilities. Moreover, a significant interaction ‘carrier phrase length’ x ‘target word’ was observed: [rANOVA F (1,21) =4.33, p=0.008, ŋp2 =0.17]. Post-hoc pairwise comparisons (see Supplemental Table 2) indicate that for all carrier phrases except CP3, performance to |Number| was significantly higher than for |Color|. This result also suggests that a roll-over effect i.e., a lack or limit to the improvement in performance is evident for carrier phrases longer than CP2 across environments (Brandewie & Zahorik, 2013; McWalter & McDermott, 2018).

A potential explanation for poorer performance in accurately reporting |Color| relative to |Number| for the shortest carrier phrase (CP0) impact of the utterance of |Color|, namely its immediate appearance at the start of the phrase i.e., the impact of “order/certainty or predictability” in statistical learning (Conway et al., 2010; Daikoku & Yumoto, 2023). One way to determine whether |Color| presented in the context of the CP0 is at all subject to statistical learning (and is therefore not wholly reliant on the short-term accumulation of information within a CP0 trial; see Figure 3G) is to assess whether performance for |Color| for the shortest carrier phrase (CP0) improves from the longer-term accumulation of knowledge i.e., the process of meta-adaptation (Robinson et al., 2016) in which adaptation to short-term statistics improves with repeated exposure to those statistics. Specifically, we tested the hypothesis that, independent of the environment, performance for |Color| in later steady trials (here, trials 9-10) of the shortest carrier phrase, CP0, are better compared to initial (1-2) trials (Figure 2F). If this hypothesis is supported, it suggests performance for |Color| for CP0 benefits from meta-adaptive information conveyed in later trials as knowledge about the global structure of the environment accumulates over time. Consistent with this hypothesis, a Wilcoxon signed rank test revealed significant better performance (n=22, Z= −2.16, p=0.03) for |Color| for steady trials (9-10; assumed to reflect a meta-adaptive state; Robinson et al., 2016) compared to initial trials (1-2; i.e., short-term adaptation). Interestingly, in anechoic conditions i.e., in the absence of reverberation, this meta-adaptive process—the expected improvement in performance between initial and steady trials—was not observed: Wilcoxon signed rank test [n=10, Z= −0.71, p=0.48]. This suggests that performance for |Color| conveyed in the shortest carrier phrase, CP0, improves over time even in the absence of immediate information in the form of any preceding carrier phrase, with knowledge of the statistical/acoustical properties of environments accumulating over the course of the experimental task.

Learning effects in ‘sham’ and ‘TMS’ conditions.

A. Schematic representation of functional connections between the dorsolateral prefrontal cortex (dlPFC) and primary auditory cortex (A1) under transcranial magnetic stimulation (TMS). B. Average exposure time (s) ‘sham TMS’ and ‘real TMS’ conditions. Correct responses across time spent in each room for ‘sham’ and ‘TMS’ conditions. Solid curves in pink and gray correspond to mean data from ‘TMS’ and ‘Sham’ conditions respectively (after 5-time point moving averages, equivalent to ∼7s, were applied), with the associated shaded areas representing standard errors of the mean. Dashed black (sham TMS) and pink (real TMS) lines represent the time point at which participants reached a stable performance in each environment [i.e., ± 10% of the Final Hit Rate (FHR)], here ‘global learning’, no statistical differences were found for ‘global learning’ between TMS conditions. C. Shows a correlation analysis between FHR and the time where 10% of the FHR is achieved in under TMS (pink) and Sham (gray) conditions. Significant negative correlations (Pearson) were observed under Sham stimulation, suggesting the earlier participants reach a stable performance, the higher their FHR is. This relationship was disrupted (lack of significant correlation) under TMS conditions. D. Overall Performance in TMS and Sham exposed participants i.e., collapsed sensitivity (d’) to |Color| and |Number| for all carrier phrases and environments (significant main effect of TMS conditions: Univariate ANOVA). d’ values plotted in bright pink corresponds to the ‘real TMS’ condition whereas the ‘sham TMS’ accuracy is plotted in gray. Mean d’ is denoted as circles in the boxplot (n=11 for ‘sham TMS’ conditions and n=10 for ‘real TMS’ conditions). E. Significant Interaction ‘TMS condition’ x ‘carrier phrase length’. d’ was significantly better as the ‘carrier phrases length’ increased for ‘sham TMS’ compared to ‘real TMS’ except for CPO (n.s), i.e., our proxy for short-term adaptation and minimal exposure to acoustic environments. F. Performance to |Color| and |Number| for CP0 i.e., our proxy for short-term adaptation and the carrier phrase where exposure to an environment remained minimal. Here all the environments have been collapsed as only a main effect of “target word”, performance to number always significantly higher than for colour independent of TMS condition. G. Schematic for assessing short-term adaptation (performance to |Number| in the Initial Trials) and meta-adaptation (performance to |Color| between Initial and Steady Trials). H. Isolated short-term adaptation effects for performance to |Number| in CP0. Hit Rate for Initial (1-2) trials only for |Number| in the CP0 condition for both TMS (pink) and Sham (gray) conditions. Notice that no differences were observed between performance |Number| under Sham or TMS conditions, suggesting no disruption of short-term adaptation.I. Hit Rate for Initial (1-2) and Steady (9-10) trials only for |Color| in the CP0 condition for both TMS (pink) and Sham (gray) conditions. Notice that only a significant improvement in performance for from Intitial to Steady trials was observed in the ‘sham TMS’ condition. https://doi.org/10.25949/24295342.v1

Improvements in performance are explained by exposure to the environment not talker idiosyncrasies

Normal-hearing listeners are reported to understand speech slightly better when listening to male talkers in a mixture of male and female talkers (Larsby et al., 2015), and quickly adapt to talkers’ idiosyncrasies such as non-native accented speech (Idemaru & Holt, 2011, 2014; Liu & Holt, 2015). Here, we wondered if exposure to different talkers (3 female and 3 male) could explain the improvement in speech as the length of the carrier phrase was increased (Supplemental Figure 1). However, this was not the case. Despite listening performance to some talkers appearing overall better than for others [main effect of talkers: rANOVA: F (5,105) = 27.19, p<0.001, ŋp2 =0.56, i.e., significantly worse performance for Talker 1 (male) and Talker 6 (female), see Supplemental Figure 2A and Supplemental Table 3], overall differences in performance due to a specific talker did not explain the benefit to speech understanding as the length of the carrier phrase was increased in each environment (i.e., lack of significant interaction: or ‘talker’ x ‘carrier’ x ‘room’: F (30,630) = 0.73, p=0.85, ŋp2 =0.034]).

To explore further the possibility of any talker-specific effects on performance, we assessed the potential benefit of experiencing the same talker on consecutive trials across an experiment, hypothesizing that, independent of room characteristics or length of carrier phrase, experiencing the same talker on consecutive trials would provide a benefit to listening performance. Across all our participants, a total of 1262, 217, and 40 trials were identified as having two, three, or four consecutive trials in which the same talker appeared (see Supplemental Figure 3). Despite the possibility that sustained experience of the same talker might lead to improved listening performance, a Wilcoxon signed rank test revealed no significant differences in performance between trial 1 vs. 2 consecutive/same talker trials (n=1262, Z= −0.42, p=0.68), trial 1 vs. 3 consecutive/same talker trials (n=217, Z= −0.17, p=0.87) or trial 1 vs. 4 consecutive/same talker trials (n=40, Z= −0.159, p=0.11). Our data suggest that listeners do not use talker identity, at least in the task reported here, to benefit their speech-in-noise understanding.

We also explored the extent to which a talker and/or the task itself could be learned by analysing improvements in performance with increasing exposure to carrier phrases and talkers in the absence of reverberation i.e., under anechoic conditions. Although a significant rANOVA was observed: F (3,27) = 9.10, p=0.007, ŋp2 =0.57; post-hoc pairwise comparisons, with Bonferroni corrections, revealed that only a significant improvement in performance was observed only between CP0 and CP2 [mean difference = 15.67, p = 0.02; see Supplemental Table 4], suggesting that in anechoic conditions very little improvement in performance is achieved by learning the talker/task as length of carrier phrase increases. These data support us employing a listening task with high ecological relevance—comprehending speech—to demonstrate the potential benefits of learning background acoustic features to leverage listening performance without requiring listeners to attend to or report features related to the statistics of presumed background sounds per se (e.g., Agus et al., 2014, McWalter & McDermott 2018, Bianco et al., 2020).

TMS disrupts long-but not short-term adaptation to an environment’s reverberation profile

The capacity for learning the statistical structure of acoustic environments, the better to understand speech embedded in background noise, suggests a real-world benefit to this form of learned listening. To determine possible brain mechanisms underlying this ability, we reversibly impaired dorsolateral prefrontal cortex (dlPFC, Figure 3A) — a brain region implicated in listening performance in noise (Houtgast & Steeneken, 1973; Knudsen, 1929; Lochner & Burger, 1961)—using repetitive transcranial magnetic stimulation (rTMS) and then assessed the ability of listeners to recall |Color| and |Number| in our modified CRM sentences. Repetitive TMS is posited to elicit a period of depressed cortical excitability in the targeted area that persists for about an hour post-stimulation (Gamboa et al., 2010; Hoogendam et al., 2010; Huang et al., 2005). Specifically, 10 naïve, normal-hearing listeners were subjected to ‘real’ TMS— continuous theta burst on dlPFC for 40s on each side—and then transferred to the anechoic chamber, where they were presented sequences of CRM phrases of varying phrase length in background noise convolved with one of the three listening environments (Lecture Room, Open-Plan Office, and Underground Car Park) as before. A second, control, group of 11 naïve participants underwent ‘sham’ TMS stimulation in which an otherwise-identical TMS procedure was performed with a ‘sham’ TMS coil. All participants were naïve to differences in the TMS procedure as well as to the listening task; indeed, as with the experimental group, participants in the ‘sham’ group experienced a short period of single-pulse TMS stimulation to obtain motor thresholds, followed by the ‘sham’ TMS with the stimulator positioned over dlPFC. This process familiarised all participants with the procedure and the influence of TMS in generating involuntary finger movements through direct stimulation of motor cortex. We assume that participants in the ‘sham’ group believed later assessment of speech recall occurred under the influence of ‘real’ TMS.

Given the reduction in sample size, as well as the potential placebo effects of a ‘sham’ TMS stimulation, we first confirmed that our sample of 11 ‘sham’ TMS participants exhibited similar behavioural performance to the larger sample of 22 participants who had not been exposed to any TMS manipulation. To avoid sample size imbalances in these comparisons, a random sample of 11 subjects was selected from the 22 participants. A Univariate analysis confirmed that overall performance in these two populations was comparable, with no significant main effect of conditions (‘sham’ vs. no exposure to TMS) observed: [F (1,263) =1.10, p=0.29, ŋp2 = 0.005]. As expected, significant main effect of rooms [F (2,131) = 83.84, p<0.001, ŋp2 = 0.41] and carrier phrase length [F (3, 43) =20.52, p< 0.001, ŋp2 = 0.20] were observed, confirming that participants experiencing ‘sham’ TMS did not perform significantly differently from the ‘no exposure to TMS’ population.

We first assessed learning over the time-course of the task itself—‘global learning’— as the mean cumulative hit rates for listeners reporting |Color| and |Number|, fitted for each TMS condition and environment with a two-phase association model (Figure 3B and Supplemental Figure 4 and 5). Our metric for ‘global learning’ was achieved for Lecture room at: ‘sham’ TMS [40.40 ± 32.57 s] and ‘real’ TMS [63.00 ± 20.82 s]; for Open-Plan Office: ‘sham’ [53.70 ± 30.72 s] and ‘real’ TMS [36.40 ± 37.36 s]; and for Car Park: ‘sham’ [61.40 ± 41.42 s] and ‘real’ TMS [69.80 ± 37.31 s]. A Univariate ANOVA between ‘sham’ TMS and ‘real’ TMS exposed participants revealed no statistical differences in global learning between conditions (mean global learning ‘sham’ TMS= 47.93 ± 33.34s) and mean global learning ‘real’ TMS= 56.40 ± 34.85s) suggesting that participants achieved a stable level of performance (± 10% of FHR) at a similar time point in the task under both ‘real’ and ‘sham’ TMS. However, Pearson’s correlation analysis (collapsed across all environments) revealed a negative correlation between FHR and ‘global learning’ for ‘sham’ TMS listeners (see Figure 3C): [r(11) =-0.45, p = 0.01, 95% CI [−0.70, −0.17] but not for ‘real’ TMS listeners: [r(10) = 0.08, p = 0.69, 95% CI [−0.22, 0.37]. Specifically, participants undergoing ‘sham’ TMS who achieved stable performance (±10% of the Final Hit Rate) relatively early in the task maintained this level of performance throughout. Relative delay in attaining stable performance was associated with a lower FHR, with the time at which early ‘global learning’ is achieved predicting overall task performance. This relationship was not evident in listeners subject to ‘real’ TMS, where the time to reach a stable performance was not predictive of the magnitude of final performance (FHR). Under TMS manipulation of dlPFC, ‘global learning’ had no predictive value, indicating that task performance was less stable and reliable for participants exposed to TMS.

We next employed a Univariate analysis to compare overall performance (d’)— including FHR and False Alarm Rate, (see Figure 3D)— of ‘sham’ participants with those who received ‘real’ TMS, and observed a main effect of TMS conditions [Univariate ANOVA [F (1,501) = 26.34, p<0.001, ŋp2 = 0.1], where ‘sham’ participants showed significantly better abilities to recall |Color| and |Number| compared to participants subjected to ‘real’ TMS [post-hoc pairwise comparisons with Bonferroni corrections: [mean difference = 0.32, p < 0.001]. In addition, a significant interaction ‘TMS condition’ x ‘carrier phrase length’ was observed, Figure 3E: [F (3,123) = 2.82, p=0.039, ŋp2 = 0.02]. A post-hoc pairwise comparison with Bonferroni corrections showed that performance in the CP0 condition was not statistically different between ‘sham’ and ‘real’ TMS conditions (mean difference = 0.02, p = 0.86), however performance in the recall task (see Figure 3E and Supplemental Table 5) was better found for ‘sham’ compared to ‘real’ TMS for CP1 (mean difference = 0.34, p = 0.006), CP2 (mean difference = 0.41, p = 0.001), and CP3 (mean difference = 0.50, p < 0.001). Listeners exposed to ‘sham’ TMS retained the improvements of performance as the length of carrier phrase was increased whereas those listeners receiving ‘real’ TMS— in which activity in dlFPC is presumably impaired—appeared unable to accumulate over time knowledge of the sound environment.

The different time courses over which statistical learning might occur suggests the involvement of multiple neural mechanisms and brain circuits (Anderson & Malmierca, 2013; Antunes & Malmierca, 2011; Robinson et al., 2016), with different cadences of learning potentially controlled, or at least modulated, by different brain centres. The time-course of midbrain neurons to adapt to different sound environments, for example, is slowed, and the capacity for retaining a ‘memory’ of those sound environments disappears, when cortex is inactivated through cooling (Robinson et al., 2016). This suggests feed-forward and feed-back mechanisms, with their own time-courses or time-constants, contribute to overall performance, including the rate at which sound environments are learned. We specifically wondered if ‘real’ TMS had a detrimental effect on short-term adaptation i.e., the observed advantage in performance for |Number| compared to |Color|, for the shortest carrier phrase CP0 (Figure 2E and 3F). A Univariate analysis revealed a main effect of ‘target word’, where performance for |Number| was always significantly better than for |Color| for all the environments explored and across TMS conditions [F (1,498) = 53.53, p<0.001, ŋp2 = 0.11]. However, no interaction between ‘TMS conditions’ and ‘target word’ was observed, indicating that in both ‘real’ and ‘sham’ TMS conditions, the same trend of better performance to |Number| compared to |Color| was evident, Figure 3E: [‘sham’ TMS: mean difference = 0.81, p < 0.001 and ‘real’ TMS: mean difference = 0.66, p < 0.001]. However, overall d’ has contributions of short-term and meta-adaptive (long-term accumulation of knowledge about the environments) across the task.

To dissociate specifically short-term adaptation from longer-term meta-adaptation, we compared performance (Hit Rate) to |Number| for initial trials (1-2), for CP0 only, as performance for this length of carrier phrase should be influenced only by short-term accumulation of information (i.e., within trial; see Figure 3G). A non-parametric Wilcoxon signed rank test revealed no significant differences in |Number| performance for initial trials between ‘sham’ and ‘real’ TMS exposed listeners (n=10, Z= −0.28, p=0.78; Figure 3H), suggesting that rapid learning of the statistical structure of the reverberant environment—in the order of a few hundreds of milliseconds—is resistant to the effects of ‘real’ TMS applied to dlPFC. To determine how much meta-adaptative improvement in performance could be disrupted by applying ‘real’ TMS to dlPFC, we then compared initial trials (1-2) and steady trials (9-10) for CP0—our proxy for meta-adaptation (Figure 3G). We tested the hypothesis that disruption of dlPFC would affect late (meta-) but not early adaptation. To this end, we expected no differences between performance in initial trials between ‘real’ and ‘sham’ TMS (as initial trials are influenced only by short-term adaptation), and a relative lack of improvement in performance in later, steady trials (meta-adaptation influenced) in ‘real’ compared to ‘sham’ TMS. Consistent with our hypothesis, a Wilcoxon signed rank test revealed no significant differences in performance for initial trials between ‘sham’ and ‘real’ TMS exposed listeners (n=10, Z= −0.21, p=0.83), but improved performance for steady compared to initial trials (n=10, Z=-2.06, p=0.04) for ‘sham’ TMS only (Figure 3I). These data suggest that performance in |Color| for the shortest carrier phrase, CP0, was disrupted due to impaired accumulation of knowledge about the environments encountered. In turn, performance under ‘TMS’ in steady trials (presumed to be influenced by meta-adaptation) was reminiscent of performance observed in initial trials where only immediate knowledge could be used to improve performance. This is reminiscent of adaptation to sound environments reported in vivo (Dean et al., 2005, 2008; Robinson et al., 2016; Wen et al., 2009), where interrupting efferent feedback by cortical cooling impairs the capacity of midbrain neurons to learn the statistical structure of sound environments; though neurons adapt to the short-term statistical structure of a sound environment each time they are exposed to it, they fail to demonstrate the acceleration of adaptation and improvement in (neural) discrimination performance as the same environment is re-encountered, adapting to it only as if exposed to it the first time.

Statistical learning of room acoustics is tuned to commonly experienced reverberation times

Reverberation is a common feature of many acoustic environments—natural and built (Traer & McDermott, 2016). Interestingly, Zahorik and Brandewie (2016) reported that improvements in speech understanding when a ‘simulated room’ is repeatedly encountered were best when listeners experienced moderately reverberant environments. This suggests that brain mechanisms responsible for listening in noise might be adapted to conditions of reverberation. Francl & McDermott (2022) described an auditory model able to reproduce several ‘human-like’ spatial hearing features when successfully trained under realistic listening conditions, such as in noise and reverberation. To this end, we wondered if the learning effects we observed similarly rely on the combination of real-world environments we employed, i.e., the RT60s in which listening performance was assessed. Specifically, we hypothesized that the ability to understand speech in noise depends on exposure to specific values of RT60 across our experimental paradigm, including those in the range commonly experienced by human listeners in natural and built environments (Traer & McDermott, 2016). To test this hypothesis, we recruited a new population of naïve participants with no previous exposure to our experimental paradigm and assessed their ability to understand speech in noise using the same CRM corpus as before, but in different combinations of acoustic environments i.e., room contexts defined by their RT60s.

We first tested the ability of 10 participants to recall |Color| and |Number| in Open-Plan Office and Underground Car Park as before, but with the Lecture Room (RT60iso = 0.42 s) swapped out for a more highly reflective elevator lobby (, Figure 4) with an RT60iso of 1.55 s—i.e., between that of the Open-Plan Office (0.96s) and the Underground Car Park (2.42s). We therefore changed the ‘room-context’ in which Open-Plan Office and Underground Car Park were learned, replacing the Lecture Room with a Highly Reflective Room. When contrasting (Univariate ANOVA) the performance (hit rate) of these 10 naïve participants vs. the 11 subjects randomly selected from the initial 22 participants sample exposed to the initial three rooms (i.e., Lecture Room, Open-Plan Office and Underground Car Park), overall performance was significantly better in the context of the Lecture Room compared to the Highly Reflective Room (Figure 4 A-B), with a main effect of room-context [F(1,251) = 207.49, p<0.001, ŋp2 = 0.46], mean difference = 20.09, p < 0.001] and a significant interaction of room-context x room [F(2, 125) = 26.64, p<0.001, ŋp2 = 0.18]. Post hoc comparisons focused only on the overlapping rooms i.e., Underground Carpark and Open-Plan Office as it was expected that longer reverberation times, such as those in the Highly Reflective Room, have detrimental effects in speech understanding compared to shorter reverberation times i.e., Lecture Room (Houtgast & Steeneken, 1973; Knudsen, 1929; Lochner & Burger, 1961) [mean difference = 34.48, p < 0.001]. Performance in the Underground Car Park was significantly better when presented in the context of the Lecture Room than in the context of the Highly Reflective Room [mean difference = 13.30, p < 0.001]. A similar trend of improved performance in the Open-Plan Office was observed in the context of the Lecture Room when compared to the Highly Reflective Room context: [mean difference = 12.50, p < 0.001].

Performance in different 3-room combinations where acoustics of 3 rooms span an ecological range.

A. Overall hit rate for different 3-room combinations. Mean data for four different 3-room combinations are plotted where Open-Plan Office (dark blue) and Underground Car Park (yellow) were always included and the context room was either Anechoic (black), Living Room (red), Lecture Room (green), Lecture Room (green) or Highly Reflective (purple). B. Performance in Open-Plan Office and Car Park when paired with different context rooms. Final hit rates were highest for Office and Car Park when presented in conjunction with the Lecture Room, followed by the Living Room and Anechoic with poorest performance observed when the Highly Reflective space was the context. C. Environments tested compared to ecological range. The RT60traers for the five test rooms, calculated as the median RT60 value across 31 frequency sub-bands (Traer & McDermott, 2016), are plotted (colored boxes) above a histogram of the median RT60traers calculated for 199 indoor (filled bars) and 72 outdoor (empty bars) spaces recorded by Traer and McDermott (2016). D. Frequency dependence of reverberation time (RT60traer) in test rooms compared to ecological range. RT60s of the 5 test rooms are displayed for frequency sub-bands used to calculate RT60traer (colored lines). Quartiles for combined indoor and outdoor spaces (Traer & McDermott, 2016) are plotted as black dashed lines. Decreasing reverberation time above 0.5kHz is observed for Living Room, Lecture Room and Car Park as has been typically described for indoor spaces (Traer & McDermott, 2016); however RT60traer profiles for Open-Plan Office and Highly Reflective space were notable for their longer reverberation times at and above 1kHz. https://doi.org/10.25949/24295342.v1

Wondering whether the ‘room-context’ effect was determined by task difficulty i.e., the more-highly reverberant lobby being intrinsically more difficult for listening than the Lecture Room, we recruited a further 11 naïve participants and performed the same task in which they were exposed to another combination of three rooms, here with the Lecture Room swapped for a Living Room with a shorter RT60iso (0.33 s) i.e., an expected ‘easier’ room. Surprisingly, although performance was improved overall relative to that in the combination of the three highly reverberant rooms (Figure 4 A-B), it remained marginally poorer [mean difference = 4.04] than when the Lecture Room was included in combination with the two fixed rooms, i.e., the Open-Plan Office and Underground Car Park (main effect of room-context [F (1,263) = 9.32, p=0.002, ŋp2 = 0.04].

Although reverberation might be considered detrimental to speech-in-noise performance, short-term benefits of more, compared to less, room reverberation for speech understanding have, in fact, been reported, including for the task we report here (Brandewie & Zahorik, 2010, 2013). Anechoic spaces—rooms whose walls are treated to remove completely reflected sound energy—have been extensively reported to aid speech understanding, especially in background noise. Notwithstanding this possibility, however, anechoic listening environments are rare— whether natural or built—and it is possible that listening performance in noise is indeed better when listeners have access to more commonly experienced, or ethologically realistic, levels of reverberation, a possibility suggested by the small, but significant, reduction in performance in all three rooms when the living room replaced the Lecture Room. To test this hypothesis directly, we recruited a further 10 naïve participants and compared their speech-in-noise performance in a combination of an Anechoic Room and the original Open-Plan Office and Underground Car Park, with the performance of the 11 participants in the original three rooms. Although task difficulty might be expected to be at its lowest i.e., easiest, in the less-reverberant Anechoic Room, this was not the case. Performance was poorer when the combination of three rooms contained an Anechoic Room and the other two, more highly reverberant rooms (Figure 4 A-B), with a main effect of ‘room-context’ [F (1,251) = 34.42, p<0.001, ŋp2 = 0.12], [mean difference = 9.19, p < 0.001].

Seeking to explain why some rooms might be better than others for the improvement of speech understanding over time, we recalculated reverberation times for our 5 reverberant rooms according to the method of Traer and McDermott (2016) where RT60traer equals the median RT60 measured in 31 sub-bands with centre frequencies between 80 Hz and 10 kHz (see Methods). We then compared these with the values collected for a wide range (n= 271) of built and natural environments by the same authors (Figure 4C-D. RT60traers for 5 rooms conditions and Traer’s data, plotted as log10 RT60traer for visualisation). In particular, the RT60traer of the Lecture Room, at 0.49 s, lies extremely close to the median and mean RT60traer values of the built environments [0.42 s and 0.50 s respectively generated by the skewed distribution (n= 199), Figure 4C]. This suggests that superior performance in understanding speech understanding in noise in the Lecture Room—and the positive impact on speech understanding when the Lecture Room is included in the three-room listening task— is related to its reverberant characteristics being commonly encountered in everyday listening situations. Notably, the mean and median RT60traers of the recorded natural environments were much lower (0.12 s and 0.16 s, respectively [n = 72], Figure 4C) than the RT60traers of the Lecture Room. This suggests that brain mechanisms contributing to effective speech understanding in reverberant environments might have adapted to the range of environments experienced over some longer time course than we assessed here.

Discussion

We assessed the ability of listeners to understand speech embedded in background noise and convolved with the room impulse responses (RIRs) of different indoor environments defined by their reverberation decay time, or RT60—the time taken for reverberant sound energy to decay 60 decibels (dB). We found speech understanding improved with repeated exposure to an environment, a form of learning that was impaired following continuous, bilateral theta-burst transcranial magnetic stimulation (TMS) of dorsolateral prefrontal cortex (dlPFC). Specifically, we observed rapid learning i.e., an improvement in speech understanding over the first few seconds of room exposure, likely due to listeners learning the acoustic reverberation—the only variable held constant across trials for a given environment. This learning, on the timescale of several seconds, was impaired by TMS, whilst learning at shorter or longer timescales was not, suggesting that TMS applied to dlPFC specifically disrupted learning of the reverberant characteristics of an environment. Listeners also showed better listening performance in moderately reverberant environments, and this performance was transferable between different acoustic environments. Specifically, the ability to correctly report keywords spoken in environments with the more extreme—lower or higher—RIRs was best when these environments were encountered in experimental blocks also containing the moderately reverberant Lecture Room or Living Room. This transference suggests an ethological tuning to more commonly encountered environments when learning acoustic backgrounds.

A role for dlPFC in statistical learning of room acoustics

Inactivation of dlPFC using theta-burst TMS reduced overall performance in our speech-in-noise task, assessed in terms of the hit rate for |Color| and |Number| in CRM phrases as a function of duration of exposure to a sound environment. Notably, however, TMS did not impair listening performance for the shortest-duration phrase, ‘|Color| |Number| now’. For this duration of carrier phrase, overall performance, and the tendency to report more accurately |Number| compared to |Color|—despite |Number| having twice as many options as |Color|—was unaffected by TMS stimulation of dlPFC. The robustness of listening performance to the shortest carrier phrase suggests that the ability to hear out speech in background noise and potentially to learn this over time in complex acoustic environments does not rely solely on cortical feedback from dlPFC. The ability to leverage performance on |Number| by exposure to |Color| presented in background noise convolved with the room impulse response is evident even when the function of dlPFC is (presumably) impaired. It is the case, however, that the rate of learning—i.e., how rapidly performance improved with increasing time of exposure to an acoustic environment—was influenced when dlPFC was impaired. Under the influence of prior TMS stimulation, listeners were less able, and less reliable when learning the acoustic background features of an environment to enhance their speech understanding, particularly in the more challenging environments of the Underground Car Park and Open-Plan Office.

Dense connections between dlPFC and the sensory cortices provide the conduit for prediction-based, top-down regulation (Morrone, 2010). These model-based predictions rely on learned templates of sensory environments (Alexander & Brown, 2018), which are critical for making inferences under situations of sensory uncertainty such as those encountered in noisy and challenging listening environments (Bartolo & Averbeck, 2021). Dorsolateral prefrontal cortex can influence auditory processing by means of direct connections to auditory cortex (Hackett et al., 1999; Plakke & Romanski, 2014)—see Figure 3Ai—and to the auditory thalamic reticular nucleus, a key modulator of auditory thalamo-cortical loops (Zikopoulos & Barbas, 2006). In return, dlPFC receives projections from primary and secondary auditory cortex (Barbas & Pandya, 1987; Barbas & Pandya, 1991; Goldman-Rakic & Schwartz, 1982; Pandya & Barnes, 2019; Petrides & Pandya, 2002), indicating that a reciprocal auditory-to-prefrontal ‘listening loop’ exists to support hierarchical predictions and prediction-error feedback. These ‘listening loops’ could potentially extend to cortical efferent feedback (Blackwell et al., 2020; McAlpine & de Hoz, 2023) and modulate the function of the the inner-ear sensitivity in attended speech-in-noise tasks (de Boer & Thornton, 2008; Garinis et al., 2011; Giraud et al., 1997; Mishra & Lutman, 2014; Hernández-Pérez et al., 2021).

The specific mechanism by which TMS-induced modulation of dlPFC impairs listening performance remains unknown. Neurostimulation studies targeting dlPFC during learning paradigms have generated conflicting directions of effect i.e., stimulation of dlPFC is reported to impair implicit learning (Nydam et al., 2018; Pascual-Leone et al., 1996) or enhance it (Ambrus et al., 2020). Ambrus and colleagues argued that impairment of dlPFC during implicit/statistical learning allows for increased engagement of a learning mechanism that is ‘model-free’ i.e., does not rely on inherent sensory representations and therefore takes longer to consolidate. Here, we speculate that the short periods over which learning is required in our task are subserved by a pre-frontally mediated, ‘model-based’ learning able to construct predictions ‘on the fly’ (Daw et al., 2005). We suggest, in our study, that disrupting a rapidly acting model-based form of learning supported by dlPFC, as part of the auditory-prefrontal ‘listening loop’, leads to impaired learning of reverberant environments.

Dorsolateral prefrontal cortex is also implicated in several, high-level cognitive functions including multi-sensory integration (Fuster et al., 2000), executive functions such as inhibition and working memory (Castro-Meneses et al., 2016; Coltheart et al., 2018; Wang et al., 2015) and listening in noisy environments (Du et al., 2016). The impact of TMS on the overall performance of our speech-in-noise task may be linked to the involvement of dlPFC in more generalized cognitive functions such as executive, memory, and attention processes. Dissociating the potential influence of TMS on such processes relative to a specific role in speech-in-noise understanding in noisy, reverberant environments would be an ideal next step. Specifically frontal regions have been linked to sensorimotor integration during speech-in-noise perception in young and older human participants (Du et al., 2016), suggesting that TMS might disrupt global functioning required for the analysis and understanding of speech. Whilst not entirely unrelated, all participants completed a familiarization task preceding the main experiment, which consisted of 10 trials presented in anechoic conditions at 0 dB Signal-to-Noise Ratio (SNR). Participants exposed to either ‘real’ or ‘sham’ TMS underwent this task following cTBS procedures. Notably, all participants, including those subjected to cTBS, achieved 75-100% accuracy during the familiarization phase, with no statistical differences between TMS groups [mean difference=1.25, t(1, 9)=0.43, p=0.68, d=0.14], confirming their comprehension of the task and successful procedural learning (i.e. learning the task per se). In addition, speech understanding in the most challenging condition (CP0) was not disrupted following TMS manipulations, suggesting that the effect of TMS was specifically in terms of influencing the way these environments were learnt over time. Nevertheless, we acknowledge that the primary task may have imposed greater cognitive demands, potentially requiring a more significant involvement of the dlPFC to meet such task requirements.

What is being learned in statistical learning of acoustic features?

The need to communicate in reverberant environments is common to daily life. It is well known that long reverberation times have detrimental effects on speech quality and intelligibility (Brandewie & Zahorik, 2010, 2013; Srinivasan & Zahorik, 2012; Zahorik & Brandewie, 2016) and that, specifically, reverberation blurs phoneme boundaries, increasing the extent to which similar words might be confused, e.g., ‘sir’/’stir’ (Watkins, 2005b; Watkins & Makin, 2007). However, with sufficient exposure to longer reverberation times in the form of a carrier phrase, similar levels of word identification to those achieved in minimal reverberation are observed (Watkins, 2005b; Watkins & Makin, 2007). Based on the premise that providing sufficient contextual information enables the auditory system to adapt and compensate for the detrimental effects of reverberation, Zahorik and colleagues (Brandewie & Zahorik, 2010, 2013; Srinivasan & Zahorik, 2012; Zahorik & Brandewie, 2016) and, here, ourselves, demonstrate that prior exposure to the reverberant characteristics of a room can enhance sentences understanding when that environment is re-encountered. Moreover, Zahorik and colleagues, and our own data (see Figure 4A), demonstrate that improvements in speech understanding are always greater in reverberant compared to anechoic, environments (Brandewie & Zahorik, 2010, 2013; Srinivasan & Zahorik, 2012), i.e., improved performance arises only through exposure to a reverberant environment rather than to the speech material per se.

Adaptation to continuous noise is a well-known phenomenon in the auditory system (Costalupes et al., 1984; Gibson et al., 1985; Phillips, 1985; Rees & Palmer, 1988) that could potentially also explain the speech improvements observed among carrier lengths. This is particularly relevant in our experiments because the masking noise preceded the onset of the speech material, potentially providing a window for noise-adaptation to occur (Ainsworth & Meyer, 1994; Ben-David et al., 2016). However, Zahorik and Brandewie reported similar effect sizes in speech enhancements when 1s (Brandewie & Zahorik, 2010) or 150 ms (Brandewie & Zahorik, 2013) of white noise was presented prior to the carrier phrase (i.e., CP0—|Color| |Number|). From this the authors concluded that noise alone did not convey sufficient information about the acoustic environment to elicit improvements in speech understanding.

Adult listeners can also acclimatise to speech acoustics that deviate from the norm or from long-term language regularities such as dialects and foreign accents, with sufficient exposure (Idemaru & Holt, 2011, 2014; Liu & Holt, 2015). Even if our participants were all Australian-English native speakers exposed to an American-English speech corpus—the CRM—some perceptual learning of individual talker’s idiosyncratic speech patterns might arise over the course of the task (Choi & Perrachione, 2019; Liu & Holt, 2015; Stilp, 2020). In addition, it has been reported that ‘target voice continuity’ i.e., the same talker within trials—similar to our design—can enhance the build-up of selective attention and improve listeners abilities to report correctly digits sequences in the presence of other talkers (Best et al., 2008). However, when the room build-up effect—the effect assessed here—is explored in an isolated anechoic space i.e., not intermingled with room acoustic of other reverberant spaces, there is little or no improvement in speech understanding with increasing exposure to the talker per se; i.e., in the form of an increasing length of carrier phrase (Brandewie & Zahorik, 2010, 2013). Interrupting the continuity of the room—it’s reverberation profile—rather than the continuity of the target talker, is the factor reported to disrupt significantly the improvement in speech understanding with increasing exposure (Brandewie & Zahorik, 2018). If listeners adapted to a talker in the course of our experiment, this occurred rapidly, within a few speech tokens (Cooke et al., 2022; Kakehi, 1992; Kato & Kakehi, 1988) and contributed little to the enhancement of speech understanding we observed in reverberant rooms.

A key feature of our study, one that distinguishes it from previous assessments of statistical learning of acoustic features in human listeners, is our use of an ethologically valid listening task—understanding speech in background noise. Given the perceptual biases inherent in sensory processing, it is notable that investigators have previously reported different noise tokens to be differently learnable (Agus et al., 2014; Daikhin et al., 2017), for example, and that textures must be carefully controlled to ensure listeners are not exploiting subtle spectro-temporal features in their judgments of statistical similarity (McDermott et al., 2013). The potential for listeners to hear out specific—potentially unique to them—spectro-temporal features of otherwise statistically identical sound tokens, or to make judgments on the regularity or similarity of tone sequences based on unique listening experiences (Barascud et al., 2016; Bianco et al., 2020), presents a potential confound for these studies. Our study countermands this problem by actively exploiting the propensity for human listeners to ascribe meaning to spectro-temporal fluctuations (here, speech) in acoustic waveforms (Brandewie & Zahorik, 2010, 2013). Though still targeting the learning of background acoustic features, it engages listeners in a highly relevant listening task, one for which humans have rapidly evolved—i.e., understanding speech in background noise. Listeners’ attention, therefore, was called away from the background acoustic features to the ethologically relevant foreground task of attending to human speech, void of semantics; additionally, length of phrase, talker, and |Callsign| were uninformative to |Color| and |Number| in our utterances, as were |Color| and |Number| to each other. This misdirection allows us to exploit speech as a ‘reporter’ or biomarker for statistical learning of background, more-abstract, acoustic features, and without the potential confounding factor of (individualised) perceptual bias. Further, we have demonstrated that prior exposure to ethologically relevant environments matters when learning less-commonly encountered acoustic scenes.

The impact on speech understanding of less-ecologically realistic reverberant environments is reminiscent of the detrimental effect of TMS, with lower overall listening performance. This suggests that learning the statistical structure of sound environments, the better to understand speech in reverberant background noise, requires some longer form of memory (experience, developmental, or evolutionary). Brain circuits, therefore, might act to improve listening performance, operating best when at least some of the exposure to listening environments includes plausible, and commonly experienced, reverberation times (Traer & McDermott, 2016). When the acoustics of commonly encountered rooms were replaced with those of other environments having higher or lower amounts of reverberation, performance declined over the course of an experimental session. Intriguingly, the most effective reverberation time we employed, at least in terms of its capacity to be learned and potentially from which learning could be transferred—the RT60traer of 0.49 s of the small Lecture Room is very close to the median and mean RT60traer of 0.42 and 0.50 s previously recorded for a wide range of built environments (indoors and outdoors). This suggests that statistical learning of acoustic background features might be tuned to ethologically relevant environments. Alternatively, built environments might be constructed to generate reverberation subjectively most suited to maximising speech-in-noise understanding, though we are unaware as to whether such a constraint is consciously applied in the design of listening spaces given contingencies such as the range and relative consistency and constancy of potential contents of any given space that influence its reverberant characteristics. Compared to built environments, however, the RT60traers of natural environments were much lower than all the rooms we assessed, save for that of the anechoic room, which was suboptimal in terms of listening performance.

Neural adaptation as a contributing mechanism to the learning of room acoustics

Our data are consistent with improved performance in a relevant ‘foreground’ task emerging as the brain adapts to the statistics of background features of the listening environment. Adaptation to stimulus statistics has been proposed as a neurophysiological mechanism underlying selective suppression of background noise in auditory cortex (Fuglsang et al., 2017; Kell & McDermott, 2019; Khalighinejad et al., 2019; Mesgarani et al., 2014). More recently, it has been shown that spectro-temporal receptive fields of auditory-cortical neurons are sensitive to the RT60s of reverberant noise i.e., a form of adaptation specific to reverberation and consistent with a de-reverberation of the environment encountered (Ivanov et al., 2022). Cortical adaptation to reverberation has been also reported in awake listeners (Fuglsang et al., 2017; Mesgarani et al., 2014) and to this end, our data suggest a role for top-down mechanisms in improving over time an attended task—recalling spoken words—in reverberant environments. It is therefore possible that goal-directed behaviours and the feedback the auditory cortex receives form areas such as dlPFC may contribute to fine-tuning the adaptation to reverberant environments.

Our data are also consistent with reports of midbrain auditory neurons recorded in vivo adapting to the statistical structure of evolving sound environments over the course of several hundred milliseconds (Dean et al., 2005, 2008) and this capacity for adaptation increases with repeated exposures to the same statistically structured environment. Importantly, whilst rapid adaptation is retained when descending cortical influences are disrupted (through cortical cooling), the longer-term learning effect—a speeding up of adaptation referred to as meta-adaptation—is abolished (Robinson et al., 2016). As with auditory neurons recorded in vivo (Bakay et al., 2018; Dean et al., 2005, 2008; Ivanov et al., 2022), it seems listeners might adapt to the statistical structure of a sound environment within a few hundred milliseconds of exposure to enhance performance in a listening task, an initial, rapid, learning phase impervious to inactivation of the dorsolateral prefrontal cortex by theta-burst TMS. The nested set of temporal sensitivities to room acoustics we observe is consistent with features of statistical learning that emerges from the level of the auditory nerve to primary cortex in vivo (Dean et al., 2005, 2008; Watkins & Barbour, 2008; Wen et al., 2009), and likely modulated by efferent influences that span the entire auditory pathway all the way to the sensory receptors of the inner ear (Hernández-Pérez et al., 2021; Perrot et al., 2006; Terreros & Delano, 2015) and even the mechanical sensitivity to sound of the eardrum and ossicles (middle-ear bones) (Gruters et al., 2018).

Dissociating auditory cortical feedback circuits to the midbrain impairs the capacity of midbrain neurons to ‘recall’ previous experience of that environment (Bajo et al., 2019; Robinson et al., 2016). Like the performance in our listening tasks, these neurons were still able to adapt rapidly (within hundreds of milliseconds) each time an environment was encountered as if for the first time, but showed no capacity to exploit a memory of that environment to improve neural coding (i.e., meta-adaptation) (Robinson et al., 2016). Meta-adaptation was consistent with neurons in lower brain centres adapting to the current sound environment, and those in higher brain centres learning the longer-term statistical structure of changing environments. Once higher brain centres learnt the experienced environment, this information, conveyed to early brain centres, ensures they were ‘primed’ for coding an environment when it is re-encountered.

Inclusion and Ethics statement

This study was approved by the Human Research Ethics Committee of Macquarie University (ref: 5201833344874). Each participant signed a written informed consent form and was given a small financial remuneration for their time.

eLife Assessment:

This study addresses valuable questions about the neural mechanisms underlying statistical learning of room acoustics, combining robust behavioral measures with non-invasive brain stimulation. The behavioral findings are strong and extend previous work in psychoacoustics, but the TMS results are modest, with methodological limitations and over-interpretation that weaken the mechanistic conclusions. The strength of evidence is therefore incomplete, and a more cautious interpretation of the stimulation findings, alongside strengthened analyses, would improve the manuscript.

Reviewer #1 (Public review):

Summary:

This manuscript describes the results of an experiment that demonstrates a disruption in statistical learning of room acoustics when transcranial magnetic stimulation (TMS) is applied to the dorsolateral prefrontal cortex in human listeners. The work uses a testing paradigm designed by the Zahorik group that has shown improvement in speech understanding as a function of listening exposure time in a room, presumably through a mechanism of statistical learning. The manuscript is comprehensive and clear, with detailed figures that show key results. Overall, this work provides an explanation for the mechanisms that support such statistical learning of room acoustics and, therefore, represents a major advancement for the field.

Strengths:

The primary strength of the work is its simple and clear result, that the dorsolateral prefrontal cortex is involved in human room acoustic learning.

Weaknesses:

A potential weakness of this work is that the manuscript is quite lengthy and complex.

Reviewer #2 (Public review):

Summary:

This study investigated how listeners adapt to and utilize statistical properties of different acoustic spaces to improve speech perception. The researchers used repetitive TMS to perturb neural activity in DLPFC, inhibiting statistical learning compared to sham conditions. The authors also identified the most effective room types for the effective use of reverberations in speech in noise perception, with regular human-built environments bringing greater benefits than modified rooms with lower or higher reverberation times.

Strengths:

The introduction and discussion sections of the paper are very interesting and highlight the importance of the current study, particularly with regard to the use of ecologically valid stimuli in investigating statistical learning. However, they could be condensed into parts. TMS parameters and task conditions were well-considered and clearly explained.

Weaknesses

(1) The Results section is difficult to follow and includes a lot of detail, which could be removed. As such, it presents as confusing and speculative at times.

(2) The hypotheses for the study are not clearly stated.

(3) Multiple statistical models are implemented without correcting the alpha value. This leaves the analyses vulnerable to Type I errors.

(4) It is confusing to understand how many discrete experiments are included in the study as a whole, and how many participants are involved in each experiment.

(5) The TMS study is significantly underpowered and not robust. Sample size calculations need further explanation (effect sizes appear to be based on behavioural studies?). I would caution an exploratory presentation of these data, and calculate a posteriori the full sample size based on effect sizes observed in the TMS data.

Reviewer #3 (Public review):

Summary:

This manuscript presents a well-designed and insightful behavioural study examining human adaptation to room acoustics, building on prior work by Brandewie & Zahorik. The psychophysical results are convincing and add incremental but meaningful knowledge to our understanding of reverberation learning. However, I find the transcranial magnetic stimulation (TMS) component to be over-interpreted. The TMS protocol, while interesting, lacks sufficient anatomical specificity and mechanistic explanation to support the strong claims made regarding a unique role of the dorsolateral prefrontal cortex (dlPFC) in this learning process. More cautious interpretation is warranted, especially given the modest statistical effects, the fact that the main TMS result of interest is a null result, the imprecise targeting of dlPFC (which is not validated), and the lack of knowledge about the timescale of TMS effects in relation to the behavioural task. I recommend revising the manuscript to shift emphasis toward the stronger behavioural findings and to present a more measured and transparent discussion of the TMS results and their limitations.

Strengths:

(1) Well-designed acoustical stimuli and psychophysical task.

(2) Comparisons across room combinations are well conducted.

(3) The virtual acoustic environment is impressive and applied well here.

(4) A timely study with interesting behavioural results.

Weaknesses:

(1) Lack of hypotheses, particularly for TMS.

(2) Lack of evidence for targeting TMS in [brain] space and time.

(3) The most interesting effect of TMS is a null result compared to a weak statistical effect for "meta adaptation"

Reviewer #4 (Public review):

Summary:

Several behavioral experiments and one TMS experiment were performed to examine adaptation to room reverberation for speech intelligibility in noise. This is an important topic that has been extensively studied by several groups over the years. And the study is unique in that it examines one candidate brain area, dlPFC, potentially involved in this learning, and finds that disrupting this area by TMS results in a reduction in the learning. The behavioral conditions are in many ways similar to previous studies. However, they find results that do not match previous results (e.g., performance in anechoic condition is worse than in reverberation), making it difficult to assess the validity of the methods used. One unique aspect of the behavioral experiments is that Ambisonics was used to simulate the spaces, while headphone simulation was mostly used previously. The main behavioral experiment was performed by interleaving 3 different rooms and measuring speech intelligibility as a function of the number of words preceding the target in a given room on a given trial. The findings are that performance improves on the time scale of seconds (as the number of words preceding the target increases), but also on a much larger time scale of tens to hundreds of seconds (corresponding to multiple trials), while for some listeners it is degraded for the first couple of trials. The study also finds that the performance is best in the room that matches the T60 most commonly observed in everyday environments. These are potentially interesting results. However, there are issues with the design of the study and analysis methods that make it difficult to verify the conclusions based on the data.

Strengths:

(1) Analysis of the adaptation to reverberation on multiple time scales, for multiple reverberant and anechoic environments, and also considering contextual effects of one environment interleaved with the other two environments.

(2) TMS experiment showing reduction of some of the learning effects by temporarily disabling the dlPFC.

Weaknesses:

While the study examines the adaptation for different carrier lengths, it keeps multiple characteristics (mainly talker voice and location) fixed in addition to reverberation. Therefore, it is possible that the subjects adapt to other aspects of the stimuli, not just to reverberation. A condition in which only reverberation would switch for the target would allow the authors to separate these confounding alternatives. Now, the authors try to address the concerns by indirect evidence/analyses. However, the evidence provided does not appear sufficient.

The authors use terms that are either not defined or that seem to be defined incorrectly. The main issue then is the results, which are based on analysis of what the authors call d', Hit Rate, and Final Hit rate. First of all, they randomly switch between these measures. Second, it's not clear how they define them, given that their responses are either 4-alternative or 8-alternative forced choice. d', Hit Rate, and False Alarm Rate are defined in Signal detection theory for the detection of the presence of a target. It can be easily extended to a 2-alternative forced choice. But how does one define a Hit, and, in particular, a False Alarm, in a 4/8-alternative? The authors do not state how they did it, and without that, the computation of d' based on HR and FAR is dubious. Also, what the authors call Hit Rate, is presumably the percent correct performance (PCC), but even that is not clear. Then they use FHR and act as if this was the asymptotic value of their HR, even though in many conditions their learning has not ended, and randomly define a variable of +-10 from FHR, which must produce different results depending on whether the asymptote was reached or not. Other examples of usage of strange usage of terms: they talk about "global likelihood learning" (L426) without a definition or a reference, or about "cumulative hit rate" (L1738), where it is not clear to me what "cumulative" means there.

There are not enough acoustic details about the stimuli. The authors find that reverberant performance is overall better than anechoic in 2 rooms. This goes contrary to previous results. And the authors do not provide enough acoustic details to establish that this is not an artefact of how the stimuli were normalized (e.g., what were the total signal and noise levels at the two ears in the anechoic and reverberant conditions?).

There are some concerns about the use of statistics. For example, the authors perform two-way ANOVA (L724-728) in which one factor is room, but that factor does not have the same 3 levels across the two levels of the other factor. Also, in some comparisons, they randomly select 11 out of 22 subjects even though appropriate test correct for such imbalances without adding additional randomness of whether the 11 selected subjects happened to be the good or the bad ones.

Details of the experiments are not sufficiently described in the methods (L194-205) to be able to follow what was done. It should be stated that 1 main experiment was performed using 3 rooms, and that 3 follow-ups were done on a new set of subjects, each with the room swapped.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

This manuscript describes the results of an experiment that demonstrates a disruption in statistical learning of room acoustics when transcranial magnetic stimulation (TMS) is applied to the dorsolateral prefrontal cortex in human listeners. The work uses a testing paradigm designed by the Zahorik group that has shown improvement in speech understanding as a function of listening exposure time in a room, presumably through a mechanism of statistical learning. The manuscript is comprehensive and clear, with detailed figures that show key results. Overall, this work provides an explanation for the mechanisms that support such statistical learning of room acoustics and, therefore, represents a major advancement for the field.

Strengths:

The primary strength of the work is its simple and clear result, that the dorsolateral prefrontal cortex is involved in human room acoustic learning.

Weaknesses:

A potential weakness of this work is that the manuscript is quite lengthy and complex.

Reviewer #2 (Public review):

Summary:

This study investigated how listeners adapt to and utilize statistical properties of different acoustic spaces to improve speech perception. The researchers used repetitive TMS to perturb neural activity in DLPFC, inhibiting statistical learning compared to sham conditions. The authors also identified the most effective room types for the effective use of reverberations in speech in noise perception, with regular human-built environments bringing greater benefits than modified rooms with lower or higher reverberation times.

Strengths:

The introduction and discussion sections of the paper are very interesting and highlight the importance of the current study, particularly with regard to the use of ecologically valid stimuli in investigating statistical learning. However, they could be condensed into parts. TMS parameters and task conditions were well-considered and clearly explained.

Weaknesses

(1) The Results section is difficult to follow and includes a lot of detail, which could be removed. As such, it presents as confusing and speculative at times.

(2) The hypotheses for the study are not clearly stated.

(3) Multiple statistical models are implemented without correcting the alpha value. This leaves the analyses vulnerable to Type I errors.

(4) It is confusing to understand how many discrete experiments are included in the study as a whole, and how many participants are involved in each experiment.

(5) The TMS study is significantly underpowered and not robust. Sample size calculations need further explanation (effect sizes appear to be based on behavioural studies?). I would caution an exploratory presentation of these data, and calculate a posteriori the full sample size based on effect sizes observed in the TMS data.

Reviewer #3 (Public review):

Summary:

This manuscript presents a well-designed and insightful behavioural study examining human adaptation to room acoustics, building on prior work by Brandewie & Zahorik. The psychophysical results are convincing and add incremental but meaningful knowledge to our understanding of reverberation learning. However, I find the transcranial magnetic stimulation (TMS) component to be over-interpreted. The TMS protocol, while interesting, lacks sufficient anatomical specificity and mechanistic explanation to support the strong claims made regarding a unique role of the dorsolateral prefrontal cortex (dlPFC) in this learning process. More cautious interpretation is warranted, especially given the modest statistical effects, the fact that the main TMS result of interest is a null result, the imprecise targeting of dlPFC (which is not validated), and the lack of knowledge about the timescale of TMS effects in relation to the behavioural task. I recommend revising the manuscript to shift emphasis toward the stronger behavioural findings and to present a more measured and transparent discussion of the TMS results and their limitations.

Strengths:

(1) Well-designed acoustical stimuli and psychophysical task.

(2) Comparisons across room combinations are well conducted.

(3) The virtual acoustic environment is impressive and applied well here.

(4) A timely study with interesting behavioural results.

Weaknesses:

(1) Lack of hypotheses, particularly for TMS.

(2) Lack of evidence for targeting TMS in [brain] space and time.

(3) The most interesting effect of TMS is a null result compared to a weak statistical effect for "meta adaptation"

Reviewer #4 (Public review):

Summary:

Several behavioral experiments and one TMS experiment were performed to examine adaptation to room reverberation for speech intelligibility in noise. This is an important topic that has been extensively studied by several groups over the years. And the study is unique in that it examines one candidate brain area, dlPFC, potentially involved in this learning, and finds that disrupting this area by TMS results in a reduction in the learning. The behavioral conditions are in many ways similar to previous studies. However, they find results that do not match previous results (e.g., performance in anechoic condition is worse than in reverberation), making it difficult to assess the validity of the methods used. One unique aspect of the behavioral experiments is that Ambisonics was used to simulate the spaces, while headphone simulation was mostly used previously. The main behavioral experiment was performed by interleaving 3 different rooms and measuring speech intelligibility as a function of the number of words preceding the target in a given room on a given trial. The findings are that performance improves on the time scale of seconds (as the number of words preceding the target increases), but also on a much larger time scale of tens to hundreds of seconds (corresponding to multiple trials), while for some listeners it is degraded for the first couple of trials. The study also finds that the performance is best in the room that matches the T60 most commonly observed in everyday environments. These are potentially interesting results. However, there are issues with the design of the study and analysis methods that make it difficult to verify the conclusions based on the data.

Strengths:

(1) Analysis of the adaptation to reverberation on multiple time scales, for multiple reverberant and anechoic environments, and also considering contextual effects of one environment interleaved with the other two environments.

(2) TMS experiment showing reduction of some of the learning effects by temporarily disabling the dlPFC.

Weaknesses:

While the study examines the adaptation for different carrier lengths, it keeps multiple characteristics (mainly talker voice and location) fixed in addition to reverberation. Therefore, it is possible that the subjects adapt to other aspects of the stimuli, not just to reverberation. A condition in which only reverberation would switch for the target would allow the authors to separate these confounding alternatives. Now, the authors try to address the concerns by indirect evidence/analyses. However, the evidence provided does not appear sufficient.

The authors use terms that are either not defined or that seem to be defined incorrectly. The main issue then is the results, which are based on analysis of what the authors call d', Hit Rate, and Final Hit rate. First of all, they randomly switch between these measures. Second, it's not clear how they define them, given that their responses are either 4-alternative or 8-alternative forced choice. d', Hit Rate, and False Alarm Rate are defined in Signal detection theory for the detection of the presence of a target. It can be easily extended to a 2-alternative forced choice. But how does one define a Hit, and, in particular, a False Alarm, in a 4/8-alternative? The authors do not state how they did it, and without that, the computation of d' based on HR and FAR is dubious. Also, what the authors call Hit Rate, is presumably the percent correct performance (PCC), but even that is not clear. Then they use FHR and act as if this was the asymptotic value of their HR, even though in many conditions their learning has not ended, and randomly define a variable of +-10 from FHR, which must produce different results depending on whether the asymptote was reached or not. Other examples of usage of strange usage of terms: they talk about "global likelihood learning" (L426) without a definition or a reference, or about "cumulative hit rate" (L1738), where it is not clear to me what "cumulative" means there.

There are not enough acoustic details about the stimuli. The authors find that reverberant performance is overall better than anechoic in 2 rooms. This goes contrary to previous results. And the authors do not provide enough acoustic details to establish that this is not an artefact of how the stimuli were normalized (e.g., what were the total signal and noise levels at the two ears in the anechoic and reverberant conditions?).

There are some concerns about the use of statistics. For example, the authors perform two-way ANOVA (L724-728) in which one factor is room, but that factor does not have the same 3 levels across the two levels of the other factor. Also, in some comparisons, they randomly select 11 out of 22 subjects even though appropriate test correct for such imbalances without adding additional randomness of whether the 11 selected subjects happened to be the good or the bad ones.

Details of the experiments are not sufficiently described in the methods (L194-205) to be able to follow what was done. It should be stated that 1 main experiment was performed using 3 rooms, and that 3 follow-ups were done on a new set of subjects, each with the room swapped.

We sincerely thank the Editor and the Reviewers for their careful evaluation of our manuscript and for their constructive and insightful comments. We greatly appreciate the time and expertise invested in reviewing our work. The feedback has been invaluable in improving the clarity, rigor, and overall presentation of the manuscript.

In response to the reviewers’ comments, we have carefully revised the manuscript throughout. The revisions include clarification of the study hypotheses, re-analysis of the TMS data using a mixed ANOVA framework, additional methodological details regarding the TMS procedures and behavioural analyses, expanded justification of the statistical modelling approach, clarification of the room-acoustics paradigm, revision of figures and figure legends, additional discussion of study limitations, and a more balanced interpretation of the TMS findings. We have also substantially revised the Discussion section and improved the overall structure and readability of the manuscript.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) It is understood that this topical area is necessarily detail-heavy, but if there are ways to streamline the manuscript to more quickly arrive at key results (Figure 3?), the work might have an even greater overall impact.

We appreciate the reviewer’s feedback and have carefully revised the manuscript to address all comments from all reviewers. However, we have retained the existing order of figures and results to maintain consistency and avoid extensive structural changes that could compromise the clarity and flow of the manuscript.

(2) Minor point: I believe Equation 1 should be d' = z(H) – z(F).

Thanks for noticing this, we have fixed the equation.

Reviewer #2 (Recommendations for the authors):

(1) Line 115: Towards the end of the introduction, the hypotheses for the current study remain unclear. Please explicitly outline each hypothesis before the Methods section.

We have now specified the hypotheses tested at the end of the Introduction.

(2) Line 229: Please state the minimum MEP amplitude criterion used during TMS thresholding (usually 50 µV). Please also reference the EMG hardware and software used to measure MEPs.

We are grateful to the reviewer for noticing that a few details regarding our TMS procedure were missing in the “Continuous theta-burst stimulation” section of the Methods. We have extensively revised this section to include the required details. We did not, however, use MEP amplitude as a criterion for estimating motor thresholds. For single-pulse TMS-induced motor threshold determination, we used visual observation of the first dorsal interosseous (FDI) muscle twitch (Varnava et al., 2011).

(3) Line 248: The Coordinate Response Measure corpus contains combinations of callsigns, colours, and numbers. What was the rationale in asking participants to only identify the colour and the number spoken, and not the callsign?

The rationale for reporting only Color and Number was to ensure an equal number of keyword identifications across phrase lengths—for example, CP0 and CP1 do not contain a callsign. This has now been clarified in the Procedure section of the Methods.

(4) Lines 327–334: The analyses outlined here are unclear – it appears that there are multiple different statistical tests being conducted on the same outcome variable in this study. Given that the design includes a between-subjects factor (TMS condition) and two within-subjects factors (Rooms, Carrier Length Phrase), could the analyses be simplified by employing a mixed ANOVA as opposed to separate repeated-measures and univariate ANOVAs? If this is, in fact, the analysis that was conducted, please improve the wording. Clarification/correction is also needed surrounding the use of the term “univariate” ANOVA, as this can refer to different statistical tests.

We appreciate the reviewer pointing this out. We have re-analysed the TMS data using a mixed ANOVA design and reported it as such in the Results section. Although the numerical values have changed, the significant findings and conclusions remain the same.

(5) Line 340: Whilst the authors identify that an alpha value of 0.05 and Bonferroni corrections were used for statistical inference in two-tailed t-tests, there is no indication of the alpha value used in the interpretation of the ANOVA results. Please include this before the Results section. Furthermore, given that multiple tests are being conducted in this project, the alpha value used to infer statistical significance should be corrected in accordance with the number of hypotheses, to reduce Type I error rates (e.g., 4 hypotheses would result in an alpha inference criterion of α = 0.0125).

We appreciate the reviewer’s observation. The corrected alpha value was not explicitly reported because all statistical analyses were conducted using IBM SPSS Statistics for Windows, Version 29.0.2.0 (IBM Corp., Armonk, NY; RRID:SCR_002865). In SPSS, Bonferroni corrections are applied by adjusting the p-values rather than the alpha threshold itself. For transparency, we have now clarified in the manuscript the factors included in each statistical comparison to make the tested hypotheses fully explicit.

(6) Line 346: The sample size calculation used in the current study could be improved. It is unclear why a sample size estimation of ≥18 is used when the actual sample recruited is significantly greater than this (62). Is this because 62 participants were divided across multiple experiments in this paper? Further clarification is needed. The alpha value used in this calculation should also be corrected to account for multiple statistical models.

We thank the reviewer for noting this. We have clarified that the initial sample size estimation (n = 18) referred to individual ANOVA analyses. In the revised manuscript, we specify in each experimental section (Identity of Sound Environments and Continuous Theta Stimulation) the exact number of participants recruited per experiment, which together sum to a total of 74 participants across all experiments (also clarified in the Participants section of the Materials and Methods).

(8) Line 438: Many of the statistical tests presented in the Results section have not been outlined in the Methods section or had the rationale explained in the Introduction – this makes the analyses feel confusing and speculative. Explicitly identifying the core hypotheses earlier in the manuscript and clearly stating which hypotheses are confirmatory or exploratory would be an important improvement.

We appreciate the reviewer noticing this. We have clarified the hypotheses tested in the Introduction (4th and 5th paragraphs) and provided a more detailed account of the statistical analyses in the Materials and Methods (Statistical Analysis section).

(9) Line 565: It is unclear how the 62 participants recruited in the study were divided across each of the experimental conditions. I would recommend expanding the Participants section in Methods to outline the number of participants involved in each stage of the study.

Thanks for noticing this discrepancy—this calculation was indeed confusing and incorrect. We ultimately tested a total of 74 participants. We have specified the sample size per experiment in both the “Identity of selected sound environments” and “Continuous theta-burst stimulation” sections of the Methods, and reiterated this in the Results section to prevent confusion.

(9) Line 792: The Results section as a whole is incredibly complex and lacks structure. This could be condensed significantly – details regarding previous research should be removed from Results, as this should already be outlined in the Introduction as rationale for the current project. It would be clearer to outline each confirmatory and exploratory hypothesis in the Introduction, then identify at each stage in the Results section where it is being tested.

We appreciate the reviewer’s point. We have rewritten the 4th and 5th paragraphs of the Introduction to outline the confirmatory and exploratory hypotheses. While we have streamlined parts of the Results section to enhance clarity, we have retained contextual information for each analysis to help readers follow the logic of the findings, given the complexity and scope of the study.

Reviewer #3 (Recommendations for the authors):

(1) Introduction – Overinterpretation and Hypothesis Clarity. The final paragraph of the Introduction (page 4) discusses the experimental findings and their interpretation, which belongs in the Discussion. This section should instead clearly state hypotheses for both the behavioural and TMS experiments. In particular, the TMS experiment lacks a clear rationale: what mechanism is being tested, and what behavioural outcome is predicted? Please revise this section to focus on the theoretical motivation, clearly defined hypotheses, and expected results, and less on summarizing the results.

We appreciate this point and have accordingly deleted the final paragraph of the Introduction, as it is already incorporated into the Discussion section. We have stated the hypotheses/rationale more clearly in the Introduction.

(2) TMS – Mechanism, Timeframe, and Clarity. The manuscript does not adequately explain how TMS produces long-lasting effects relevant to the task, which occur minutes (or possibly longer) after stimulation.

(a) What is the specific timeframe between stimulation and behavioural testing?

The timeframe between stimulation and behavioural testing has been clarified in the Methods section, e.g.: “Participants exposed to ‘real’ or ‘sham’ TMS completed the familiarization and behavioural task right after cTBS procedures.”

(b) What is the evidence that TMS to the prefrontal cortex affects function on this timescale?

We have added the following paragraph to the Discussion to address the timescale of continuous theta-burst stimulation (cTBS): cTBS, as used in our study, typically induces aftereffects lasting 20–50 minutes (Huang et al., 2005; Wischnewski & Schutter, 2015). While these effects are well established in the motor cortex—with motor-evoked potential changes persisting for up to one hour—recent evidence suggests that similar durations of cortical modulation can also occur in the prefrontal cortex (Taylor et al., 2025). Specifically, studies applying inhibitory rTMS to the dorsolateral prefrontal cortex (dlPFC) during cognitive tasks have demonstrated functional effects lasting up to one hour in healthy participants (Wagner et al., 2006). Furthermore, Tupak et al. (2013) showed that inhibitory rTMS to the dlPFC leads to reduced oxygenation levels, reflecting decreased cortical activity, for at least 45 minutes—the same duration as the experimental task in our study. Given that changes in cerebral haemoglobin concentration closely correspond to neuronal activation (Liao et al., 2013), the fNIRS-measured alterations in local cerebral blood oxygenation provide an indirect but reliable indicator of TMS-induced neural modulation within this timescale.

(c) Can post-stimulation effects be objectively measured or confirmed?

Although no objective post-stimulation measures were collected in the present study, we acknowledge this as a limitation. However, previous research has shown that inhibitory rTMS to the dlPFC leads to reduced oxygenation levels—reflecting decreased cortical activity—for at least 45 minutes (Tupak et al., 2013). Given that changes in cerebral haemoglobin concentration closely correspond to neuronal activation (Liao et al., 2013), these findings indicate that fNIRS can serve as an indirect but reliable method for confirming TMS-induced neural modulation. We plan to incorporate such objective measures in future studies.

We thank the reviewer for raising these important points and have addressed them by updating the Methods and Results sections and adding a paragraph to the Discussion. We would like to clarify, however, that the TMS effects observed in our study are not weak: the statistically significant differences between sham and TMS conditions were accompanied by large effect sizes, indicating that bilateral inhibitory stimulation of the dlPFC produced a robust and consistent effect across participants—specifically, a reduction in performance, reflecting decreased improvement in speech understanding with increasing exposure to the reverberant environment.

(3) Figure 3A – Anatomical Specificity and Interpretation. Figure 3A implies precise stimulation of dlPFC and its projections to auditory cortex (A1), but the authors cannot actually target dlPFC or its connections specifically with this approach. Rather, the TMS protocol disrupts an undetermined region of PFC, with diffuse downstream effects. This should be clearly acknowledged in the figure legend and main text.

We appreciate this comment and have acknowledged this point in the figure legend and Discussion, e.g., Figure 3 legend: “Although the TMS protocol was intended to target the dlPFC, it likely affected adjacent prefrontal regions, leading to diffuse downstream effects that may have included modulation of A1.”

(4) Additionally, the Discussion overstates the evidence for a specific dlPFC → AC role in reverberation learning. The weak and poorly localized TMS effect does not support strong claims about this pathway. Please scale back this interpretation and focus more on the robust psychophysical results, which are the manuscript's stronger contribution.

We thank the reviewer for this comment. We have revised our interpretation to clarify that the proposed dlPFC–auditory cortex link is speculative, and have added caveats regarding the limited spatial precision of TMS targeting and individual variability in its effects, in the Discussion section “A role for dlPFC in statistical learning of room acoustics.”

(4a) Line 152: Extra comma after “of”? Also, why are there square brackets around “callsigns” etc.?

Fixed.

(4b) Line 156: “RRID:SCR_001622” is unexplained and likely unnecessary—consider removing.

Removed.

(4c) Line 159: Why was no ramping applied at the end of the noise? Please clarify.

Similar to Brandewie & Zahorik (2013), no ramping was applied. This has been clarified in the Methods (“Acoustic Stimuli”).

(4d) Line 207: Methods do not describe the sham TMS protocol—please add this information.

Thanks for noticing this. Information on the sham TMS protocol has been added.

(4e) Line 352: Fix bracket formatting.

Fixed.

(4f) Line 382: Sentence is grammatically incorrect—please revise.

Fixed.

(4g) Line 394: Unclear use of square brackets—clarify or standardize.

Fixed.

(4h) Line 401: It is unclear how interleaving the talker and length ensures a different room each trial. Aren't these variables independent?

The reviewer is correct: the only variable that was pseudorandomized was room order, to prevent carry-over effects, similar to Brandewie & Zahorik (2013). This has been clarified in the Methods section.

(4i) Figure 1: Clarify that AI-generated images refer only to the room images, not other components.

Fixed.

(4j) Figure 2A: Confirm that “overall” includes all speakers and durations—clarify in legend.

Fixed.

(4k) The interesting duration effects in Figure 1D are not discussed in the text and appear before overall room effects in Figure 2A—please reorder and comment on these results.

Fixed.

(4l) Supplementary Figure 2: Caption contains a typo (“Lecture Room/Open-Plan Office”).

Typo has been fixed.

Also, I recommend adding this result to the main figure set—e.g., include overall d′ for all six talkers in Figure 2 alongside rooms (2A) and lengths (2D).

We thank the reviewer for the suggestion. Including overall performance for all six talkers in Figure 2 would require substantial restructuring and risk making the figure crowded. We have therefore retained these results in the Supplementary Materials (Supplementary 2 and 3), as originally presented.

(4m) Line 555: Phrase “to better understand” could be clearer—consider rewording.

This section has been reworded.

(4n) Lines 587–592: The lack of main effect of TMS is helpful, but more important is whether interactions between TMS and room/length variables occur. Please report these interactions, as they are central to interpreting the TMS effects.

We appreciate the reviewer highlighting this. We have reviewed this analysis and reported the interaction Condition x CP length as follows: “A mixed ANOVA with a between-group factor of Condition and within-group factors of room and CP length confirmed that performance in these two populations was comparable, with no significant main effect of TMS conditions observed (‘sham’ vs. no exposure to TMS): [F (1,31) =0.01, p=0.91, ŋp2 = 0.00], and no significant interaction Condition x CP length was observed: [F (3,93) =0.48, p=0.69, ŋp2 = 0.01]; confirming that participants experiencing ‘sham’ TMS did not perform significantly differently from the ‘no exposure to TMS’ population”.

(4o) Line 601: Reiterate the timeframe of the TMS-behaviour gap. Is there supporting evidence that TMS can affect behaviour over this duration? Could null effects reflect fading TMS efficacy?

We appreciate the reviewer pointing this out. We have clarified the timeframe of TMS stimulation in both the Methods and Results, e.g.: “The procedure began with TMS manipulation, and although the behavioural task lasted 45 minutes, the inhibitory effects of TMS extended for at least 60 minutes post-stimulation (Huang et al., 2005; Gamboa et al., 2010; Hoogendam, Ramakers, & Di Lazzaro, 2010; Romero et al., 2022).” However, we cannot dismiss individual differences in the duration of TMS effects, nor differences in efficacy duration between anatomical areas (motor cortex vs. dlPFC). We have noted this in the Discussion section “A role for dlPFC in statistical learning of room acoustics” (Pallant, 2011).

(4p) Figure 3I: The “meta-adaptation” effect is marginal in both Exp 1 (p = 0.03) and Exp 2 sham (p = 0.04). These should be interpreted cautiously, given their statistical fragility.

We appreciate this comment. We have now calculated effect sizes for all Wilcoxon signed-rank tests (Pearson’s r) and report them. For the two comparisons noted by the reviewer, the effect sizes are medium (Exp 1) and large (Exp 2). We are therefore confident that, even where the p-values are not extremely low, the statistical differences are reliable.

(4q) Line 696: Reverberation is described as “common,” but it is nearly universal. Consider rephrasing to reflect this.

We appreciate this suggestion and have rephrased this line.

(4r) Line 816: The authors state that TMS reduced overall performance, but the earlier ANOVA (lines 587–592) shows no such effect. Please correct this discrepancy.

We appreciate the reviewer noticing this. This section has been clarified: the lack of statistical significance at lines 587–592 relates to the comparison between a subset of ‘no-TMS-exposed’ listeners and ‘sham’-TMS-exposed listeners, made only to demonstrate the absence of placebo effects in the sham sample. Following the reviewers’ suggestions, we also re-analysed the data using a one-way ANOVA; this slightly changed the numerical values of the reported main effect but did not change the statistical outcome.

Reviewer #4 (Recommendations for the authors):

(1) Lines 201–202: It's not clear what is meant by combination and by carrier length here.

This section has been rewritten for clarity.

(2) Line 330: What is meant by “Univariate” here? I think this was a mixed ANOVA, with a betweensubject factor of TMS exposure and the remaining factors within-subject.

We appreciate this suggestion; we have re-analysed this section to use a mixed-ANOVA design. The numerical results differ, but the statistical outcome remains the same.

(3) Lines 335–343: This is impossible to follow if one does not understand that there were 3 followup experiments.

Thank you for highlighting that this section was confusing. We have rewritten it to clarify the following: “Three follow-up experiments were performed (univariate ANOVA) to assess whether speech understanding was affected by room context (i.e., the third room in which Open-Plan Office and Underground Car Park were learnt), with one between-subjects factor: room context (levels: Anechoic Room, Living Room, Lecture Room, and Highly Reflectant Room).” We have also added the following earlier in the Methods: “Additionally, we performed three follow-up experiments in different groups of listeners, assessing performance across combinations of three rooms: (1) Anechoic Room/Open-Plan Office/Underground Car Park (10 naïve listeners); (2) Living Room/Open-Plan Office/Underground Car Park (11 naïve listeners); and (3) Highly Reflectant Room/Open-Plan Office/Underground Car Park (10 naïve listeners). These conditions were used to determine whether a specific room combination was required to observe improvements in speech performance with increasing exposure to room acoustics i.e., with increasing carrier phrase length—and were assessed in the same way as Brandewie & Zahorik (2013).”

(4) Lines 414–416: The review of Tsironis et al. (2024) (doi:10.1177/23312165241273399) does not provide strong evidence that there are multiple scales for adaptation to room reverberation (most adaptation effects stabilize within 1 sec).

We apologize for this mistake, which arose from an issue with our reference manager. It has been corrected to: Robinson, Harper, & McAlpine (2016), Nature Communications, and Simpson, Harper, Reiss, & McAlpine (2014), Journal of Neuroscience.

(5) Line 427: The supplementary figure shows that many subjects did not achieve asymptotic performance.

We appreciate the reviewer pointing this out. The fittings have been extensively reviewed; please see the Methods and Results for the new fitting analysis. Indeed, some participants, although very slowly, keep improving over time without reaching clearly asymptotic behaviour. This section, however, referred specifically to the point at which performance stabilised within ±10% of final performance.

(6) Line 428: The ±10% statistic is random (as discussed below). And why switch to HR now? And what is its meaning when the HR did not converge by the end of the run?

We appreciate the reviewer raising the inconsistent use of d′ versus HR. d′ is referred to only in statistical analyses that do not bear a specific relation to the time-course analysis; time-course analysis does not allow us to calculate d′ at each trial or time point, owing to the lack of HR and FA values for single trials. We have clarified this in the Methods: “Given how d′ was calculated for |Color| and |Number|, it was not considered a useful metric for describing performance across time in different environments, owing to the paucity of data for each |Color| and |Number| affecting the temporal resolution of any generated curve. We therefore analysed the development of individual and average performance in each acoustic environment using cumulative hit rates, applying a 5-point moving average (~7 seconds) to each trace and plotting performance as a function of mean cumulative exposure time.”

We have also extensively reviewed our fits, following these steps: (i) we compared single- vs. double-exponential fits across 22 participants (Supplementary Figure 1), which showed that double exponentials provided a better R2 for the majority of participants; (ii) we re-ran all analyses forcing the fits to the HR endpoint; (iii) characterizing taus was not informative for our sample, given flat-like performance for some participants (Supplementary Figure 1)—in these cases, taus do not aid understanding of how performance stabilizes over time, particularly given the use of two taus; and (iv) cutting initial points differs by participant.

(7) Line 434: Or that they learned/adapted to other characteristics that were fixed.

Thank you—we have added “adapted to” in the sentence.

(8) Supplementary tables often show differences, but the actual values are not shown. Also, the tables and figures randomly switch between d' and HR.

Supplementary tables are intended only to show additional detail not reported in the main text or figures, to avoid redundancy; means (referred to in the Supplementary tables) are always shown in the main figures. We appreciate the reviewer raising the inconsistent use of d′ versus HR, addressed above, and have clarified this throughout the Methods and Results.

(9) Lines 436–442: There seem to be a lot of issues with the fitting shown in Supplemental Figure 1 and Figure 2B:

(a) It does not seem to converge, especially for the green line. So, presumably, the asymptotic value obtained for tau_slow was the upper bound set to 2000 s for many subjects' conditions. But those values are never shown—they should be in Supplemental Figure 1.

(b) Then the FHR value, derived from that, is completely dependent on what the bound was set to, and is therefore arbitrary. And its value of 10% is also arbitrary. Why do this when tau itself of an exponential model represents the time it takes to reach 67% of the asymptotic value, from which one can derive whatever time it should take to reach the final 10%?

(c) Even the use of the model specified by Equation 2 seems arbitrary. Average data in Figure 2B do not provide strong evidence for two time scales. If the authors are worried about the instability of the data at the beginning, a simple exponential with a weighted fit that prioritizes the later portions seems sufficient.

(d) Lines 430–434: This conclusion seems wrong, based only on the arbitrary measure chosen for “global likelihood learning.” Looking at Figure 2B, there is no evidence that the green graph reached any asymptote, while for the yellow and blue it appears to have. The authors should try fitting a simple exponential function to it to show that tau is larger.

We appreciate the reviewer’s comments and have significantly revised these sections of the Methods and Results. In summary, we fitted the data with single- and double-exponential functions. Double exponentials were fitted to the full time course. Single exponentials were fitted to both the full time course (Supplementary Figure 1) and a truncated version excluding the first 10 points (~14 s) to mitigate initial variability (Supplementary Figure 2). AIC comparisons heavily favoured the double-exponential model for the full time course (mean AIC: double −442.8 vs. single −376.4), providing a better fit for ≥20/22 subjects across all environments. Compared against the truncated single-exponential fit, the double-exponential model retained a lower mean AIC (−420.7 vs. −399.1) and remained the preferred model for approximately half the subjects. The double-exponential model was preferred not only for its automated nature (requiring no manual truncation) but also for the magnitude of improvement: when the single model was superior, the advantage was marginal (ΔAIC = 9.7 ± 1.4), whereas the double model’s advantage was substantial (ΔAIC = 52.8 ± 8.9). We therefore used the double-exponential fit for further analysis.

(10) Lines 436–446: How can this analysis be performed if asymptotic performance was not achieved in any of the conditions (nothing has plateaued in Figure 2B)? Also, the 10% FHR measure is dependent on the FHR estimate; correlating two measures based on the same measure is, by definition, expected to be correlated. This result seems to reflect that if one's learning is faster within a fixed number of trials (150), one has more opportunity to reach a higher final PCC even if asymptotic performance is identical.

To clarify, the variables being correlated are not the FHR and ±10% of the FHR values themselves, but rather the time points at which each participant reached ±10% of their individual FHR during the task. This analysis therefore does not involve two measures derived directly from the same estimate. The timing of reaching ±10% of the FHR reflects the learning-settling trajectory rather than the FHR magnitude, so there is no a priori reason for the two measures to be intrinsically correlated. While asymptotic performance was not reached within 150 trials for some participants, the estimated FHR still provides a consistent individual marker of learning rate, allowing comparison of relative learning dynamics across participants and conditions.

(11) Lines 448–461: Brandewie & Zahorik (2013) show that a large portion of that improvement is due to tuning to the voice and location of the speaker. Also, in the current study, there are some issues with the anechoic condition (see below).

We thank the reviewer for this comment. This section refers specifically to results related to carrier phrase length, not to speaker identity (addressed separately below) or location, both of which were fixed in our study and therefore unlikely to account for the observed effects. We address the reviewer’s concerns about the anechoic condition in our responses below.

(12) Line 491: What were the average trial numbers for the steady and initial trials? Also for the anechoic condition?

We appreciate the reviewer raising this. We analysed the average trial number at which initial and steady trials occurred across a total of 360 trials: for all 22 participants, initial trials mean = 6 ± 4 and steady trials mean = 37 ± 9; sham TMS: initial trials mean = 5 ± 4, steady trials mean = 34 ± 6; real TMS: initial trials mean = 10 ± 9, steady trials mean = 38 ± 12; anechoic condition: initial trials mean = 6 ± 3, steady trials mean = 36 ± 8; and the 11 randomly selected subjects: initial trials mean = 6 ± 5, steady trials mean = 38 ± 7. This information has been added to the relevant Results sections.

(13) Line 498: It's still not clear when the anechoic condition was performed. Lines 200–205 talk about combinations in which the Lecture Room was swapped, but it's impossible to follow when and how often that occurred. Given that the anechoic room was not included in the same way as the main three rooms, the conclusion at lines 500–504 is questionable.

Thank you for noticing this. We have rewritten the relevant section of the Methods (“Identity of sound environments”) as follows: “Additionally, we performed three follow-up experiments in different groups of listeners, assessing performance across combinations of three rooms: (1) Anechoic Room/Open-Plan Office/Underground Car Park (10 naïve listeners); (2) Living Room/Open-Plan Office/Underground Car Park (11 naïve listeners); and (3) Highly Reflectant Room/Open-Plan Office/Underground Car Park (10 naïve listeners). These conditions were used to determine whether a specific room combination was required to observe improvements in speech performance with increasing exposure to room acoustics—i.e., with increasing carrier phrase length—and were assessed in the same way as Brandewie & Zahorik (2013).” The anechoic room was therefore explored in the same manner as the main three rooms.

(14) Lines 508–521: Tuning to the talker's voice/location would not predict that the effect would be different for a different voice.

We thank the reviewer for this point. If the improvement in speech understanding were due to tuning to a specific talker’s voice or location, we would expect the effect to differ across talkers. However, our analysis across six talkers (three female, three male) showed no significant interaction between talker, carrier phrase length, and room (F(30,630) = 0.73, p = 0.85, ηp2 = 0.034). Although overall performance differed across talkers (main effect of talker: F(5,105) = 27.19, p < 0.001, ηp2 = 0.56), these differences did not modulate the carrier phrase effect. We therefore conclude that the improvement in speech understanding with increasing carrier phrase length is consistent across talkers.

(15) Lines 523–536: Neither of these tests addresses the question directly. That would require switching the talker randomly between the carrier and target (or throughout the sentence).

We thank the reviewer for this comment. We respectfully disagree that our analyses fail to address the question. While our experiment was not specifically designed to test the effect of switching talkers between the carrier and target segments, we examined whether adaptation to a talker could explain the improvement in performance with increasing carrier phrase length through three complementary analyses: (1) a repeated-measures ANOVA testing for interactions between talker and carrier phrase length (see response above); (2) an analysis of potential carry-over effects across consecutive same-talker trials; and (3) an assessment of talker-learning effects in the absence of reverberation (anechoic condition). As detailed in the Results section “Improvements in performance are explained by exposure to the environment, not talker idiosyncrasies,” none of these analyses revealed evidence that talker identity influenced the observed improvement in speech understanding. We therefore conclude that the performance improvements with increasing carrier phrase length are better explained by adaptation to the acoustic environment than to specific talkers.

(16) Line 544: Why is FHR used in this measure when d' is used for the standard analysis in Figure 2D?

As noted above, d′ is referred to only in statistical analyses that do not bear a specific relation to the time-course analysis. Given how d′ was calculated for | Colour | and |Number|, it was not considered a useful metric for describing performance across time in different environments, owing to the paucity of data affecting temporal resolution. We therefore used cumulative hit rates, with a 5-point moving average (~7 seconds), plotted against mean cumulative exposure time. This has been clarified in the Methods.

(17) Also, why is FHR, as opposed to HR (which I assume is really PCC), computed across the whole experiment?

FHR refers to the Final Cumulative Performance. This naming was used to distinguish it from trial-by-trial Hit Rate used in the time-course analysis. FHR is the final data point of the cumulative hit rate—i.e., after all responses have been accumulated in that listening environment. This has been clarified throughout the manuscript.

(18) Still worse, it's also not clear when these anechoic trials were measured.

We appreciate the reviewer noting a lack of clarity here. We performed three follow-up experiments in different, naïve groups of listeners, assessing performance across combinations of three rooms, including Anechoic Room/Open-Plan Office/Underground Car Park (10 naïve listeners), assessed in the same way as Brandewie & Zahorik (2013). The anechoic room was therefore explored in the same manner as the main three rooms; this has been clarified in the Methods and reiterated in the Results.

(19) And specifically, from Supplemental Table 4, it looks like the improvement was considerable (up to 15%), supporting that the effect is occurring. Also, note that there seems to be something numerically wrong in Supplemental Table 4: the improvement CP0–CP1 is −6, CP1–CP2 is −9.667, and CP2–CP3 is −2.5. Based on this, CP0–CP2 is expected to be −15.667 (which it is), but CP0–CP3 is expected to be −18.167, yet it's stated as −13.167.

We appreciate the reviewer pointing this out. Our statistical analysis does not match the calculations the reviewer derived from Supplemental Table 4. For transparency, we report below the means for each carrier phrase in the anechoic room, exported directly from SPSS, which are the values reported in the manuscript. We have reviewed this section to improve clarity and have included a link to the raw supplemental data.

CP0: Mean 40.500, SE 4.548, 95% CI [30.212, 50.788]

CP1: Mean 46.500, SE 5.296, 95% CI [34.519, 58.481]

CP2: Mean 56.167, SE 3.777, 95% CI [47.624, 64.710]

CP3: Mean 53.667, SE 3.966, 95% CI [44.695, 62.638]

(20) Lines 555–557: This sentence seems grammatically incorrect.

It has been corrected.

(21) Lines 559–560: The sentence “a brain region implicated in listening performance in noise (Houtgast & Steeneken, 1973; Knudsen, 1929; Lochner & Burger, 1961)” seems to imply that the cited studies support dlPFC being the brain region implicated in hearing in noise. None of these studies does that.

Thank you for noticing this—this was an error introduced by our reference manager and has been corrected.

(22) Lines 586–592: What was the “overall performance” measure—d′, HR, or PCC? Also, what is “univariate” analysis here? A mixed ANOVA with a between-group factor of condition and withingroup factors of room and CP length would be appropriate, and the whole group of 22 subjects should be used for the “no-exposure” group, rather than a random selection of an 11-subject subgroup.

We appreciate this comment and we have revised this analysis to include a mix ANOVA as suggested by the reviewer. It reads as follows in the Manuscript: “Given the potential placebo effects of a ‘sham’ TMS stimulation, we first tested whether our sample of 11 ‘sham’ TMS participants exhibited similar behavioural performance to the larger sample of 22 participants who had not been exposed to any TMS manipulation. A mixed ANOVA with a between-group factor of Condition and within-group factors of room and CP length confirmed that performance in these two populations was comparable, with no significant main effect of TMS conditions observed (‘sham’ vs. no exposure to TMS): [F (1,31) =0.01, p=0.91, ŋp2 = 0.00], confirming that participants experiencing ‘sham’ TMS did not perform significantly differently from the ‘no exposure to TMS’ population.”

(23) Lines 600–601: By “univariate ANOVA” is meant one-way ANOVA? And why wasn't it a two-way ANOVA with factors of room and sham/real TMS? More importantly, the 10% of FHR measure is arbitrary and should be replaced by standard fitting, as discussed earlier. Looking at Figure 3B, the black line appears near an asymptote while the red one is still growing toward the end, and that should be reflected in tau.

The ANOVAs in this and other sections have been revised based on the reviewers’ suggestions; mixed ANOVAs have instead been performed and reported, yielding similar results. The fittings and 10% FHR calculations have also been extensively revised. We now show that double exponentials are better suited to our dataset, and that two-tau parameters are not informative about when performance reaches a stable point during the task.

(24) Also, why is the exposure time on the x-axis different in Figure 3B from Figure 2B (150 vs 500)? And it would be good to see where the across-room average no-TMS data would lie here (or show the equivalent average in Figure 2B).

We appreciate the reviewer noticing this mismatch. Figure 2B shows the time course for each environment (150 s of exposure to each), whereas Figure 3B shows all environments collapsed (150 s × 3). This is because, for the 22 listeners without TMS exposure, a Rooms main effect was observed, justifying separate time courses per room; however, for listeners exposed to sham and real TMS, no Rooms × TMS interaction was observed, so separating time courses per room was not statistically justified. As the only significant effect was TMS condition, we grouped the time spent across all environments by TMS condition.

(25) Lines 606–617: This analysis and Figure 3C have the same issues as described for Figure 2C— asymptotic performance was not achieved for many conditions, so the 10% measure is arbitrary, as is the resulting correlation.

We appreciate the reviewer raising these fitting issues. This part of the manuscript has been extensively revised, including new analyses and figures, although our results have not changed. Additional detail has been added to the Methods (“Speech Performance Analysis and Timecourse Fittings of Mean Cumulative Hit Rates”), and the following summary has been added to the Results (“Statistical learning of reverberant environments occurs over long and short time courses”): we fitted data with single- and double-exponential functions; double exponentials were fitted to the full time course, and single exponentials to both the full time course (Supplementary Figure 1) and a truncated version excluding the first 10 points (~14 s) (Supplementary Figure 2). AIC comparisons heavily favoured the double-exponential model for the full time course (mean AIC: double −442.8 vs. single −376.4), providing a better fit for ≥20/22 subjects across all environments, and remained preferred when compared against the truncated single-exponential fit (−420.7 vs. −399.1, preferred for roughly half the subjects). The double-exponential model was preferred for both its automated nature and the magnitude of improvement (marginal ΔAIC = 9.7 ± 1.4 when the single model won, versus substantial ΔAIC = 52.8 ± 8.9 when the double model won). We therefore used the double-exponential fit for further analysis.

(26) Lines 619–620: Was d′ really calculated using FHR (the final value) and a non-final False Alarm Rate? This would be arbitrary. It is still unclear how HR and FAR are defined here. There is no apparent benefit to switching between HR (Figure 3B/C, presumably overall percent correct, PCC), d′ (D, E, F), and back to HR (H, I).

We appreciate the reviewer raising this. We have clarified in the Methods (“Speech performance analysis and time-course fittings of mean cumulative hit rates”) how Hit Rate, False Alarm Rate, and d′ were calculated.

(27) Line 655: Figure 3E should be Figure 3F.

Corrected.

(28) Lines 661–670: Why was Number only analyzed for initial trials, while Color was analyzed for both initial and steady trials? Also, the choice of trials 9–10 for “steady” is arbitrary and should be shown somewhere in Figure 3B.

We analysed performance for |Number| on initial trials (1–2) only, for CP0, because performance for this speech token could only improve if positively influenced by short-term, within-trial accumulation of information (acknowledging that | Colour | precedes |Number|). To determine how much knowledge accumulated over repeated exposures — i.e., metaadaptation—we instead needed to compare performance on a speech token whose improvement could only stem from knowledge gained across previous trials, not within a single carrier phrase. We compared | Colour | performance on CP0 between initial trials (1–2) and later, steady trials (9–10). If this hypothesis is supported, it suggests that | Colour | performance for CP0 benefits from meta-adaptive information conveyed across trials as knowledge of the environment’s global structure accumulates—our proxy for meta-adaptation (Figure 3G). The choice of trials 9–10 as “steady” follows work on animal models of meta-adaptation (Robinson, Harper, & McAlpine, 2016), which described a faster adaptation rate after the eighth presentation of an environment. This has been clarified in the manuscript.

(29) Lines 724–728: This description is confusing. The main effect of “Lecture Room” vs. “Highly Reflectant” context is that performance is very good in the Lecture Room (green line) and poor in the Highly Reflectant Room (purple). Averaging that with OPO and CP and reporting “mean difference = 20.09” (in what units?) as “overall performance” distracts from the main point. Moreover, how can that be entered into an ANOVA when the room contexts differ (LR+OPO+CP vs. HR+OPO+CP)? That ANOVA seems incorrect; it should only be performed on OPO+CP across the two contexts.

This section has been revised and re-analysed as suggested. Redundant and unnecessary statistical comparisons were removed, retaining only those that show the effect of context on OPO and CP when comparing the different contexts in which these common environments were learned.

(30) Lines 740–750: Again, it is not surprising that when Living Room replaces Lecture Room—and performance in Living Room is worse than in Lecture Room—the average of LiR+OPO+CP is lower than LER+OPO+CP, if OPO+CP performance is unchanged. The interesting question is whether anything changed in OPO+CP performance, as suggested for the previous point.

This section has been revised as suggested by the reviewer.

(31) Lines 752–771: Again, the same issue—the main effect is that performance in Anechoic trials is worse than in Lecture Room or Living Room trials, which alone explains the group difference. I am also sceptical of the finding that Anechoic performance is worse than reverberant performance, contrary to typical spatial-release-from-masking results, where reverberation degrades performance by adding noise energy at the better ear and reducing binaural benefit through decorrelation. This may be an artefact of how target and noise levels were normalized after convolution with HRTFs/BRIRs (or the use of Ambisonics); no acoustic analysis of the stimuli is provided. At minimum, the total received level at the two ears for target and masker in every environment should be reported. Brandewie & Zahorik (2013), using equivalent anechoic and reverberant conditions, never observed reverberant performance to exceed anechoic, contrary to what is stated here (lines 754–755).

We appreciate the reviewer raising these points. The statistical analysis in this section has been revised: only the common rooms across the three-room conditions (Open-Plan Office and Car Park) were directly compared. Performance in Living Room and Lecture Room was not statistically different (Results, paragraph 4, “Statistical learning of room acoustics is tuned to universally experienced reverberation times”). However, Anechoic and Lecture Room performance remained significantly different (mean difference = 11.08, t(9) = 2.66, p = 0.013, Cohen’s d = 0.84), as did performance in the common rooms when learned in the context of Lecture Room versus other contexts (mean difference = 9.7, F(1,41) = 13.24, p < 0.001, ηp2 = 0.26).

While this setup resembles many masking studies, Brandewie & Zahorik (2013) tested four rooms simultaneously, whereas we tested three-room conditions explicitly designed to test environment-mix adaptation. We observed a synergistic relationship between performance in ‘good’ reverberant rooms (Lecture Room, Living Room) and the common but less favourable rooms (Open-Plan Office, Car Park, with longer RT60): in the absence of a ‘good-reverb anchor,’ performance in the common rooms improved less over time, possibly because participants had less to leverage in anechoic environments.

We agree that verifying at-ear acoustic levels is critical to ruling out a normalization artefact. Our stimuli were normalized in the 41-channel sound field, not at the listener’s ears: source speech and noise were convolved with the 41-channel anechoic or reverberant impulse responses, the 41-channel noise energy was scaled to a target of 70 dB, and the 41-channel speech field was scaled to the target SNR; the 41-channel signals were then rendered to two channels using a Higher Order Ambisonics-to-binaural decoder (hoa2bin), preserving natural head-related acoustic effects such as head shadow.

To verify that this did not create an at-ear artefact, we extracted simulated at-ear RMS energy for speech and noise after hoa2bin rendering and mapped these to approximate dB SPL using the 70 dB sound-field anchor (see supplied table, Summary Reverb Data). In the anechoic condition, the noise (positioned to the left) is strongly attenuated at the right ear by head shadow (dropping from ~57 dB to ~51 dB), giving the frontal target speech a highly favourable SNR at the better ear. In the reverberant condition, room reflections fill in the head shadow, raising noise level at the right ear to 55–56 dB depending on room, substantially lowering the ear SNR relative to the anechoic condition. The improved behavioural performance in reverberation therefore occurred despite a poorer acoustic SNR at the better ear, confirming this is not a normalization artefact but rather a genuine perceptual spatial release from masking, likely driven by early reflections aiding target integration and late reverberation decorrelating the noise binaurally. We have added the at-ear acoustic details to the Methods (“Stimulus Normalization and Binaural Rendering”) and Table 2, and updated the Discussion to clarify this mechanism.

(32) Lines 755–759: Describing anechoic spaces as “rare” and as rooms whose “walls are treated” states the facts backwards. Open spaces (e.g., a grass lawn) are largely anechoic fields, and people spend considerable time in such environments. An anechoic room may be artificial, but an anechoic (or near-anechoic) space is very common, and the room is simply an attempt to simulate that within an enclosure.

We appreciate this point and have rewritten this section to reflect it.

(33) Lines 808–812: This sentence appears incorrect. It refers to “the ability to correctly report keywords spoken in environments with the more extreme—lower or higher—RIRs,” presumably meaning OPO and CP, but these are not the environments with extreme RIRs; or, if referring to An and HR, those were not “encountered in experimental blocks also containing the moderately reverberant Lecture Room or Living Room.”

Thank you for noting this. We have rephrased this section to refer only to the extreme high-RIR environments encountered.

(34) Line 843: Should Fig 3Ai be Fig 3A? Also, in that figure there are arrows between dlPFC and A1, and between A1 and (the cerebellum?)—it's unclear what these represent.

The arrows were intended to represent feedforward and feedback information flow to lower auditory brain centres. We acknowledge they were confusing and have removed them from the figure.

(35) Lines 929–949: The authors did not account for listeners tuning to voice and location (as now cited via Best et al.), and their own and Brandewie & Zahorik's data show improvement due to carrier phrase even in the anechoic case (with the inconsistency in Supplemental Table 4 noted earlier). A direct test—switching the environment between carrier phrase and target phrase, as in Brandewie & Zahorik and Vlahou et al.—would be needed to fully attribute the effect to reverberation rather than other factors.

We thank the reviewer for this detailed comment and agree that directly manipulating the environment between carrier and target phrase would provide the most direct test of environment-specific adaptation. While our study did not implement this manipulation, our data provide converging evidence: (1) listeners showed improvement with longer carrier phrases even in the anechoic condition, consistent with previous reports, but this improvement did not interact with talker identity, carrier phrase length, or room, indicating it is not driven by tuning to specific voices or locations; and (2) regarding Supplemental Table 4, the calculations suggested by the reviewer do not match our statistical analysis—we have reported the SPSS-exported means directly (shown above) and reviewed this section for clarity, including a link to the raw supplemental data. Taken together, while we cannot fully quantify the proportion of adaptation attributable to reverberation versus other factors without the direct environment-switch manipulation, our results indicate that the observed improvements primarily relate to exposure to the environment rather than talker-specific effects.

(36) Lines 951–953: It is unclear what about “understanding speech in background noise” distinguishes this study from previous studies of adaptation to reverberation, many of which also examined speech in noise (as reviewed in Tsironis et al., 2024). Rather than reviewing pertinent studies on adaptation to reverberation for speech tokens, the authors cite abstract noise-texture studies that are only partially relevant, given the prevalence of speech in everyday listening (lines 955–958).

We appreciate the reviewer raising this point. Our intention was to refer specifically to statistical learning of implicit environmental acoustic features such as reverberation, rather than to speech-in-noise perception per se. We have revised the text accordingly: “A key feature of our study, which distinguishes it from previous investigations of statistical learning of acoustic features in human listeners, is the use of an ethologically valid listening task—understanding speech in background noise while listeners implicitly learn repeated acoustic features.”

(37) Lines 978–980: In what way? For environments with large T60, a simpler explanation than “ecological validity” is that there is more late reverberant energy in the target acting as a masker, predictable from DRR.

This section has been rewritten to clarify that it is the decline in performance at longer RT60 that is reminiscent of the decline observed under rTMS.

(38) Line 981: What does “the better to understand speech in reverberant background noise” mean?

This sentence has been revised.

(39) Lines 987–988: When did “performance decline over the course of an experimental session”? Figures 2B, 3B, and 4A all show performance improving over the session.

We have rephrased this sentence to refer to a decline in overall performance.

(40) Lines 990–992: Many previous studies report better adaptation to reverberation for some rooms than others (e.g., Brandewie & Zahorik, 2010; Vlahou et al., 2021), but none have reported decreased performance for an anechoic space relative to a reverberant one. This anomaly should be explained and reconciled with the existing literature before invoking ecological explanations such as “ethologically relevant environments.”

We thank the reviewer for raising this important point. We agree that verifying the at-ear acoustic levels is critical to ruling out a normalization artefact, particularly given our finding that reverberant performance exceeded anechoic performance. To address this directly: our stimuli were normalized in the 41-channel sound field, not at the listener’s ears. The source speech and noise were convolved with the 41-channel anechoic or reverberant impulse responses; the 41-channel noise field was scaled to a target of 70 dB and the 41-channel speech field scaled to the target SNR; the 41-channel signals were then rendered to two channels via a Higher Order Ambisonics-to-binaural decoder (hoa2bin), preserving natural head-related effects such as head shadow because normalization preceded binaural rendering.

To verify that this sound-field normalization did not create an at-ear artefact, we extracted simulated at-ear RMS energy for speech and noise after hoa2bin rendering and mapped these to approximate dB SPL using the 70 dB sound-field anchor (Summary Reverb Data table). In the anechoic condition, the noise (positioned left) is strongly attenuated at the right ear by head shadow (dropping from ~57 dB to ~51 dB), giving the frontal target speech a highly favourable SNR at the better ear. In the reverberant condition, room reflections fill in the head shadow, increasing right-ear noise level to 55–56 dB depending on room, substantially worsening the atear SNR relative to the anechoic condition. The improved behavioural performance in reverberation therefore occurred despite a poorer acoustic SNR at the better ear, confirming the finding is not a normalization artefact but instead reflects a genuine perceptual spatial release from masking—likely driven by early reflections aiding target integration and late reverberation decorrelating the noise binaurally. We have added these at-ear acoustic details to the Methods (“Stimulus Normalization and Binaural Rendering”) and Table 2 to clarify this mechanism.

(41) Discussion: Given the questions about the results, the discussion might need to be rewritten to only discuss claims that are actually supported.

The Discussion section has indeed been extensively revised.

(42) The hippocampus and other areas have been proposed for statistical learning, and studies also show that disruption of DLPFC can boost statistical learning (https://doi.org/10.1016/j.jml.2020.104144).

We appreciate the reviewer raising this. We have cited Ambrus et al. (doi:10.1016/j.jml.2020.104144) in the Discussion (line 888) as evidence of opposing effects of dlPFC stimulation on statistical learning, and have further revised lines 891–903 of the Discussion to more clearly describe the known projections and functional interactions between dlPFC, hippocampus, striatum, and basal ganglia that support implicit and statistical learning.

(43) Line 1739: What is “cumulative” here?

“Cumulative” has been deleted.

Version 2

Abstract

Navigating complex sensory environments is critical to survival, and brain mechanisms have evolved to cope with the wide range of surroundings we encounter. To determine how listeners learn the statistical properties of acoustic spaces, we assessed their ability to perceive speech in a range of noisy and reverberant rooms. Listeners were also exposed to repetitive transcranial stimulation (rTMS) to disrupt the dorsolateral prefrontal cortex (dlPFC) activity, a region believed to play a role in statistical learning. Our data suggest listeners rapidly adapt to statistical characteristics of an environment to improve speech understanding. This ability is impaired when rTMS is applied bilaterally to the dlPFC. The data demonstrate that speech understanding in noise is best when exposed to a room with reverberant characteristics common to human-built environments, with performance declining for higher and lower reverberation times, including fully anechoic (non-reverberant) environments. Our findings provide evidence for a reverberation “sweet spot” and the presence of brain mechanisms that might have evolved to cope with the acoustic characteristics of listening environments encountered every day.

Introduction

Learning occurs over multiple time scales—evolutionary, developmental and moment-to-moment—to support a diverse range of abilities: navigating complex sensory environments (Bregman, 1994; Lewicki et al., 2014; Smith & Lewicki, 2006), expressing intricate behaviours in social settings (Gariépy et al., 2014; van den Bos et al., 2013), or acquiring communication skills such as bird song (Brainard & Doupe, 2002; Lauay et al., 2004; Marler, 1970) or spoken language (Aslin et al., 1998; Saffran et al., 1996; Saffran, 2003). Although some learning requires active or explicit involvement in a task (Huyck & Wright, 2011; Mathews et al., 1989; Rebuschat, 2015), perhaps with a system of rewards and punishments to make the learning ‘stick’ (Barberis, 2013; Schultz, 2002; Wächter et al., 2009), other forms of learning seem automatic or implicit (Reber, 1967), acquired with individuals seemingly unaware it is taking place. This type of learning, referred to as ‘statistical learning’ (Ambrus et al., 2020; Saffran et al., 1996), ‘sequence learning’ (Nissen & Bullemer, 1987; Vékony et al., 2022) or ‘sequential learning’ (Conway & Christiansen, 2001; Vékony et al., 2022), is thought to entail automatic and incidental extraction of regularities or patterns within external stimuli or in the environment (Conway, 2020; Takács et al., 2021). Evident across sensory modalities to support automatic learning of tonal (Saffran et al., 1999) or linguistic (Saffran et al., 1996) sequences, strings of letters (Reber, 1967), visual scenes and shapes (Fiser & Aslin, 2001), visual-motor patterns (Nissen & Bullemer, 1987), and even tactile input (Conway & Christiansen, 2005), statistical learning appears to be a unitary, domain-general phenomenon, potentially governed by a single mechanism or neurocognitive principle (Conway, 2020; Kirkham et al., 2002).

Increasing evidence suggests that statistical learning is a critical part of how listeners deal with complex and cluttered acoustic scenes. Potential background sounds such as rain or insects—referred to as sound textures (Hicks & McDermott, 2024; McDermott et al., 2013; McWalter & McDermott, 2018) as well as changes in the regularity of sound patterns (Barascud et al., 2016; Bianco et al., 2020) are processed—seemingly unconsciously—in terms of their summary statistics. This form of statistical learning likely contributes to our ability to follow conversations in background noise (‘cocktail party listening’; Cherry, 1953) and to deal with reverberant spaces where multiple, delayed copies of the same sound reach a listener in the form of reflections from walls and other acoustically opaque surfaces (Blesser & Salter, 2009; Sabine, 1953; Schroeder, 1962). Though listeners rely on early-arriving sound energy to determine source location—suppressing potentially conflicting localization cues in later-arriving, often more intense, sound energy (Bradley et al., 1999; Culling et al., 2003; Houtgast & Steeneken, 1985; Nielsen & Dau, 2010)—the perception of reflected sound energy is informative of the listening environment more broadly (Bronkhorst & Houtgast, 1999; Shinn-Cunningham, 2000). This includes whether environments are real or synthetic i.e., deviated from natural reverberation characteristics (Traer & McDermott, 2016), room dimensions (Cabrera et al., 2005; Kolarik et al., 2021; Zahorik & Wightman, 2001) and source distance (Bronkhorst & Houtgast, 1999; Zahorik & Wightman, 2001).

Nevertheless, despite its potential utility for understanding background features of a sound environment, the accumulation over time of late-arriving, reverberant sound energy is thought to generate an additional burden on listening performance beyond that from sound energy direct from interfering sources (Houtgast & Steeneken, 1973; Knudsen, 1929; Lochner & Burger, 1961; Santon, 1976; Shinn-Cunningham & Kawakyu, 2003), smearing the acoustic waveform and occluding temporal gaps that might otherwise be helpful for ‘glimpsing’ speech in background noise (Cooke, 2006). However, if listeners are able to utilize late-arriving, reverberant energy of known acoustic environments to suppress disruptive spatial cues and enhance speech understanding (Brandewie & Zahorik, 2010, 2013; Vlahou et al., 2019; Watkins, 2005a, 2005b), then the acoustic characteristics of an environment can be learned and stored for later use in complex listening tasks.

Here, using an ecologically relevant listening task—understanding speech in background noise—we assessed the ability of human listeners to learn the statistical structure of different sound environments defined by their RT60, the time it takes for reverberant energy to decay by 60 decibels (dB), and confirmed that speech understanding in background noise improved over time. Specifically, when asked to report words from unfamiliar and semantically uninformative spoken sentences of varying duration in noise convolved with the reverberant qualities of different acoustic environments, listeners’ performance improved with increasing sentence duration, and with repeated exposure to each environment (Brandewie & Zahorik, 2010, 2013).

Building upon this literature, we wondered, if the learning of acoustic environments i.e., improvements in performance with increasing sentence duration, followed a similar time-course to in vivo experimental evidence of increased adaptive capacity for neural learning of sound environments upon repeated exposure to those environments (Dean et al., 2005, 2008), and whether this adaptive capacity was modulated by cortical circuits (Robinson et al., 2016) such as the dorsolateral prefrontal cortex (dlPFC), a cortical locus hypothesised to contribute to statistical learning in human listeners (Ambrus et al., 2020; Vékony et al., 2022). Specifically, we asked whether knowledge of the acoustic environment accumulated over time—that is, whether performance during initial exposures was poorer than during later ones—and whether this accumulation was reduced when bilateral dlPFC activity was disrupted by TMS. We found that the capacity to learn the acoustic environment to aid speech understanding was diminished when repetitive transcranial magnetic stimulation (rTMS) was applied bilaterally to impair the function of dlPFC, suggesting that indeed human listeners have neural mechanisms supporting listening in noise and reverberant conditions.

Consistent with this, Francl and McDermott (2022) demonstrated that an auditory model trained under realistic listening conditions—incorporating both noise and reverberation—was able to reproduce several human-like spatial hearing characteristics. To this end, we wondered if the learning effects we observed relied on the combination of real-world environments we employed, i.e., the RT60s in which listening performance was assessed. Specifically, we hypothesized that the ability to understand speech in noise depends on exposure to specific values of RT60 across our experimental paradigm, including those in the range commonly experienced by human listeners in natural and built environments (Traer & McDermott, 2016). Counter to expectations that reverberant energy can only harm speech understanding, listeners’ abilities to leverage the knowledge of a sound environment (talker identity, speech corpus, noise characteristics, spatial configuration) to improve speech understanding was best when a ‘typical’ room with reverberant characteristics close to the average of a wide range of common (built) environments (Traer & McDermott, 2016) was included as one of the three environments presented in a single experimental run. Performance declined systematically when this room was switched for one with more, or less, reverberation, including a fully anechoic (i.e., non-reverberant or dry) environment in which listeners initially trained on the recall task. Importantly in the context of other potentially learnable acoustic features, talker identity (three male and three female) represented a random variable in our experimental design. Evidence for a reverberation ‘sweet spot’ suggests the existence of brain mechanisms adapted to the longer-term structure of common listening environments (Traer & McDermott, 2016).

Materials and Methods

Participants

A total of 74 participants (53 were females), aged between 19-26 years old (mean ± SD = 22 ± 2 years old) were recruited across all experiments. They were Australian native-English speakers, had normal pure tone thresholds (< 20 dB HL tested at octave intervals between 0.5-8 kHz (Hughson & Westlake, 1944); Interacoustics Hearing Aid Fitting Analyzer Affinity 2.0 Audiometry) and normal middle ear function (assessed using standard 226 Hz tympanometry; Titan, Interacoustics). To ensure normal outer hair cell function, all participants were screened for Distortion Product Otoacoustic Emissions (DPOAEs) between 0.5-10 kHz (stimulus parameters: f1/f2 =1.2, f1 =65 dB SPL; f2 =55 dB SPL; responses parameters: SNR > 6 dB, signal level > −10 dB SPL, and reliability > 98% (DPOAE440; Titan, Interacoustics). All participants had normal steady-state ipsilateral broadband noise middle ear muscle reflexes (MEMR) (Titan, Interacoustics). The audiological testing took 15-20 minutes to complete.

Acoustic stimuli

To assess listeners’ abilities to understand speech in background noise across different sound environments, they were asked to verbally report keywords spoken by a talker virtually located in front of the participant that were masked by white noise arriving from a virtual source 90° to the left (Figure 1A) (Brandewie & Zahorik, 2010, 2013). A binaural configuration was necessary as speech intelligibility improvements following exposure to room acoustics is significantly reduced under monaural listening conditions (Brandewie & Zahorik, 2010). The speech stimuli used in this study were from the Coordinate Response Measure (CRM) corpus (Bolia et al., 2000), and all combinations of Callsigns (‘Baron’, ‘Eagle’, ‘Charlie’, ‘Tiger’, ‘Arrow’, ‘Ringo’, ‘Laker’, ‘Hoper’), Colors (four monosyllabic choices, ‘red’, ‘white’, ‘blue’, or ‘green’), and Numbers (the English digits between ‘one’ and ‘eight’) were used in this experiment (Figure 1B) (Brandewie & Zahorik, 2010, 2013). The masking white noise was randomly generated on a PC using MATLAB (The MathWorks, Inc. of Natick, MA: https://au.mathworks.com/) and was presented with the speech. The masking noise preceded the speech stimuli by 150 ms. The masker’s amplitude was set to increase linearly from zero to the target level (70 dB SPL), and ended at the same time as the speech stimuli, without ramping (Brandewie & Zahorik, 2013).

Experimental environment.

A. Speech-in-noise immersed in real reverberant environments reproduced in an anechoic chamber over a 41-loudspeaker array. All speech material was rendered as if it was originating from directly in front of a listener and was masked by spatially separated (90° to the left) white noise at −15 dB signal-to-noise ratio (SNR). The noise masker level was fixed at 70 dB SPL. The speech material was convolved with the measured impulse response of a variety of different real rooms, from anechoic to highly reverberant real rooms such as an underground car park (see Table 1 methods for acoustic details). B. Structure of the speech material. Sentences from the Coordinated Response Measures (CRM) corpus were modified to vary their duration depending on how many words preceded the target phrase i.e., ‘carrier phrase’ (CP) length. Participants were asked to recognize and repeat target words: |Color| from a list of four and |Number| from a 1-8 list. C. Example of the experimental paradigm. Trials consisted of ‘carrier phrases’ spoken in a specific room i.e., ‘go to blue 1 now’ in the Lecture Room. Participants were never exposed to the same room consecutively and the task lasted no longer than 45 minutes. D. Sensitivity to |Color| and |Number| combined in the Lecture Room, Open-Plan Office and Underground Car Park. Mean d’ (measure of accuracy calculated as: Z (correct responses) – Z (false alarm)] denoted as circles in the boxplot (n=22 for all rooms). The horizontal line denotes the median. Upper and lower limits of the boxplot represent 1st (q1) and 3rd (q3) quartiles respectively, while whiskers denote the interquartile range (IQR =q3-q1). As previously reported (Brandewie & Zahorik, 2010, 2013), we observed an increase in performance for longer ‘carrier phrases’ in all environments (data available: https://doi.org/10.25949/24295342). Only room images in this figure were generated using artificial intelligence, with the exception of the Anechoic Room photograph, which was taken by the authors in the chamber where the experiments took place.

Virtual Environments

Reverberant environments were reproduced by convolving the generated acoustic signals with Room Impulse Responses (RIRs) obtained using 62-channel microphone array recordings of real rooms that were subsequently decoded into 41 higher-order Ambisonic (HOA) channels (Badajoz-Davila et al., 2020; Weisser et al., 2019). These final decoded channels corresponded to the spherical array of 41 Tannoy V8 concentric loudspeakers (Tannoy) installed in the anechoic chamber at the Australian Hearing Hub where testing took place. To optimize the “directionality” of the acoustic stimuli, the speech and masking noise signals were separately convolved with RIRs that had enhanced direct sound components in loudspeakers associated with their respective virtual source locations (i.e., 0° azimuth loudspeaker for the target talker and 90° azimuth loudspeaker for the masking noise) (Badajoz-Davila et al., 2020; Weisser et al., 2019). In the anechoic condition, RIRs only incorporating these enhanced direct sound components were convolved with the speech and masking noise signals. The SNR for speech-in-noise during data collection was −15 dB to avoid ceiling performance and was manipulated after convolution with the RIR by adjusting the gain of the speech target relative to a fixed masker level of 70 dB SPL. All signal processing was performed on a PC using MATLAB (Mathworks) and the final spatialized signal were presented via the RME MADI sound card (RME Audio) to two RME 32-channel digital-to-analog converters (M-32, RME Audio). These, in turn, fed 11 Yamaha XM4180 power amplifiers (Yamaha) that drove the loudspeaker array.

Identity of sound environments

Three rooms were selected, with similar Reverberation Times (RT60s) to those employed by Brandewie and Zahorik (2013) and assessed in 22 participants. As Bajadoz-Davila et al. (2020) and Brandewie and Zahorik (2013) calculated these RT60s according to ISO-3382 guidelines (ISO, 2009), they are referred to in the text as RT60iso: Lecture room (RT60iso = 0.46 s), Open-Plan Office (RT60iso = 0.96 s) and Underground Car Park (RT60iso = 2.42 s) (Table 1). Additionally, we performed 3 followed up experiments in different groups of listeners and assessed performance across combinations of three rooms: 1) Anechoic Room/Open-Plan Office/Underground Car Park (10 naïve listeners); 2) Living Room/Open-Plan Office/Underground Car Park (11 naïve listeners) and 3) Highly Reflectant Room/Open-Plan Office/Underground Car Park (10 naïve listeners). These conditions were employed to determine whether a specific combination was required to observe improvements in speech performance as exposure to room acoustics increased, that is, with increasing ‘carrier phrase length’, assessed in the same way as (Brandewie & Zahorik, 2013). For each combination, Lecture Room was swapped with one of three virtual environments: Anechoic (RT60iso = n/a), Living Room (RT60iso = 0.33 s) and Highly Reflectant Room (RT60iso of 1.55 s) (Table 1). Naïve participants were recruited for each room combination.

To compare our recordings to RIR data recorded and analyzed by Traer and McDermott (2016), we also calculated RT60s for each room by integrating across the 41 HOA channels of each RIR and then calculating the median RT60 value in 31 frequency sub-bands with centre frequencies ranging from 80 Hz to 10 kHz (Traer & McDermott, 2016). We refer to these measures of reverberation as RT60traer. The range of sub-bands and their frequency selectivity was chosen to match that of the human ear (Glasberg & Moore, 1990; McDermott & Simoncelli, 2011). See SI Materials and Methods, (Traer & McDermott, 2016), for further information concerning how individual RT60traers were extracted from each sub-band.

Stimulus Normalization and Binaural Rendering

Stimuli were generated and normalized within a 41-channel sound field prior to binaural rendering to preserve natural head-related acoustic effects (e.g., head shadow). Source speech and noise signals were first convolved with 41-channel anechoic or reverberant impulse responses. The overall energy of the 41-channel noise field was scaled to a target of 70 dB, and the 41-channel speech field was subsequently scaled to achieve the target SNR. Following sound field normalization, the 41-channel signals were rendered to two channels using a Higher Order Ambisonics to Binaural (hoa2bin) decoder. To verify the at-ear acoustic levels, simulated at-ear root-mean-square (RMS) energy was extracted for the speech and noise tokens post-rendering. These digital RMS values were mapped to approximate dB SPL, based on the 70 dB noise sound field anchor, to confirm the resulting at-ear acoustic SNRs across both anechoic and reverberant conditions (Table 2).

Continuous theta-burst stimulation

Applying repetitive Transcranial Magnetic Stimulation (rTMS)—specifically continuous theta burst stimulation (cTBS)—to the right and left dlPFC, elicits a period of depressed cortical excitability in the targeted area that lasts about 60 minutes post-stimulation (Gamboa et al., 2010; Hoogendam et al., 2010; Huang et al., 2005; Romero et al., 2022), the time during which speech recall in reverberant rooms was assessed. Participants were screened for contra-indications to rTMS (see Supplemental Information. Transcranial Magnetic Stimulation screening form). Two cTBS conditions (‘real’ and ‘sham’ TMS) were counterbalanced across normal-hearing participants, and participants were blinded to the type of manipulation they might receive.

cTBS intensity applied to dlPFC was adjusted for every participant (i.e., both ‘real’ and ‘sham’ TMS groups) as a function of their individual corticospinal excitability assessed with single-pulse TMS-motor thresholds. Single-pulse TMS was delivered using a 70 mm figure-of-eight coil (Magstim Rapid2 system, Magstim, Whitland, United Kingdom), orientated at 45° to the scalp with current flowing posterior-anterior across the primary motor cortex. For the determination of single pulse TMS-induced motor thresholds we used visual observation of the first dorsal interosseous (FDI) muscle twitch (Varnava et al., 2011). Coil position and angle was adjusted until the optimal site for consistent and larger elicitation of FDI muscle twitch. Once the optimal site was located, the minimum level of stimulator output that produces a visible twitch in either the left or right hand in each individual, following single pulse stimulation of the respective contralateral motor cortex, was determined. Subjects were instructed to relax their arm, and rest it on their lap, palm upwards, whilst gently squeezing their thumb and forefinger together – as if squeezing a pea, whilst simultaneously keeping their other fingers relaxed. We then monitored for 3/5 visible twitches in the FDI in response to single pulse TMS. Subjects were occasionally asked to shake their arm to relax everything and then to resume holding their hand as before (Coltheart et al., 2018; Dienes & Hutton, 2013). Participants’ motor thresholds ranged from 46-57% of the maximum stimulator output (mean = 51 ± 5). Participants’ individual motor threshold for left and right hemisphere were recorded to set the intensity for bilateral cTBS stimulation which was administered by placing the same figure-of-eight coil over the right and left dlPFC located over electrode positions F3 and F4 (10-20 system EEG) (Herwig et al., 2003; Jurcak et al., 2005). Bilateral TMS was performed serially i.e., first the right or the left dlFPC (order was alternated among subjects), was chosen to limit possible compensation effects of the non-stimulated hemisphere (Ambrus et al., 2020). Each cTBS burst consisted of three pulses at 50 Hz, with bursts repeated at a frequency of 5 Hz, applied continuously for 40s, and delivered at an intensity of 80% of the right or left resting motor threshold (Huang et al., 2005).

Both real (applied to 10 naïve listeners) and sham (applied to 11 naïve listeners) TMS stimulations were identical and delivered to two groups of listeners, with participants counterbalanced between the two conditions. The sole difference between the ‘real’ and ‘sham’ TMS was the figure-of-8 coil utilized. Real TMS was administered using an active figure-of-8 coil, while the sham stimulation employed a sham coil (Magstim Rapid2 system, Magstim, Whitland, United Kingdom). The sham coil is non-stimulating, meaning it does not induce a significant magnetic field, and is designed to be indistinguishable from the active coil in appearance, feel, and sound. The duration of the TMS procedure was 20–30 minutes, primarily due to the estimation of motor thresholds, with cTBS stimulation delivered in the final 2–5 minutes.

Procedure

Familiarization

All participants first completed (within 5 minutes) a familiarization task that consisted of 10 trials (i.e., different ‘carrier phrase lengths’) presented in anechoic conditions at 0 dB SNR where listeners were asked to report |Color| and |Number|. Participants exposed to ‘real’ or ‘sham’ TMS completed the familiarization and behavioural task right after cTBS procedures. All participants performed the familiarization phase with 80-100% accuracy, confirming that all participants including exposed to cTBS understood the task and procedural learning was achieved and not affected by either ‘sham’ or ‘real’ TMS exposure.

Behavioural task

The CRM sentences were spoken by six talkers (3 males and 3 females) and were of varying duration where all, some, or none of the preceding sentence (the ‘carrier phrase’ (CP, Figure 1B)) before ‘|Color| |Number|’ was included (Brandewie & Zahorik, 2010, 2013). Thus, for CP0, listeners heard ‘|Color| |Number| now’, CP1—‘go to |Color| |Number| now’, CP2—‘|Callsign| go to |Color| |Number| now’, CP3—‘Ready |Callsign| go to |Color| |Number| now’ (Figure 1B and C). After each phrase was presented, participants reported the |Color| and |Number| they heard to the experimenter and performance was assessed based on keywords correctly identified. The rationale for only reporting |Color| and |Number| was to ensure the same amount of keywords identifications across different phrases lengths. For the reverberant conditions SNR was set to −15 dB, which we found in pilot testing resulted in a level of performance similar to (Brandewie & Zahorik, 2010, 2011, 2013; Zahorik, 2009). The room environment was selected randomly for each trial except that the same room could not appear in two consecutive trials. The target |Color|, |Number| and talker were selected randomly for each trial. For each experimental session, three different listening environments were assessed, and listeners were presented 360 trials (30 repeats x ‘room = 3’ x ‘carrier phrase length =4’). Each listener participated in one session only. Throughout the session, listeners were required to verbally repeat the appropriate |Color| and |Number| combination they heard from corpus lists physically displayed on both sides of the 0° azimuth loudspeaker. The CRM is a closed-set corpus, therefore participants were provided with |Color| and |Number| choices to select from (Brandewie & Zahorik, 2010, 2011, 2013; Zahorik, 2009). The experimenter continuously monitored participants’ responses while scoring performance, but no feedback was provided. The behavioural total test time was 45 min.

Speech performance analysis and time course-fittings of mean cumulative hit rates

d’ was first calculated individually for both |Color| and |Number| in different room/carrier combinations using Equation 1, where z(H) and z(F) were the z transforms of the hit rate and false alarm, respectively. Hit rates consisted of the correct |Color| or |Number| being selected for each trial, whereas false alarms were considered when a |Color| or |Number| was reported incorrectly to the presentation of any other |Color| or |Number|. To prevent infinite values of d’, hit rates/false alarms of 1 were set to 0.99, and corresponding false alarms/hit rates of 0 were set to 0.01 leading to maximal/minimal d’ values of ± 4.65. The total d’ for a carrier phrase length in a particular room was calculated by averaging |Color| and |Number| d’ values for that combination of room and ‘carrier phrase length’ across individuals. (1) d′=z(H)−z(F) Because d’ is calculated independently for |Color| and |Number|, it lacked the temporal resolution necessary to describe performance over time due to data paucity. Consequently, we analysed the development of individual and average performance in each acoustic environment using cumulative hit rates. We applied a 5-point moving average (∼7 seconds) to each trace and plotted performance as a function of mean cumulative exposure time.

To quantify performance trajectories, we fitted the time-series data with both single (Eq. 2) and double (Eq. 3) exponential functions: (2) f(t)=C+ mag single⋅exp⁡(−tτsingle) (3) f(t)=C+magfast⋅exp⁡(−tτfast)+magslow⋅exp⁡(−tτslow) where, f(t) represents the fitted performance at time t, and C denotes the constant y-intercept. The terms mag and τ represent the magnitude and time constant (in seconds), respectively, for the single, fast, and slow components.

Constrained optimization was performed using MATLAB’s fmincon function to minimize the mean squared error. To avoid local minima, fits were initialized using a gridded search. For single exponential models, we tested a 25 × 25 grid with τsingle ranging from 0–100 s and magsingle from −50 to 50. For double exponential models, we utilized a grid of 5 linearly spaced values for τslow (50–500 s), τfast (0–100 s), and their corresponding magnitudes (−10 to 10), resulting in 625 initialization points per model type. Upper bounds were set at 2000 s for τsingle and τslow, and 500 s for τfast. The optimal fit for a listening environment was selected from the resulting collection of single/double exponential curves by identifying fits with the highest adjusted R2. During optimization, fits were weighted by their variance; therefore, greater importance was attributed to performance at later exposure times. Furthermore, we forced all final fitted curves to pass through the final hit rate i.e. last data point, by penalizing those that did not during fitting.

Double exponential models were fitted exclusively to the full dataset, as they are capable of accounting for early data reversals. However, to account for inconsistent initial results, single exponential models were calculated for both the full dataset and a truncated dataset (excluding the first 10 points, (∼14 seconds). To determine the optimal model, we calculated the Akaike Information Criterion (AIC). AIC estimates prediction error by balancing goodness-of-fit against model complexity, defined as: (4) AIC=2k−2ln⁡(L^ Where k is the number of estimated parameters and L̂ is the maximum likelihood of the model. In the context of least-squares fitting, this is derived from the sum of squared errors. To ensure valid comparisons, AIC values were computed using identical data ranges for both candidate models (comparing fits on the full dataset against one another or fits on the truncated dataset against one another).

For analysis of ‘Learning Settling Time”, we calculated the absolute percentage deviation of cumulative hit rate curves relative to the final time point in each curve i.e., their Final Hit Rate (FHR), we then calculated back in time to determine where fitted curves first deviated from the FHR by ±10%. Note that FHR is the final data point of the cumulative hit rate i.e., after all responses have been accumulated in that listening environment.

Statistical analysis

Repeated measures ANOVAs (rANOVA) were performed to assess speech understanding (d’) within subjects with factors: Rooms (levels: Lecture Room, Open-Plan Office and Underground Car park); carrier phrase length (levels: CP0, CP1, CP2 and CP3) and target word (levels: |Color| and |Number|). A mixed ANOVA was employed to assess whether speech understanding was affected by the between factor: TMS exposure (levels ‘sham’ and ‘real’ TMS), and within-subject’s factors: Rooms (levels: Lecture Room, Open-Plan Office and Underground Car park); carrier phrase length (levels: CP0, CP1, CP2 and CP3) and target word (levels: |Color| and |Number|). Additionally, three follow up experiments were performed and a one-way ANOVA was used to assess whether speech understanding was affected by the room-context (i.e., the 3rd room in which Open-Plan Office and Underground Car Park were learned). Only one between-subject’s factor was assessed: room context.

Effect sizes were calculated for all post hoc statistical analyses (Partial Eta-squared (ηp2), Cohen’s d (parametric analysis) and Person’s r (for nonparametric Wilcoxon Signed-Rank test) (Pallant, 2011). Further two-tailed t-tests (original alpha=0.05, with post hoc Bonferroni corrections for multiple comparisons) were performed. Corrected alpha value is not reported as all statistical analyses were performed in SPSS: RRID:SCR_002865 (IBM Corp. Released 2023. IBM SPSS Statistics for Windows, Version 29.0.2.0 Armonk, NY), and SPSS output adjusts the p-value instead of the alpha value itself.

Power Analysis

Initial sample size estimation for individual ANOVA analysis (18) was computed using G*Power: RRID:SCR_013726 (Faul et al., 2007) (Effect size f = 0.25; α err prob = 0.05; Power (1-β err prob) = 0.95). However, given the large effect size observed in previous studies (Brandewie & Zahorik, 2010, 2013; Srinivasan & Zahorik, 2012; Pavel Zahorik & Brandewie, 2016) using sample sizes = 9-16 listeners, we expected relatively large effect sizes with a sample ≥ 10. The variance of all variables was tested for normal distribution (Shapiro-Wilk test, Shapiro & Wilk, 1965).

Results

Human listeners can incorporate knowledge of a listening environment, specifically its reverberant characteristics, to improve speech understanding in noise (Brandewie & Zahorik, 2010, 2013; Srinivasan & Zahorik, 2012; Zahorik & Brandewie, 2016). To understand how this learning of sound environments is achieved, we recreated acoustic characteristics of real rooms using an array of loudspeakers located in an anechoic chamber (Figure 1A) and assessed listeners’ performance in a speech-in-noise task using sentences from the Coordinate Response Measure (CRM) corpus—“Ready |Callsign| go to |Color| |Number| now” (Figure 1B). We employed six environments, Figure 1A, that varied from fully anechoic to a highly reverberant underground car park with RT60iso = 2.42 s (RT60s according to ISO-3382 guidelines (ISO, 2009).

The CRM corpus is commonly used to quantify listening performance in noisy or cluttered environments (Bolia et al., 2000). Listeners reported the |Color| (one of four monosyllabic choices, ‘red’, ‘white’, ‘blue’, or ‘green’) and the |Number| (the English digits between ‘one’ and ‘eight’) they heard for sentences of varying duration where all, some, or none of the preceding sentence (the ‘carrier phrase length’) before ‘|Color| |Number|’ was included (Figure 1B). Importantly, words in the CRM phrase preceding ‘|Color| |Number|’ are uninformative as to the color and number (Brandewie & Zahorik, 2010, 2013). CRM sentences—complete or partial (see Methods and Figure 1B)—were presented as if originating from in front of a participant, whilst a randomly generated white noise was presented from a source 90° to their left. Listeners’ abilities to recognize |Color| and |Number| from whole or partial CRM sentences were assessed. Although sentences varied in duration from trial to trial, they always contained the element ‘|Color| |Number| now’ embedded in noise and convolved with the impulse response of the real rooms (Figure 1C). Sentence length varied by adjusting the duration of the preceding ‘carrier phrase’ (CP; Figure 1B-C), which was always constructed from the same CRM sentence embedded in noise and convolved with the same impulse response as the remainder of the phrase. Thus, for CP0, listeners heard only ‘|Color| |Number| now’, CP1—‘go to |Color| |Number| now’, CP2—‘|Callsign| go to |Color| |Number| now’, CP3—‘Ready |Callsign| go to |Color| |Number| now’.

After each phrase was presented, participants verbally reported the |Color| and |Number| they heard to the experimenter and performance was assessed based on keywords correctly identified. CRM phrases of varying length were spoken by one of 6 talkers (3 female, 3 male) and the different environments were presented in a pseudorandom order to avoid the same environment being presented in consecutive trials (e.g., Figure 1C). Our initial assessment, Figure 1D, examined listening performance in three environments: a Lecture Room (RT60iso = 0.45 s), an Open-Plan Office (RT60iso = 0.96 s), and an Underground Car Park (RT60iso = 2.42 s). We confirmed, in 22 listeners, that understanding speech in background noise depends on the reverberant characteristics of the listening spaces from which impulse responses were obtained (Figure 1D (expanded datasets) and Figure 2A (specific comparisons between environments)). Overall, listeners performed better (quantified in terms of hit rate for reporting the correct |Color| and |Number|) in the Lecture Room—the least reverberant of the 3 environments (Figure 1D and 2A)—with a significant main effect of room [rANOVA: F (2,42) = 76.75, p<0.001, ηp2 = 0.79]. Bonferroni-corrected post-hoc pairwise comparisons demonstrated that performance in the Lecture Room was significantly better than in the Open-Plan Office [mean difference = 0.41, p < 0.001] and Underground Car Park [mean difference = 1.05, p < 0.001], (Figure 2A). Performance was also significantly better in the Open-Plan Office when compared to the Underground Car Park [mean difference = 0.65, p < 0.001], Figure 1D and Figure 2A.

Learning effects in three real rooms.

A. Performance (d’) (including correct and incorrect responses to all |Colors| and |Numbers|) in the Lecture Room (LR), Open-Plan Office (OPO) and Car Park (CP). Performance includes all speakers and phrases durations. Mean d’ denoted as circles in the boxplot (n=22 for all rooms). The horizontal line denotes the median. Upper and lower limits of the boxplot represent 1st (q1) and 3rd (q3) quartiles respectively, while whiskers denote the interquartile range (IQR =q3-q1). B. Time course of performance in each room for 22 listeners. Correct responses (Hit Rate) across time spent in each room are shown, solid curves in color correspond to mean data after 5-time point moving averages, equivalent to ∼7s, were applied). Optimal time course-fittings (using a two-phase association model) are plotted as color markers in each room. Associated shaded areas representing standard errors of the mean. Dashed coloured lines represent the time point at which participants reached a stable performance in each environment [i.e., ± 10% of the Final Hit Rate (FHR)], here ‘Learning Settling Time (LST)’, this variable was computed for each listener. Notice that no statistical differences were found for LST among the different rooms. C. Shows a correlation analysis between FHR and SLT in each environment. Significant negative Pearson correlations were observed for Lecture Room, and Open Plan Office, suggesting the earlier participants reach a stable performance, the higher their FHR is. D. Performance for each carrier phrase length, d’ was significantly better as the ‘carrier phrases length’ increased except for CP2 vs. CP3 (n.s), where a roll over effect was observed. E. Performance to |Color| and |Number| for CP0 i.e., our proxy for short-term adaptation the carrier phrase where exposure to an environment remained minimal. Here all the environments have been collapsed i.e., rANOVA, main effect of “target word”. F-G. Hit Rate for Initial (1-2) and Steady (9-10) trials for keyword |Color| in the CP0 condition: (F) in Reverberant conditions, indicating a significant improvement in performance for Steady trials, likely due to the accumulation of acoustic/environmental knowledge i.e., meta-adaptation (G) in anechoic (no echoes) condition, showing a lack of improvement in performance i.e., lack of meta-adaptation in the absence of reverberation. All data available: https://doi.org/10.25949/24295342

Statistical learning of reverberant environments occurs over long and short-time courses

Statistical learning likely occurs over different time courses subject to a range of possible brain mechanisms controlling or modulating the different cadences over which learning emerges (Robinson et al., 2016; Simpson et al., 2014). We sought to distinguish short-term from longer-term learning by assessing how prior exposure to the listening environment benefits word recall over multiple time courses. The design of our paradigm—with talker, length of carrier phrase, target words |Color| & |Number| and reverberation time (listening environment) constituting random variables—means that the initial phase of learning may also be highly variable; listeners, idiosyncratically, likely experience very different parameters over the first few trials, making it difficult to assess the rate at which learning accumulates over these early epochs, and we indeed found this to be the case (see Supplemental Figure 1).

To characterize behavioural response trajectories, we fitted data with single and double exponential functions. Double exponentials were fitted to the full time course (Supp. Fig 1 and 2). Single exponentials were fitted to both the full time course (Supp. Fig. 1) and a truncated version excluding the first 10 points (∼14 s) to mitigate initial variability (Supp. Fig. 2). AIC (Akaike Information Criterion) comparisons heavily favoured the double exponential model for the full time course (mean ΔAIC Double Vs Single (fitted to full time course) = −61.41 ± 7.13), providing a better fit for ≥ 20/22 subjects in all three environments. When compared against the truncated single-exponential fit, the double exponential model still produced smaller AIC values on average (mean ΔAIC Double Vs Single (fitted to truncated timecourse) = −21.55 ± 5.90), and remained the preferred model for approximately half the subjects. Crucially, the double exponential model was preferred not only for its automated nature (requiring no manual truncation) but also for the magnitude of improvement: when the single model was superior, the advantage was marginal (ΔAIC = 9.7 ± 1.4), whereas the double model’s advantage was substantial (ΔAIC = 52.8 ± 8.9). We therefore used the double exponential fit for further analysis.

Using these fittings we assessed the ‘learning settling time (LST)’—defined here as the time at which participants achieved stable performance in each environment and quantified as the time point at which performance stabilised to ± 10% of the Final Hit Rate (FHR). For the 22 listeners, LST (Figure 2B; Supplemental Figure 2A-C) was similar across all rooms (a total of 180 trials per room and average trial length of 1.5s): Lecture Room: [45.45 ± 41.70 s]; Open-Plan Office: [58.46 ± 32.34 s] and Car Park: [64.52 ± 34.47 s], with a one-way ANOVA revealing no statistical differences in LST for the three reverberant environments [F(2,42)=2.33, p=0.11, ηp2 =0.10] suggesting listeners learned/adapted to these environments at the same rate.

We next wondered whether the FHR achieved by a participant within an environment was related to the time point at which LST was reached. To this end, a Pearson’s correlation analysis revealed FHR and the timepoint at which stable performance was achieved were negatively correlated for the Lecture Room: r(22) = −0.66, p < 0.001, 95% CI [−0.86, −0.31] and Open Plan Office r(22) =-0.50, p = 0.02, 95% CI [−0.74, −0.20], but not for the more-highly reverberant Car Park r(22) = −0.08, p = 0.73, 95% CI [−0.44, 0.35], Figure 2C. This analysis suggests that the earlier in time a listener achieves asymptotic performance (i.e., the time point defined as global learning)—presumably by accumulating knowledge about these environments over time—the higher their performance (defined by FHR) in the task, at least in the two less-reverberant environments assessed here.

We next analysed whether speech understanding—quantified in terms of hit rates—improves as the length of the ‘carrier phrase’ was increased (Figure 2D). Consistent with Brandewie and Zahorik (2010; 2013), we found a significant main effect of length of ‘carrier phrase’: [rANOVA: F (3,63) =26.59, p<0.001, ηp2 =0.56]. Bonferroni-corrected post-hoc pairwise comparisons revealed a significant effect for most carrier phrase lengths (Supplemental Table 1) other than between the two longest, CP2 vs. CP3. This plateau in performance between CP2 and CP3 suggests an upper limit in the ability to exploit/accumulate information in reverberant listening environments with increasing sound duration, consistent with Brandewie and Zahorik’s (Brandewie & Zahorik, 2013) observation of a plateau (and a decline in some reverberant environments) in listening performance with increasing length of carrier phrase, and the existence of an upper limit of listeners to exploit prior exposure to sound environments to benefit listening (Hicks & McDermott, 2024; McDermott et al., 2013; McWalter & McDermott, 2018).

We hypothesized that if prior exposure to the statistics of reverberant rooms arises from an increase in the length of the carrier phrase, then performance to |Number| will always be better than to |Color| for the shortest carrier phrase (CP0: ‘|Color| |Number| now’; Figure 2E) as |Number| is subject to a longer preceding phrase than |Color|. Assessed in terms of reporting |Color| and |Number| alone, listeners performed significantly better (i.e., showed greater sensitivity) for |Number| compared to |Color|: main effect of target word: [rANOVA F (1,21) =48.79, p<0.001, ηp2 =0.70], [mean difference = 0.38, p < 0.001] despite there being twice as many (eight compared to four) possibilities. Moreover, a significant interaction ‘carrier phrase length’ x ‘target word’ was observed: [rANOVA F (1,21) =4.33, p=0.008, ηp2 =0.17]. Post-hoc pairwise comparisons (see Supplemental Table 2) indicate that for all carrier phrases except CP3, performance to |Number| was significantly higher than for |Color|. This result also suggests that a roll-over effect i.e., a lack or limit to the improvement in performance is evident for carrier phrases longer than CP2 across environments (Brandewie & Zahorik, 2013; McWalter & McDermott, 2018).

A potential explanation for poorer performance in reporting |Color| relative to |Number| for the shortest carrier phrase (CP0) is the impact of the utterance of |Color|, namely its immediate appearance at the start of the phrase i.e., the impact of “order/certainty or predictability” in statistical learning (Conway et al., 2010; Daikoku & Yumoto, 2023). One way to determine whether |Color| presented in the context of the CP0 is at all subject to statistical learning (and is therefore not solely reliant on the short-term accumulation of information within a CP0 trial; see Figure 3G) is to assess whether performance for |Color| for the shortest carrier phrase (CP0) improves from the longer-term accumulation of knowledge i.e., the process of meta-adaptation (Robinson et al., 2016) in which adaptation to short-term statistics improves with repeated exposure to those statistics (after 8 exposures). Specifically, we tested the hypothesis that, independent of the environment, performance for |Color| in later steady trials (here, trials 9-10) of the shortest carrier phrase, CP0, are better compared to initial (1-2) trials (Figure 2F). Across the total of 360 trials (across carrier phrase lengths and environments), CP0 initial trials occurred early (trial 6±4), while steady CP0 trials (9th–10th occurrences) did not emerge until trial 38±7. If this hypothesis is supported, it suggests performance for |Color| for CP0 benefits from meta-adaptive information conveyed in later trials as knowledge about the global structure of the environment accumulates over time.

Learning effects in ‘sham’ and ‘TMS’ conditions.

A. Schematic representation of functional connections between the dorsolateral prefrontal cortex (dlPFC) and primary auditory cortex (A1) under transcranial magnetic stimulation (TMS). Although the TMS protocol was intended to target the dlPFC, it likely affected adjacent prefrontal regions, leading to diffuse downstream effects that may have included modulation of A1. B. Average exposure time (s) ‘sham TMS’ and ‘real TMS’ conditions. Correct responses across time spent in each room for ‘sham’ and ‘TMS’ conditions. Solid curves in pink and gray correspond to mean data from ‘TMS’ and ‘Sham’ conditions respectively (after 5-time point moving averages, equivalent to ∼7s, were applied), with the associated shaded areas representing standard errors of the mean. Dashed black (sham TMS) and pink (real TMS) lines represent the time point at which participants reached a stable performance in each environment [i.e., ± 10% of the Final Hit Rate (FHR)], here ‘global learning’, no statistical differences were found for ‘global learning’ between TMS conditions. C. Shows a correlation analysis between FHR and the time where 10% of the FHR is achieved in under TMS (pink) and Sham (gray) conditions. Significant negative correlations (Pearson) were observed under Sham stimulation, suggesting the earlier participants reach a stable performance, the higher their FHR is. This relationship was disrupted (lack of significant correlation) under TMS conditions. D. Performance in TMS and Sham exposed participants i.e., collapsed sensitivity (d’) to |Color| and |Number| for all carrier phrases and environments (significant main effect of TMS conditions (between-subject’s factor): mixed ANOVA). d’ values plotted in bright pink corresponds to the ‘real TMS’ condition whereas the ‘sham TMS’ accuracy is plotted in gray. Mean d’ is denoted as circles in the boxplot (n=11 for ‘sham TMS’ conditions and n=10 for ‘real TMS’ conditions). E. Significant Interaction ‘TMS condition’ x ‘carrier phrase length’. d’ was significantly better as the ‘carrier phrases length’ increased for ‘sham TMS’ compared to ‘real TMS’ except for CPO (n.s), i.e., our proxy for short-term adaptation and minimal exposure to acoustic environments. F. Performance to |Color| and |Number| for CP0 i.e., our proxy for short-term adaptation and the carrier phrase where exposure to an environment remained minimal. Here all the environments have been collapsed as only a main effect of “target word”, performance to number always significantly higher than for colour independent of TMS condition. G. Schematic for assessing short-term adaptation (performance to |Number| in the Initial Trials) and meta-adaptation (performance to |Color| between Initial and Steady Trials). H. Isolated short-term adaptation effects for performance to |Number| in CP0. Hit Rate for Initial (1-2) trials only for |Number| in the CP0 condition for both TMS (pink) and Sham (gray) conditions. Notice that no differences were observed between performance |Number| under Sham or TMS conditions, suggesting no disruption of short-term adaptation. I. Hit Rate for Initial (1-2) and Steady (9-10) trials only for |Color| in the CP0 condition for both TMS (pink) and Sham (gray) conditions. Notice that only a significant improvement in performance for from Initial to Steady trials was observed in the ‘sham TMS’ condition. All data available: https://doi.org/10.25949/24295342

Consistent with this hypothesis, a Wilcoxon signed rank test revealed significant better performance (n=22, Z= −2.16, p=0.03, medium effect size (r=0.46)) for |Color| for steady trials (9-10; assumed to reflect a meta-adaptive state; Robinson et al., 2016) compared to initial trials (1-2; i.e., short-term adaptation). Interestingly, in anechoic conditions i.e., in the absence of reverberation, this meta-adaptive process—the expected improvement in performance between initial and steady trials—was not observed: Wilcoxon signed rank test [n=10, Z= −0.71, p=0.48, small effect size (r=0.23)], with Initial trials appearing also early in the task: 6±3, whereas steady trials did not appear later until trial 36±8. This suggests that performance for |Color| conveyed in the shortest carrier phrase, CP0, improves over time even in the absence of immediate information in the form of any preceding carrier phrase, with knowledge of the statistical/acoustical properties of environments accumulating over the course of the experimental task.

Improvements in performance are explained by exposure to the environment not talker idiosyncrasies

Normal-hearing listeners are reported to understand speech slightly better when listening to male talkers in a mixture of male and female talkers (Larsby et al., 2015), and quickly adapt to talkers’ idiosyncrasies such as non-native accented speech (Idemaru & Holt, 2011, 2014; Liu & Holt, 2015). Here, we wondered if exposure to the different talkers (3 female and 3 male) could explain the improvement in speech as the length of the carrier phrase was increased in the 22 listeners (Figure 2D). We performed a rANOVA with the rooms, carrier length and talker identity as within-subject’s factor. Despite listening performance to some talkers (Supplemental Figure 3A) appearing better than for others [main effect of talkers: [rANOVA: F (5,105) = 27.19, p<0.001, ηp2 =0.56], i.e., significantly worse performance for Talker 1 (male) and Talker 6 (female), see Supplemental Table 3 for details, overall differences in performance due to a specific talker did not explain the benefit to speech understanding as the length of the carrier phrase was increased in each environment (i.e., lack of significant interaction: or ‘talker’ x ‘carrier’ x ‘room’: F (30,630) = 0.73, p=0.85, ηp2 =0.034]), Supplemental Figure 3 for details.

To explore further the possibility of any talker-specific effects on performance, we assessed the potential benefit of experiencing the same talker on consecutive trials across an experiment, hypothesizing that, independent of room characteristics or length of carrier phrase, experiencing the same talker on consecutive trials would provide a benefit to listening performance. Across all our participants, a total of 1262, 217, and 40 trials were identified as having two, three, or four consecutive trials in which the same talker appeared (see Supplemental Figure 4). Despite the possibility that sustained experience of the same talker might lead to improved listening performance, a Wilcoxon signed rank test revealed no significant differences in performance between trial 1 vs. 2 consecutive/same talker trials (n=1262, Z= −0.42, p=0.68, small effect size (r= 0.01)), trial 1 vs. 3 consecutive/same talker trials (n=217, Z= −0.17, p=0.87, small effect size (r= 0.01)) or trial 1 vs. 4 consecutive/same talker trials (n=40, Z= −0.159, p=0.11, small effect size (r= 0.02)). Our data suggest that listeners do not use talker identity, at least in the task reported here, to benefit their speech-in-noise understanding.

We also explored the extent to which a talker and/or the task itself could be learned by analysing improvements in performance with increasing exposure to carrier phrases and talkers in the absence of reverberation i.e., under anechoic conditions. In a follow up experiment we assessed performance (similarly to our 3 initial rooms (Brandewie & Zahorik, 2013)) but swapping the Lecture Room with an Anechoic room in 10 naïve listeners. Here, we isolated only the Hit Rate in the Anechoic room and performed an rANOVA with only carrier phrase length as the within-subject’s factor. Although a significant rANOVA was observed: F (3,27) = 9.10, p=0.007, ηp2 =0.57; post-hoc pairwise comparisons, with Bonferroni corrections, revealed that only a significant improvement in performance was observed between CP0 and CP2 [mean difference = 15.67, p = 0.02; see Supplemental Table 4], suggesting that in anechoic conditions very little improvement in performance is achieved by learning the talker/noise/location/task as length of carrier phrase increases. These data support us employing a listening task with high ecological relevance—comprehending speech—to demonstrate the potential benefits of learning background acoustic features to leverage listening performance without requiring listeners to attend to or report features related to the statistics of presumed background sounds per se (e.g., Agus et al., 2014, McWalter & McDermott 2018, Bianco et al., 2020).

TMS disrupts long- but not short-term adaptation to an environment’s reverberation profile

The capacity for learning the statistical structure of acoustic environments and use this knowledge to better understand speech embedded in background noise, suggests a real-world benefit to this form of learned listening. To determine possible brain mechanisms underlying this ability, we reversibly impaired dorsolateral prefrontal cortex (dlPFC, Figure 3A) —a brain region implicated in listening performance in noise—using repetitive transcranial magnetic stimulation (rTMS) and then assessed the ability of listeners to recall |Color| and |Number| in our modified CRM sentences. The procedure began with TMS manipulation, and although the behavioural task lasted 45 minutes, the inhibitory effects of TMS extended for at least 60 minutes post-stimulation (Gamboa et al., 2010; Hoogendam et al., 2010; Huang et al., 2005; Romero et al., 2022). Specifically, 10 naïve, normal-hearing listeners were subjected to ‘real’ TMS—continuous theta burst on dlPFC for 40s on each side—and then transferred to the anechoic chamber, where they were presented sequences of CRM phrases of varying phrase length in background noise convolved with one of the three listening environments (Lecture Room, Open-Plan Office, and Underground Car Park) as before. A second, control, group of 11 naïve participants underwent ‘sham’ TMS stimulation in which an otherwise-identical TMS procedure was performed with a ‘sham’ TMS coil. All participants were naïve to differences in the TMS procedure as well as to the listening task; indeed, as with the experimental group, participants in the ‘sham’ group experienced a short period of single-pulse TMS stimulation to obtain motor thresholds, followed by the ‘sham’ TMS with the stimulator positioned over dlPFC. This process familiarised all participants with the procedure and the influence of TMS in generating involuntary finger movements through direct stimulation of motor cortex. The ‘sham’ procedure was designed to maintain participant blinding, such that participants in the ‘sham’ group would not know whether the subsequent speech-recall task followed active or ‘sham’ stimulation.

Given the potential placebo effects of a ‘sham’ TMS stimulation, we first tested whether our sample of 11 ‘sham’ TMS participants exhibited similar behavioural performance to the larger sample of 22 participants who had not been exposed to any TMS manipulation. A mixed ANOVA with a between-group factor of Condition and within-group factors of room and CP length confirmed that performance in these two populations was comparable, with no significant main effect of TMS conditions observed (‘sham’ vs. no exposure to TMS): [F (1,31) =0.01, p=0.91, ηp2 = 0.00], and no significant interaction Condition x CP length was observed: [F (3,93) =0.48, p=0.69, ηp2 = 0.01]; confirming that participants experiencing ‘sham’ TMS did not perform significantly differently from the ‘no exposure to TMS’ population.

We first assessed learning over the time-course of the task itself—‘Learning Settling Time—as the cumulative mean hit rate for listeners reporting |Color| and |Number|, fitted for each TMS condition and environment with a two-phase association model (Figure 3B and Supplemental Figure 5 and 6). Our metric for ‘Learning Settling Time’ was achieved for Lecture room at: ‘sham’ TMS [40.22 ± 34.17 s] and ‘real’ TMS [64.86 ± 17.08 s]; for Open-Plan Office: ‘sham’ [51.50 ± 33.26 s] and ‘real’ TMS [67.32 ± 37.04 s]; and for Car Park: ‘sham’ [64 ± 39.02 s] and ‘real’ TMS [69.80 ± 37.31 s]. A Mixed ANOVA between ‘sham’ TMS and ‘real’ TMS participants revealed a between-subject’s effect of TMS Conditions: [F (1,19) = 184.42, p<0.001, ηp2 = 0.91], suggesting participants always achieved an early LST under ‘sham’ stimulation [51.91 ± 36.11s] compared to longer LST under ‘real TMS’ [71.76 ± 36.09 s]. No significant Rooms or Rooms x TMS condition effects were observed.

Pearson’s correlation analysis (collapsed across all environments) revealed a negative correlation between FHR and ‘Learning Settling Time’ for ‘sham’ TMS listeners (see Figure 3C): [r(33) =-0.45, p = 0.01, 95% CI [−0.70, −0.11] but not for ‘real’ TMS listeners: [r(30) = −0.21, p = 0.27, 95% CI [−0.53, 0.16]. Specifically, participants undergoing ‘sham’ TMS who achieved stable performance relatively early in the task and maintained this level of performance throughout. Relative delay in attaining stable performance was associated with a lower FHR, with the time at which early LST is achieved predicting overall task performance. This relationship was not evident in listeners subject to ‘real’ TMS, where the time to reach a stable performance was not predictive of the magnitude of Final Hit Rate. Under TMS manipulation of dlPFC, LST had no predictive value, indicating that task performance was less stable and reliable for participants exposed to TMS.

We next employed a mixed ANOVA to compare performance (d’), (see Figure 3D)—of ‘sham’ participants with those who received ‘real’ TMS, and observed a significant between-subject’s effect of TMS condition: ANOVA [F (1,17) = 6.18, p=0.02, ηp2 = 0.28], where ‘sham’ participants showed significantly better abilities to recall |Color| and |Number| compared to participants subjected to ‘real’ TMS [post-hoc pairwise comparisons with Bonferroni corrections: [mean difference = 0.46]. In addition, a significant interaction ‘TMS condition’ x ‘carrier phrase length’ was observed, Figure 3E: [F (3,51) = 6.72, p<0.001, ηp2 = 0.28]. A post-hoc pairwise comparison with Bonferroni corrections showed that performance in the CP0 condition was not statistically different between ‘sham’ and ‘real’ TMS conditions (mean difference = 0.14, p = 0.49), however performance was better for ‘sham’ compared to ‘real’ TMS for CP1 (mean difference = 0.47, p = 0.04), CP2 (mean difference = 0.51, p = 0.02), and CP3 (mean difference = 0.71, p < 0.001). Listeners exposed to ‘sham’ TMS retained the improvements of performance as the length of carrier phrase was increased whereas those listeners receiving ‘real’ TMS—in which activity in dlFPC is presumably impaired—appeared unable to accumulate over time knowledge of the sound environment.

The different time courses over which statistical learning might occur suggests the involvement of multiple neural mechanisms and brain circuits (Anderson & Malmierca, 2013; Antunes & Malmierca, 2011; Robinson et al., 2016), with different cadences of learning potentially controlled, or at least modulated, by different brain centres. The time-course of midbrain neurons to adapt to different sound environments, for example, is slowed, and the capacity for retaining a ‘memory’ of those sound environments disappears, when cortex is inactivated through cooling (Robinson et al., 2016). This suggests feed-forward and feed-back mechanisms, with their own time-courses or time-constants, contribute to performance, including the rate at which sound environments are learned. We specifically wondered if ‘real’ TMS had a detrimental effect on short-term adaptation i.e., the observed advantage in performance for |Number| compared to |Color|, for the shortest carrier phrase CP0 (Figure 2E and 3F). Mixed ANOVA revealed a main effect of ‘target word’, where performance for |Number| was always significantly better than for |Color| for all the environments explored and across TMS conditions [F (1,17) = 31.99, p<0.001, ηp2 = 0.65]. However, no interaction between ‘TMS conditions’ and ‘target word’ was observed, indicating that in both ‘real’ and ‘sham’ TMS conditions, the same trend of better performance to |Number| compared to |Color| was observed (Figure 3F).

To dissociate specifically short-term adaptation from longer-term meta-adaptation, we compared performance (Hit Rate) to |Number| for initial trials (1-2), for CP0 only, as performance for this length of carrier phrase should be influenced only by short-term accumulation of information (i.e., within trial; see Figure 3G). Note that for both groups, CP0 ‘initial’ trials occurred early in the task (360 total trials): trial 5±4 for ‘sham’ TMS and trial 10±9 for ‘real’ TMS. However, ‘steady’ trials (9th–10th instances) did not emerge until trial 34± 6 and 38± 12, respectively.

A non-parametric Wilcoxon signed rank test revealed no significant differences in |Number| performance for initial trials between ‘sham’ and ‘real’ TMS exposed listeners (n=10, Z= −0.28, p=0.78, with a small effect size (r= 0.01); Figure 3H), suggesting that rapid learning of the statistical structure of the reverberant environment—in the order of a few hundreds of milliseconds—is resistant to the effects of ‘real’ TMS applied to dlPFC. To determine how much meta-adaptative improvement in performance could be disrupted by applying ‘real’ TMS to dlPFC, we then compared initial trials (1-2) and steady trials (9-10) for CP0—our proxy for meta-adaptation (Figure 3G). We tested the hypothesis that disruption of dlPFC would affect late (meta-) but not early adaptation. To this end, we expected no differences between performance in initial trials between ‘real’ and ‘sham’ TMS (as initial trials are influenced only by short-term adaptation), and a relative lack of improvement in performance in later, steady trials (meta-adaptation influenced) in ‘real’ compared to ‘sham’ TMS. Consistent with our hypothesis, a Wilcoxon signed rank test revealed no significant differences in performance for initial trials between ‘sham’ and ‘real’ TMS exposed listeners (n=10, Z= −0.21, p=0.83, with a small effect size (r = 0.01), but improved performance for steady compared to initial trials (n=10, Z=-2.06, p=0.04, large effect size (r=0.65)) for ‘sham’ TMS only (Figure 3I). These data suggest that performance in |Color| for the shortest carrier phrase, CP0, was disrupted due to impaired accumulation of knowledge about the environments encountered. In turn, performance under ‘TMS’ in steady trials (presumed to be influenced by meta-adaptation) was reminiscent of performance observed in initial trials where only immediate knowledge could be used to improve performance. This is reminiscent also of adaptation to sound environments reported in vivo (Dean et al., 2005, 2008; Robinson et al., 2016; Wen et al., 2009), where interrupting efferent feedback by cortical cooling impairs the capacity of midbrain neurons to learn the statistical structure of sound environments; though neurons adapt to the short-term statistical structure of a sound environment each time they are exposed to it, they fail to demonstrate the acceleration of adaptation and improvement in (neural) discrimination performance as the same environment is re-encountered, adapting to it only as if exposed to it the first time.

Statistical learning of room acoustics is tuned to universally experienced reverberation times

Reverberation is a common feature of many acoustic environments—natural and built (Traer & McDermott, 2016). Interestingly, Zahorik and Brandewie (2013) reported that improvements in speech understanding when a ‘simulated room’ is repeatedly encountered were best when listeners experienced moderately reverberant environments. Specifically, we hypothesized that the ability to understand speech in noise depends on exposure to specific values of RT60 across our experimental paradigm, including those in the range commonly experienced by human listeners in natural and built environments (Traer & McDermott, 2016). To test this hypothesis, we recruited a new population of naïve participants with no previous exposure to our experimental paradigm and assessed their ability to understand speech in noise using the same CRM corpus as before, but in different combinations of acoustic environments i.e., room contexts defined by their RT60s.

We first tested the ability of a new sample of 10 naïve participants to recall |Color| and |Number| in Open-Plan Office and Underground Car Park as before, but with the Lecture Room (RT60iso = 0.42 s) swapped out for a more Highly Reflectant elevator lobby (Figure 4) with an RT60iso of 1.55 s—i.e., between that of the Open-Plan Office (0.96s) and the Underground Car Park (2.42s). We therefore changed the ‘room-context’ in which Open-Plan Office and Underground Car Park were learned. When contrasting the performance (Final Hit Rate) in an ANOVA considering performance only in the common rooms (Open-Plan Office and Underground Car Park) of these 10 naïve participants vs. the 11 subjects randomly selected from the initial 22 participants sample, we found that speech understanding in Open-Plan Office and Underground Car Park was significantly better when the room acoustics of these environments were learned in the context of the Lecture Room than in the context of the Highly Reflectant Room [mean difference = 14.8, F (1,41) = 30.72, p<0.001, ηp2 = 0.45].

Performance in different 3-room combinations where acoustics of 3 rooms span an ecological range.

A. Hit rates over time for different 3-room combinations. Mean data for four different 3-room combinations are plotted where Open-Plan Office (dark blue) and Underground Car Park (yellow) were always included and the context room was either Anechoic (black), Living Room (red), Lecture Room (green), Lecture Room (green) or Highly Reflectant (purple). B. Performance in Open-Plan Office and Car Park when paired with different context rooms. Final hit rates were highest for Office and Car Park when presented in conjunction with the Lecture Room, followed by the Living Room and Anechoic with poorest performance observed when the Highly Reflectant space was the context. C. Environments tested compared to ecological range. The RT60traers for the five test rooms, calculated as the median RT60 value across 31 frequency sub-bands (Traer & McDermott, 2016), are plotted (colored boxes) above a histogram of the median RT60traers calculated for 199 indoor (filled bars) and 72 outdoor (empty bars) spaces recorded by Traer and McDermott (2016). D. Frequency dependence of reverberation time (RT60traer) in test rooms compared to ecological range. RT60s of the 5 test rooms are displayed for frequency sub-bands used to calculate RT60traer (colored lines). Quartiles for combined indoor and outdoor spaces (Traer & McDermott, 2016) are plotted as black dashed lines. Decreasing reverberation time above 0.5kHz is observed for Living Room, Lecture Room and Car Park as has been typically described for indoor spaces (Traer & McDermott, 2016); however RT60traer profiles for Open-Plan Office and Highly Reflectant space were notable for their longer reverberation times at and above 1kHz. All data available: https://doi.org/10.25949/24295342

Wondering whether the ‘room-context’ effect was determined by task difficulty i.e., the more-highly reverberant lobby being intrinsically more difficult for listening than the Lecture Room [mean difference = 35.75, t (9) =18.72, p < 0.001, Cohen’s d = 5.92]. We recruited a further new sample of 11 naïve participants and performed the same task in which they were exposed to another combination of three rooms, here with the Lecture Room swapped for an Anechoic Room. Anechoic spaces—rooms whose walls are treated to completely absorb reflected sound energy—have been extensively reported to aid speech understanding, especially in background noise. To test this hypothesis directly, we recruited a further sample of 10 naïve participants and compared their speech-in-noise performance in a combination of an Anechoic Room, Open-Plan Office and Underground Car Park, with the performance of 11 subjects randomly selected from the initial 22 participants’ sample. Although task difficulty might be expected to be at its lowest i.e., easiest, in the less-reverberant Anechoic Room, this was not the case, performance in the Anechoic room was poorer compared to Lecture Room [mean difference = 11.08, t (9) =2.66, p = 0.013, Cohen’s d = 0.84] (Figure 4 A-B). Additionally, we found that performance in the common rooms: Open-Plan Office and Underground Car Park, was significantly better when these environments were again learned in the context of the Lecture Room: [mean difference = 9.7, F (1,41) = 13.24, p<0.001, ηp2 = 0.26].

Although reverberation might be considered detrimental to speech-in-noise performance, short-term benefits of more, compared to less, room reverberation for speech understanding have, in fact, been reported, including for the task we report here (Brandewie & Zahorik, 2010, 2013). Furthermore, it is likely that listening performance in noise is indeed better when listeners have access to more commonly experienced, or ethologically realistic, levels of reverberation. We tested this hypothesis by recruiting a new sample of naïve listeners and swapping again the Lecture room with a Living Room with a shorter RT60iso (0.33 s). When comparing performance between Lecture and Living Room no significant differences were observed: [mean difference = 7.04, t (10) =1.69, p = 0.061, Cohen’s d = 0.51] (Figure 4 A-B). However, significantly better performance was observed in the common rooms: Open-Plan Office and Underground Car Park when these environments are learned in the context of the Lecture room: [mean difference = 9.7, F (1,43) = 5.79, p=0.21, ηp2 = 0.13]. These results suggest that when certain reverberant characteristics are encountered e.g., Lecture Room, they might unlock a perceptual advantage in understanding speech in harder/more reverberant environments such as Open-Plan Office and Indoor Car Park.

Seeking to explain why some rooms might be better than others in aiding speech understanding, we recalculated reverberation times for our 5 reverberant rooms according to the method of Traer and McDermott (2016) where RT60traer equals the median RT60 measured in 31 sub-bands with centre frequencies between 80 Hz and 10 kHz (see Methods). We then compared these with the values collected for a wide range (n= 271) of built and natural environments by the same authors (Figure 4C-D. RT60traers for 5 rooms conditions and Traer’s data, plotted as log10 RT60traer for visualisation). In particular, the RT60traer of the Lecture Room, at 0.49 s, lies extremely close to the median and mean RT60traer values of the built environments [0.42 s and 0.50 s respectively generated by the skewed distribution (n= 199), Figure 4C]. This suggests that superior performance in understanding speech understanding in noise in the Lecture Room—and the positive impact on speech understanding when the Lecture Room is included in the three-room listening task—is related to its reverberant characteristics being commonly encountered in everyday listening situations. Notably, the mean and median RT60traers of the recorded natural environments were much lower (0.12 s and 0.16 s, respectively [n = 72], Figure 4C) than the RT60traers of the Lecture Room. This suggests that brain mechanisms contributing to effective speech understanding in reverberant environments might have adapted to the range of environments experienced over some longer time course than we assessed here.

Discussion

We assessed the ability of listeners to understand speech embedded in background noise and convolved with the room impulse responses (RIRs) of different indoor environments defined by their reverberation decay time, or RT60—the time taken for reverberant sound energy to decay 60 decibels (dB). We found speech understanding improved with repeated exposure to an environment, a form of learning that was impaired following continuous, bilateral theta-burst transcranial magnetic stimulation (TMS) of dorsolateral prefrontal cortex (dlPFC). Specifically, we observed rapid learning i.e., an improvement in speech understanding over the first few seconds of room exposure, likely due to listeners learning the acoustic reverberation—the only variable held constant across trials for a given environment. This learning, on the timescale of several seconds, was impaired by TMS, whilst learning at shorter or longer timescales was not, suggesting that TMS applied to dlPFC specifically disrupted learning of the reverberant characteristics of an environment. Listeners also showed better listening performance in moderately reverberant environments, and this performance was transferable between different acoustic environments. Specifically, the ability correctly to report keywords spoken in environments with the more extreme—higher—RIRs was best when these environments were encountered in experimental blocks also containing the moderately reverberant environments. This transference suggests an ethological tuning to more commonly encountered environments when learning acoustic backgrounds.

A role for dlPFC in statistical learning of room acoustics

Inactivation of dlPFC using theta-burst TMS reduced performance in our speech-in-noise task, assessed in terms of the hit rate for |Color| and |Number| in CRM phrases as a function of duration of exposure to a sound environment. Notably, however, TMS did not impair listening performance for the shortest-duration phrase, ‘|Color| |Number| now’. For this duration of carrier phrase, performance i.e., the tendency to report more accurately |Number| compared to |Color|—despite |Number| having twice as many options as |Color|—was unaffected by TMS stimulation of dlPFC. The robustness of listening performance to the shortest carrier phrase suggests that the ability to hear out speech in background noise and potentially to learn this over time in complex acoustic environments does not rely solely on cortical feedback from dlPFC. The ability to leverage performance on |Number| by exposure to |Color| presented in background noise convolved with the room impulse response is evident even when the function of dlPFC is (presumably) impaired. It is the case, however, that the rate of learning—i.e., how rapidly performance improved with increasing time of exposure to an acoustic environment—was influenced when dlPFC was impaired. Under the influence of prior TMS stimulation, listeners were less able, and less reliable, when learning the acoustic background features of an environment to enhance their speech understanding, particularly in the more challenging environments of the Underground Car Park and Open-Plan Office.

Dense connections between dlPFC and the sensory cortices provide the conduit for prediction-based, top-down regulation (Morrone, 2010). These model-based predictions rely on learned templates of sensory environments (Alexander & Brown, 2018) that are critical for making inferences under situations of sensory uncertainty, such as those encountered in noisy and challenging listening environments (Bartolo & Averbeck, 2021). Dorsolateral prefrontal cortex can influence auditory processing by means of direct connections to auditory cortex (Hackett et al., 1999; Plakke & Romanski, 2014)—see Figure 3Ai—and to the auditory thalamic reticular nucleus, a key modulator of auditory thalamo-cortical loops (Zikopoulos & Barbas, 2006). In return, dlPFC receives projections from primary and secondary auditory cortex (Barbas & Pandya, 1987; Barbas & Pandya, 1991; Goldman-Rakic & Schwartz, 1982; Pandya & Barnes, 2019; Petrides & Pandya, 2002), indicating that a reciprocal auditory-to-prefrontal ‘listening loop’ exists to support hierarchical predictions and prediction-error feedback. These ‘listening loops’ could potentially extend to cortical efferent feedback (Blackwell et al., 2020; McAlpine & de Hoz, 2023) and modulate the function of the inner-ear sensitivity in attended speech-in-noise tasks (de Boer & Thornton, 2008; Garinis et al., 2011; Giraud et al., 1997; Mishra & Lutman, 2014; Hernández-Pérez et al., 2021).

The specific mechanism by which TMS-induced modulation of dlPFC impairs listening performance remains unknown. Neurostimulation studies targeting dlPFC during learning paradigms have generated conflicting directions of effect i.e., stimulation of dlPFC is reported to impair implicit learning (Nydam et al., 2018; Pascual-Leone et al., 1996) or enhance it (Ambrus et al., 2020). Ambrus and colleagues argued that impairment of dlPFC during implicit/statistical learning allows for increased engagement of a learning mechanism that is ‘model-free’ i.e., does not rely on inherent sensory representations and therefore takes longer to consolidate. In the present study, although the TMS protocol was designed to target dlPFC using the 10–20 EEG cap system, this approach does not permit precise stimulation of dlPFC (Jiang et al., 2025; Nikolin et al., 2019; Trapp et al., 2020) or its direct projections to the auditory cortex. Instead, stimulation likely influenced prefrontal regions producing diffuse downstream effects including the auditory cortex and other brain regions such as hippocampus (Bennett et al., 2011; Stillman et al., 2013), striatum (Albouy et al., 2015) and basal ganglia (Packard & Knowlton, 2002). We therefore speculate that the short periods over which learning is required in our task are subserved by a pre-frontally mediated, ‘model-based’ learning able to construct predictions ‘on the fly’ (Daw et al., 2005). We suggest, in our study, that disrupting a rapidly acting model-based form of learning supported by prefrontal regions, as part of the auditory-prefrontal ‘listening loop’, may lead to impaired learning of reverberant environments.

Dorsolateral prefrontal cortex is also implicated in several, high-level cognitive functions including multi-sensory integration (Fuster et al., 2000), executive functions such as inhibition and working memory (Castro-Meneses et al., 2016; Coltheart et al., 2018; Wang et al., 2015) and listening in noisy environments (Du et al., 2016). The impact of TMS on performance of our speech-in-noise task may be linked to the involvement of dlPFC in more generalized cognitive functions such as executive, memory, and attention processes. Dissociating the potential influence of TMS on such processes relative to a specific role in speech-in-noise understanding in noisy, reverberant environments would be an ideal next step. Specifically frontal regions have been linked to sensorimotor integration during speech-in-noise perception in young and older human participants (Du et al., 2016), suggesting that TMS might disrupt global functioning required for the analysis and understanding of speech.

Whilst not entirely unrelated, all participants completed a familiarization task preceding the main experiment, which consisted of 10 trials presented in anechoic conditions at 0 dB Signal-to-Noise Ratio (SNR). Participants exposed to either ‘real’ or ‘sham’ TMS underwent this task following cTBS procedures. Notably, all participants, including those subjected to cTBS, achieved 75-100% accuracy during the familiarization phase, with no statistical differences between TMS groups [mean difference=1.25, t(1,9)=0.43, p=0.68, d=0.14], confirming their comprehension of the task and successful procedural learning (i.e. learning the task per se). In addition, speech understanding in the most challenging condition (CP0) was not disrupted following TMS manipulations, suggesting that the effect of TMS was specifically in terms of influencing the way these environments were learned over time. Nevertheless, we acknowledge that the primary task may have imposed greater cognitive demands, potentially requiring a more significant involvement of the dlPFC to meet such task requirements.

Moreover, we cannot rule out individual variability in the duration of TMS effects, nor differences in effect duration across anatomical sites—such as the motor cortex (Gamboa et al., 2010; Hoogendam et al., 2010; Huang et al., 2005; Romero et al., 2022) versus the dLPFC examined in our study. Continuous theta burst stimulation (cTBS), as used in this study, typically induces aftereffects lasting 20–50 minutes (Huang et al., 2005; Wischnewski & Schutter, 2015). While these effects are well established in the motor cortex, where motor-evoked potential changes can persist for up to one-hour, comparable durations of cortical modulation have also been reported in the prefrontal cortex (Taylor et al., 2024; Wagner et al., 2006). Specifically, inhibitory rTMS applied to the dorsolateral prefrontal cortex (dlPFC) has been shown to reduce cortical oxygenation levels—indicative of decreased neural activity—for at least 45 minutes (Tupak et al., 2013). Given that changes in cerebral haemoglobin concentration closely correspond to neuronal activation (Liao et al., 2013), these findings support the use of functional Near Infrared Spectroscopy (fNIRS) as an indirect but reliable method to infer TMS-induced neural modulation within this timescale. Although no objective post-stimulation measures were collected in the present study, we acknowledge this as a limitation and plan to incorporate such measures in future work to confirm the duration and magnitude of TMS effects more directly.

What is being learned in statistical learning of acoustic features?

The need to communicate in reverberant environments is common to daily life. It is well known that long reverberation times have detrimental effects on speech quality and intelligibility (Brandewie & Zahorik, 2010, 2013; Srinivasan & Zahorik, 2012; Pavel Zahorik & Brandewie, 2016) and that, specifically, reverberation blurs phoneme boundaries, increasing the extent to which similar words might be confused, e.g., ‘sir’/’stir’ (Watkins, 2005b; Watkins & Makin, 2007). However, with sufficient exposure to longer reverberation times in the form of a carrier phrase, similar levels of word identification to those achieved in minimal reverberation are observed (Watkins, 2005b; Watkins & Makin, 2007). Based on the premise that providing sufficient contextual information enables the auditory system to adapt and compensate for the detrimental effects of reverberation, Zahorik and colleagues (Brandewie & Zahorik, 2010, 2013; Srinivasan & Zahorik, 2012; Zahorik & Brandewie, 2016) and, here, ourselves, demonstrate that prior exposure to the reverberant characteristics of a room can enhance sentence understanding when that environment is re-encountered. Moreover, Zahorik and colleagues, and our own data (see Figure 4A), demonstrate that improvements in speech understanding are always greater in reverberant compared to anechoic, environments (Brandewie & Zahorik, 2010, 2013; Srinivasan & Zahorik, 2012), i.e., improved performance arises only through exposure to a reverberant environment rather than to the speech material per se.

Adaptation to continuous noise is a well-known phenomenon in the auditory system (Costalupes et al., 1984; Gibson et al., 1985; Phillips, 1985; Rees & Palmer, 1988) that could potentially also explain the speech improvements observed among carrier lengths. This is particularly relevant in our experiments because the masking noise preceded the onset of the speech material, potentially providing a window for noise-adaptation to occur (Ainsworth & Meyer, 1994; Ben-David et al., 2016). However, Zahorik and Brandewie reported similar effect sizes in speech enhancements when 1 s (Brandewie & Zahorik, 2010) or 150 ms (Brandewie & Zahorik, 2013) of white noise was presented prior to the carrier phrase (i.e., CP0—|Color| |Number|). From this, the authors concluded that noise alone did not convey sufficient information about the acoustic environment to elicit improvements in speech understanding.

Adult listeners can also acclimatise to speech acoustics that deviate from the norm or from long-term language regularities such as dialects and foreign accents, with sufficient exposure (Idemaru & Holt, 2011, 2014; Liu & Holt, 2015). Even if our participants were all Australian-English native speakers exposed to an American-English speech corpus—the CRM—some perceptual learning of individual talker’s idiosyncratic speech patterns might arise over the course of the task (Choi & Perrachione, 2019; Liu & Holt, 2015; Stilp, 2020). In addition, it has been reported that ‘target voice continuity’ i.e., the same talker within trials—similar to our design—can enhance the build-up of selective attention and improve listeners abilities to report correctly digits sequences in the presence of other talkers (Best et al., 2008). However, when the room build-up effect—the effect assessed here—is explored in an isolated anechoic space i.e., not intermingled with room acoustic of other reverberant spaces, there is little or no improvement in speech understanding with increasing exposure to the talker per se; i.e., in the form of an increasing length of carrier phrase (Brandewie & Zahorik, 2010, 2013). Interrupting the continuity of the room—its reverberation profile—rather than the continuity of the target talker, is the factor reported to disrupt significantly the improvement in speech understanding with increasing exposure (Brandewie & Zahorik, 2018). If listeners adapted to a talker in the course of our experiment, this occurred rapidly, within a few speech tokens (Cooke et al., 2022; Kakehi, 1992; Kato & Kakehi, 1988) and contributed little to the enhancement of speech understanding we observed in reverberant rooms.

A key feature of this study, one that distinguishes it from previous assessments of statistical learning of acoustic features in human listeners, is our use of an ethologically valid listening task—understanding speech in background noise while listeners implicitly learned background features. Given the perceptual biases inherent in sensory processing, it is notable that investigators have previously reported different noise tokens to be differently learnable (Agus et al., 2014; Daikhin et al., 2017), for example, and that textures must be carefully controlled to ensure listeners are not exploiting subtle spectro-temporal features in their judgments of statistical similarity (McDermott et al., 2013). The potential for listeners to hear out specific—potentially unique to them—spectro-temporal features of otherwise statistically identical sound tokens, or to make judgments on the regularity or similarity of tone sequences based on unique listening experiences (Barascud et al., 2016; Bianco et al., 2020), presents a potential confound for these studies. Our study countermands this problem by actively exploiting the propensity for human listeners to ascribe meaning to spectro-temporal fluctuations (here, speech) in acoustic waveforms (Brandewie & Zahorik, 2010, 2013). Though still targeting the learning of background acoustic features, it engages listeners in a highly relevant listening task, one for which humans have rapidly evolved—i.e., understanding speech in background noise. Listeners’ attention, therefore, was called away from the background acoustic features to the ethologically relevant foreground task of attending to human speech, void of semantics; additionally, length of phrase, talker, and |Callsign| were uninformative to |Color| and |Number| in our utterances, as were |Color| and |Number| to each other. This misdirection allows us to exploit speech as a ‘reporter’ or biomarker for statistical learning of background, more-abstract, acoustic features, and without the potential confounding factor of (individualised) perceptual bias. Further, we have demonstrated that prior exposure to ethologically relevant environments matters when learning less-commonly encountered acoustic scenes.

The impact on speech understanding of more extreme (longer RT60s) environments is reminiscent of the detrimental effect of TMS, with lower listening performance. This suggests that learning the statistical structure of sound environments, to aid speech understanding in reverberant background noise, requires some longer form of memory (experience, developmental, or evolutionary). Brain circuits, therefore, might act to improve listening performance, operating best when at least some of the exposure to listening environments includes plausible, and commonly experienced, reverberation times (Traer & McDermott, 2016). When the acoustics of commonly encountered rooms were replaced with those of other environments having higher or lower amounts of reverberation, performance declined. Intriguingly, the most effective reverberation time we employed, at least in terms of its capacity to be learned and potentially from which learning could be transferred—the RT60traer of 0.49 s of the small Lecture Room is very close to the median and mean RT60traer of 0.42 and 0.50 s previously recorded for a wide range of built environments (indoors and outdoors). This suggests that statistical learning of acoustic background features might be tuned to ethologically relevant environments. Alternatively, built environments might be constructed to generate reverberation subjectively most suited to maximising speech-in-noise understanding, though we are unaware as to whether such a constraint is consciously applied in the design of listening spaces given contingencies such as the range and relative consistency and constancy of potential contents of any given space that influence its reverberant characteristics. Compared to built environments, however, the RT60traers of natural environments were much lower than all the rooms we assessed, save for that of the anechoic room, which was suboptimal in terms of listening performance.

Neural adaptation as a contributing mechanism to the learning of room acoustics

Our data are consistent with improved performance in a relevant ‘foreground’ task emerging as the brain adapts to the statistics of background features of the listening environment. Adaptation to stimulus statistics has been proposed as a neurophysiological mechanism underlying selective suppression of background noise in auditory cortex (Fuglsang et al., 2017; Kell & McDermott, 2019; Khalighinejad et al., 2019; Mesgarani et al., 2014). More recently, it has been shown that spectro-temporal receptive fields of auditory-cortical neurons are sensitive to the RT60s of reverberant noise i.e., a form of adaptation specific to reverberation and consistent with a de-reverberation of the environment encountered (Ivanov et al., 2022). Cortical adaptation to reverberation has been also reported in awake listeners (Fuglsang et al., 2017; Mesgarani et al., 2014) and to this end, our data suggest a role for top-down mechanisms in improving over time an attended task—recalling spoken words—in reverberant environments. It is therefore possible that goal-directed behaviours and the feedback the auditory cortex receives from areas such as dlPFC may contribute to fine-tuning the adaptation to reverberant environments.

Our data are also consistent with reports of midbrain auditory neurons recorded in vivo adapting to the statistical structure of evolving sound environments over the course of several hundred milliseconds (Dean et al., 2005, 2008) and this capacity for adaptation increases with repeated exposures to the same statistically structured environment. Importantly, whilst rapid adaptation is retained when descending cortical influences are disrupted (through cortical cooling), the longer-term learning effect—a speeding up of adaptation referred to as meta-adaptation—is abolished (Robinson et al., 2016). As with auditory neurons recorded in vivo (Bakay et al., 2018; Dean et al., 2005, 2008; Ivanov et al., 2022), it seems listeners might adapt to the statistical structure of a sound environment within a few hundred milliseconds of exposure to enhance performance in a listening task, an initial, rapid, learning phase impervious to inactivation of the dorsolateral prefrontal cortex by theta-burst TMS. The nested set of temporal sensitivities to room acoustics we observe is consistent with features of statistical learning that emerges from the level of the auditory nerve to primary cortex in vivo (Dean et al., 2005, 2008; Watkins & Barbour, 2008; Wen et al., 2009), and likely modulated by efferent influences that span the entire auditory pathway all the way to the sensory receptors of the inner ear (Hernández-Pérez et al., 2021; Perrot et al., 2006; Terreros & Delano, 2015) and even the mechanical sensitivity to sound of the eardrum and ossicles (middle-ear bones) (Gruters et al., 2018).

Dissociating auditory cortical feedback circuits to the midbrain impairs the capacity of midbrain neurons to ‘recall’ previous experience of that environment (Bajo et al., 2019; Robinson et al., 2016). Like the performance in our listening tasks, these neurons were still able to adapt rapidly (within hundreds of milliseconds) each time an environment was encountered as if for the first time, but showed no capacity to exploit a memory of that environment to improve neural coding (i.e., meta-adaptation) (Robinson et al., 2016). Meta-adaptation was consistent with neurons in lower brain centres adapting to the current sound environment, and those in higher brain centres learning the longer-term statistical structure of changing environments. Once higher brain centres learned the experienced environment, this information, conveyed to early brain centres, ensures they were ‘primed’ for coding an environment when it is re-encountered.

Inclusion and Ethics statement

This study was approved by the Human Research Ethics Committee of Macquarie University (ref: 5201833344874). Each participant signed a written informed consent form and was given a small financial remuneration for their time.